---
title: "'Opakapaka case study"
format: html
execute:
  eval: true
vignette: >
  %\VignetteIndexEntry{'Opakapaka case study}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor_options:
  chunk_output_type: console
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

This vignette configures, fits, and cross-platform benchmarks an `opal` stock
assessment model for 'opakapaka (*Pristipomoides filamentosus*), a Hawaiian
deepwater snapper. The reference model was built in Stock Synthesis 3
(SS3 v3.30.19) as part of the
[Opaka_simulation](https://github.com/noaa-pifsc/Opaka_simulation) project,
which provides both an operating model (OM) that generates the data and an
estimation model (EM) fitted to those data.

Because the OM was built in SS3 and the underlying dynamics are therefore
known, the case study serves two purposes: it evaluates the correspondence
between the `opal` and SS3 population-dynamics calculations under identical
inputs, and it compares `opal`'s estimated spawning-biomass trajectory against
both the operating-model truth and the SS3 estimation model.

Two versions of the `opal` model are configured, both fitting two CPUE indices
(generated with negligible observation error) together with length composition
data from the commercial and survey fleets:

- a **fixed-effects** model that estimates population scale, annual recruitment
  deviations, initial age structure, and the selectivity parameters of the
  fleets with length composition data, holding the remaining life-history and
  fishery parameters at the values used to generate the simulated data; and
- a **random-effects** model configured identically, except that the annual
  recruitment deviations and initial age structure are treated as random
  effects and recruitment variability (`log_sigma_r`) is estimated.

The model covers the historical period 1949--2023 with three fleets: a
commercial fishery (Comm), a non-commercial fishery (Non_comm), and a research
fishery survey (ResFish).

This case study is illustrative and is not intended to inform management
advice.

::: {.callout-note}
## Caching

Fitting both models and running `sdreport()` is the expensive part of this
document. To keep render time low for web/page builds, the fit is **cached**:
on a cache hit nothing from `opal`/`RTMB` is rebuilt and the document consumes
only a plain-R results bundle (`paka`). The build-and-fit code below runs only
when the cache is absent, in which case it regenerates and saves the bundle.
Delete the cached `.rds` to force a refit.
:::

# Load inputs

Load the `opal` package and its dependencies. `bayesplot` and `patchwork` are
used for the posterior-predictive diagnostics; `tidyverse` for data
manipulation and plotting.

```{r}
#| label: load-pkg
#| echo: true
#| message: false
library(tidyverse)
library(opal)
library(RTMB)
library(bayesplot)
library(patchwork)
```

## Plotting palette

A shared palette sampled from the `opal` logo is used across all figures so
that a given model keeps the same colour throughout. `theme_opal()` replaces
per-figure theme calls.

```{r}
#| label: setup-palette
#| echo: true
## Colours sampled from the opal logo
opal_cols <- c(
  navy  = "#013282", ocean = "#064E8F", midblue = "#066AA7", blue  = "#0181C3",
  teal  = "#27A7BE", aqua  = "#7DD2DB", gold    = "#FCCF43", coral = "#D92208",
  truth = "#4D4D4D", obs   = "#B3B3B3", struct  = "#8C8C8C"
)

## Model dimension: every label variant used here maps to a fixed colour
opal_models <- c(
  "opal"        = unname(opal_cols["teal"]),
  "opal (RE)"   = unname(opal_cols["midblue"]),
  "SS3 EM"      = unname(opal_cols["navy"]),
  "SS3 OM"      = unname(opal_cols["truth"])
)

scale_colour_opal <- function(...) ggplot2::scale_colour_manual(values = opal_models, ...)
scale_fill_opal   <- function(...) ggplot2::scale_fill_manual(values = opal_models, ...)

## Shared theme
theme_opal <- function(base_size = 11, base_family = "") {
  ggplot2::theme_minimal(base_size = base_size, base_family = base_family) +
    ggplot2::theme(
      legend.position  = "bottom",
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank()
    )
}
theme_set(theme_opal())
```

## Data

The bundled data object `opaka_data` contains all biological parameters, catch
and CPUE observations, length structure, and prior specifications needed for
the assessment model.

SS3 indexes ages from 0 to 43 (44 age classes), while `opal` uses 1-based
internal indexing. To align the growth curve, survivorship, and all
age-structured biology vectors with SS3, we set `A1 = 0` (anchoring the
Schnute growth parameterization at biological age 0) and redefine `age_a`
accordingly. The `min_age` field is passed to `get_growth()` so that the age
sequence starts at 0 rather than 1.

```{r}
#| label: load-data
#| echo: true
#| message: false
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L

# remove priors (regenerated below)
data$priors <- NULL

# anchor the growth curve at biological age 0 (SS3 indexes ages 0-43)
data$A1 <- 0
data$A2 <- data$A1 + (data$n_age - 1)
data$min_age <- 0
ages <- data$age_a <- seq(data$min_age, by = 1, length.out = data$n_age)

# calendar years for the modelled period (1949-2023)
calendar_years <- (1949 - 1) + data$years
```

Key dimensions of the data:

```{r}
#| label: data-dims
#| echo: true
cat("Number of ages:",       data$n_age,     "\n")
cat("Number of years:",      data$n_year,    "\n")
cat("Number of fisheries:",  data$n_fishery, "\n")
cat("Number of length bins:", data$n_len,    "\n")
```

## Biological inputs

Natural mortality is constant across all ages at M = 0.135 yr$^{-1}$. Maturity
follows a length-based logistic curve with Mat50 = 40.7 cm. The length-weight
relationship uses a = 1.75 $\times 10^{-5}$ and b = 2.99. Fecundity is
proportional to body weight (SS3 option 1 with a = 1, b = 0).

The sex ratio vector (`sex_ratio`) is set to 1.0 at all ages. Although SS3
applies a female fraction of 0.5 when computing SSB, this scaling is already
implicit in `opal`'s $B_0$ parameterization: `opal` defines $B_0$ as
female-only spawning biomass (matching SS3's reported SSB_Virgin), so the
per-recruit spawning potential does not require an additional sex-ratio
multiplier.

```{r}
#| label: plot-biology
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 7
data$sex_ratio <- rep(1, data$n_age)

len_mid <- seq(data$len_bin_start + data$len_bin_width / 2,
               by = data$len_bin_width, length.out = data$n_len)

bind_rows(
  data.frame(x = ages,    y = data$M,        panel = "M-at-age"),
  data.frame(x = len_mid, y = data$maturity, panel = "Maturity-at-length")
) |>
  ggplot(aes(x, y)) +
  geom_line(colour = opal_cols["teal"], linewidth = 0.9) +
  facet_wrap(~panel, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme_opal()
```

## Length composition data

Length composition observations are loaded from the bundled `opaka_lf`
long-format object and transformed into model-ready arrays. Data are available
for two fleets: Comm (fleet 1, 1949--2023) and ResFish (fleet 3, 2017--2023).
Fleet 2 (Non_comm) has no length composition data.

```{r}
#| label: prep-lf
#| echo: true
#| message: false
data(opaka_lf)

lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(
    id_cols     = c(fishery, year, month, ts),
    names_from  = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

var_adjust_scalars <- 1 / rep(1, data$n_fishery)

data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),
                     lf_var_adjust     = var_adjust_scalars,
                     lf_addtocomp      = 1e-10)
```

## SS3-style bias ramps

SS3 applies a bias-adjustment ramp to the lognormal recruitment correction
term following Methot and Taylor (2011). The ramp reduces the correction
$-0.5 \cdot \sigma_R^2$ in years where recruitment deviations are poorly
informed, preventing mean-unbiasing from introducing systematic bias when
deviations are estimated imprecisely. It is defined by four control years
(ascending start, plateau start, plateau end, descending end) and a maximum
bias-adjustment fraction. Two ramps are constructed: one for the initial age
cohorts (via `init_bias_adj_a`) and one for the time-series cohorts (via
`bias_adj_y`), both using the SS3 EM's estimated ramp parameters.

```{r}
#| label: add-bias-ramp
#| echo: true
#| message: false
# initial cohorts: age_a = 1 (bio age 0) -> born 1949; age_a = 44 -> born 1906
init_cohort_years <- 1949 - (data$age_a - 1)

data$init_bias_adj_a <- get_bias_adj_vector(
  years            = init_cohort_years,
  do_rec_bias_ramp = 1,
  bias_years       = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj     = 0.869
)

data$bias_adj_y <- get_bias_adj_vector(
  years            = calendar_years,
  do_rec_bias_ramp = 1,
  bias_years       = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj     = 0.869
)
```

# Model setup

## Parameters

Growth parameters (`log_L1`, `log_L2`, `log_k`, `log_CV1`, `log_CV2`) and
selectivity (`par_sel`) are initialised from the SS3 estimates in the bundled
`opaka_parameters` object. The 1949 recruitment deviation seeds the age-0
cohort of the initial age structure; deviations for 1950--2023 are shifted
into `rdev_y` with a terminal 0 appended.

```{r}
#| label: get-pars
#| echo: true
#| message: false
data(opaka_parameters)

# Route the 1949 deviation to biological age 0 (index 1 in opal's 1-based ages).
init_rdev_a <- rev(c(0.0206807, 0.00321579, 0.00372620, 0.0043174, 0.00500263,
  0.00579674, 0.00671727, 0.00778422, 0.00902081, 0.0104539, 0.0121143,
  0.0140375, 0.0162641, 0.0188404, 0.0218182, 0.0252561, 0.0292182, 0.0337726,
  0.0389898, 0.0449362, 0.0516662, 0.0592049, 0.0675236, 0.0764992, 0.0858443,
  0.0950047, 0.102706, 0.107687, 0.107712, 0.0998349, 0.0808137, 0.0482339,
  0.00234443, -0.0517563, -0.102395, -0.131458, -0.121101, -0.0714826,
  -0.00742166, -0.0109308, -0.0385029, 0.0366483, 0.0165105, -0.166906))
init_rdev_a[1] <- opaka_parameters$rdev_y[1]

# Shift 1950-2023 into rdev_y and append a terminal 0 for the forecast year.
rdev_y_shifted <- c(opaka_parameters$rdev_y[2:length(opaka_parameters$rdev_y)], 0.0)

parameters <- list(
  log_B0         = 9,
  log_h          = as.numeric(opaka_parameters$log_h),
  log_sigma_r    = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q     = as.numeric(opaka_parameters$log_cpue_q),
  log_cpue_tau   = rep(log(0.1), data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau     = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1         = as.numeric(opaka_parameters$log_L1),
  log_L2         = as.numeric(opaka_parameters$log_L2),
  log_k          = as.numeric(opaka_parameters$log_k),
  log_CV1        = as.numeric(opaka_parameters$log_CV1),
  log_CV2        = as.numeric(opaka_parameters$log_CV2),
  par_sel        = as.matrix(opaka_parameters$par_sel),
  log_init_F_f   = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y         = rdev_y_shifted,
  init_rdev_a    = init_rdev_a
)

# Convert SS3 selectivity parameterization to opal's internal parameterization.
parameters$par_sel <- convert_ss3_selex_to_rtmb(parameters$par_sel,
                                                data$sel_type_f, data$len_mid)
```

## Priors

```{r}
#| label: get-priors
#| echo: true
#| message: false
data$priors <- get_priors(parameters = parameters, data = data)
evaluate_priors(parameters = parameters, priors = data$priors)
```

## Parameter map

RTMB's `map` mechanism controls which parameters are estimated versus fixed;
parameters mapped to `factor(NA)` are held at their initial values. The base
map estimates `log_B0`, `log_cpue_q`, `rdev_y`, and `init_rdev_a`, while
steepness, $\sigma_R$, growth, initial F, and selectivity are held at their SS3
values. Selectivity is freed for the length-composition fleets in the model
configuration below.

```{r}
#| label: get-map
#| echo: true
#| message: false
map_sel    <- as.vector(matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel)))
map_lf_tau <- rep(NA, length(parameters$log_lf_tau))

map <- list(
  log_h          = factor(NA),
  log_sigma_r    = factor(NA),
  log_cpue_tau   = as.factor(rep(NA, data$n_index)),
  log_cpue_omega = as.factor(rep(NA, data$n_index)),
  log_lf_tau     = factor(map_lf_tau),
  log_L1         = factor(NA),
  log_L2         = factor(NA),
  log_k          = factor(NA),
  log_CV1        = factor(NA),
  log_CV2        = factor(NA),
  par_sel        = factor(map_sel),
  log_init_F_f   = rep(factor(NA), data$n_fishery),
  init_rdev_a    = as.factor(seq_len(data$n_age)),
  rdev_y         = as.factor(seq_len(length(parameters$rdev_y)))
)
```

# Fixed- and random-effects models

The fixed-effects `opal (with LF)` model turns on the length-composition
likelihood and frees the selectivity of the two fleets with length data: the
two logistic parameters for Comm (fleet 1) and the peak, ascending-width, and
descending-width parameters for the ResFish double-normal (fleet 3). Fleet 2
(Non_comm) retains its fixed SS3 selectivity because it has no length data. All
other parameters follow the base map: $\sigma_R$, steepness, and growth remain
fixed at SS3 values; $B_0$, catchabilities, `rdev_y`, and `init_rdev_a` are
freely estimated.

The random-effects model is configured identically, with two changes: `rdev_y`
and `init_rdev_a` are integrated out as random effects (via `random =`), and
`log_sigma_r` is estimated rather than fixed.

## Model configuration

These lightweight configuration objects (data and map) are cheap and always
evaluated, so the figures below can reference them regardless of the cache
state.

```{r}
#| label: model-config
#| echo: true
#| message: false
fleet_names <- c("Comm", "Non_comm", "ResFish")

# ---- Fixed-effects data + map (opal with LF) --------------------------------
data_lf <- data
data_lf$lf_switch <- 1L   # multinomial length-composition likelihood

map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1, 1:2]        <- c(1, 2)   # Comm:    logistic inflection + width
map_sel[3, c(1, 4, 6)] <- c(3, 4, 5)  # ResFish: peak, asc-width, desc-width
map_lf <- map
map_lf$par_sel <- factor(as.vector(map_sel))

# ---- Random-effects map -----------------------------------------------------
map_re <- map_lf
map_re$log_sigma_r <- as.factor(1)   # estimate recruitment variability
```

## Cache resolution

The cached results bundle is anchored to the document directory when available;
Quarto's execution directory can vary, so several candidate locations are
checked. On a cache hit the bundle is loaded and no model is rebuilt.

```{r}
#| label: cache-setup
#| echo: true
#| message: false
doc_dir <- tryCatch(dirname(knitr::current_input(dir = TRUE)),
                    error = function(e) ".")

paka_fit_candidates <- c(
  file.path(doc_dir, "cached", "opakapaka", "opal_lf-re_fit.rds"),
  file.path(doc_dir, "assets", "cached", "opakapaka", "opal_lf-re_fit.rds"),
  file.path(getwd(),  "cached", "opakapaka", "opal_lf-re_fit.rds"),
  file.path(getwd(),  "assets", "cached", "opakapaka", "opal_lf-re_fit.rds")
)

hit_paths     <- paka_fit_candidates[file.exists(paka_fit_candidates)]
cache_hit     <- length(hit_paths) > 0
paka_fit_path <- if (cache_hit) hit_paths[1] else paka_fit_candidates[1]

if (cache_hit) paka <- readRDS(paka_fit_path)
cat("Cache", if (cache_hit) "hit" else "miss", "->", paka_fit_path, "\n")
```

## Fit (cache miss only)

The build, bound, optimisation, and `sdreport` steps below execute **only when
the cache is absent**. `nlminb` is run several times to ensure convergence; the
final `fn` call resets `last.par` to the MLE so that `sdreport()` and
`simulate()` operate at the optimum.

```{r}
#| label: fe-fit
#| echo: true
#| results: hide
if (!cache_hit) {
  obj_lf <- MakeADFun(func = cmb(opal_model, data_lf),
                      parameters = parameters, map = map_lf)

  Lwr_lf <- rep(-Inf, length(obj_lf$par)); Upr_lf <- rep(Inf, length(obj_lf$par))
  Lwr_lf[grep("log_B0",      names(obj_lf$par))] <- log(1)
  Upr_lf[grep("log_B0",      names(obj_lf$par))] <- log(exp(22))
  Lwr_lf[grep("log_cpue_q",  names(obj_lf$par))] <- log(0.001)
  Upr_lf[grep("log_cpue_q",  names(obj_lf$par))] <- log(10)
  Lwr_lf[grep("rdev_y",      names(obj_lf$par))] <- -5
  Upr_lf[grep("rdev_y",      names(obj_lf$par))] <-  5
  Lwr_lf[grep("init_rdev_a", names(obj_lf$par))] <- -5
  Upr_lf[grep("init_rdev_a", names(obj_lf$par))] <-  5
  Lwr_lf[grep("par_sel",     names(obj_lf$par))] <- -7
  Upr_lf[grep("par_sel",     names(obj_lf$par))] <-  7

  control  <- list(eval.max = 10000, iter.max = 10000)
  start_fe <- Sys.time()
  opt_lf <- nlminb(obj_lf$par, obj_lf$fn, obj_lf$gr, lower = Lwr_lf, upper = Upr_lf, control = control)
  opt_lf <- nlminb(opt_lf$par, obj_lf$fn, obj_lf$gr, lower = Lwr_lf, upper = Upr_lf, control = control)
  opt_lf <- nlminb(opt_lf$par, obj_lf$fn, obj_lf$gr, lower = Lwr_lf, upper = Upr_lf, control = control)
  invisible(obj_lf$fn(opt_lf$par))
  sdr_lf     <- RTMB::sdreport(obj_lf)
  runtime_fe <- Sys.time() - start_fe
}
```

```{r}
#| label: re-fit
#| echo: true
#| results: hide
if (!cache_hit) {
  obj_re <- RTMB::MakeADFun(func = cmb(opal_model, data_lf),
                            parameters = parameters, map = map_re,
                            random = c("init_rdev_a", "rdev_y"), silent = TRUE)

  # Only the fixed parameters (log_B0, log_cpue_q, freed par_sel, log_sigma_r)
  # appear in obj_re$par; the random effects are marginalised.
  Lwr_re <- rep(-Inf, length(obj_re$par)); Upr_re <- rep(Inf, length(obj_re$par))
  Lwr_re[grep("log_B0",     names(obj_re$par))] <- log(1)
  Upr_re[grep("log_B0",     names(obj_re$par))] <- log(exp(22))
  Lwr_re[grep("log_cpue_q", names(obj_re$par))] <- log(0.001)
  Upr_re[grep("log_cpue_q", names(obj_re$par))] <- log(10)
  Lwr_re[grep("par_sel",    names(obj_re$par))] <- -7
  Upr_re[grep("par_sel",    names(obj_re$par))] <-  7

  start_re <- Sys.time()
  opt_re <- nlminb(obj_re$par, obj_re$fn, obj_re$gr, lower = Lwr_re, upper = Upr_re, control = control)
  opt_re <- nlminb(opt_re$par, obj_re$fn, obj_re$gr, lower = Lwr_re, upper = Upr_re, control = control)
  invisible(obj_re$fn(opt_re$par))
  sdr_re     <- RTMB::sdreport(obj_re)
  runtime_re <- Sys.time() - start_re
}
```

## Assemble and save the cache

On a cache miss, the fitted models are simulated from and summarised into the
plain-R `paka` bundle, which is then written to disk. Live `MakeADFun` objects
are **not** stored: they hold external pointers to compiled AD tapes that do not
round-trip through `saveRDS()`/`readRDS()` across sessions. Only plain-R
results (the `sdreport` objects, `report()` lists, simulated CPUE draws, and
convergence scalars) are cached.

```{r}
#| label: cache-save
#| echo: true
#| message: false
if (!cache_hit) {
  invisible(obj_lf$fn(opt_lf$par))
  set.seed(124)
  sims_cpue <- replicate(500, obj_lf$simulate()$cpue_log_obs)   # [obs x draws]

  paka <- list(
    par         = opt_lf$par,
    par_re      = opt_re$par,
    pdh         = isTRUE(sdr_lf$pdHess),
    pdh_re      = isTRUE(sdr_re$pdHess),
    mgc_sci     = formatC(max(abs(obj_lf$gr(opt_lf$par))), format = "e", digits = 2),
    mgc_sci_re  = formatC(max(abs(obj_re$gr(opt_re$par))), format = "e", digits = 2),
    runtime     = runtime_fe,
    runtime_re  = runtime_re,
    report      = obj_lf$report(),
    report_re   = obj_re$report(),
    sdreport    = sdr_lf,
    sdreport_re = sdr_re,
    sims_cpue   = sims_cpue,
    ssb         = data.frame(
      year     = calendar_years,
      estimate = obj_lf$report()$spawning_biomass_y[seq_along(calendar_years)]
    )
  )

  dir.create(dirname(paka_fit_path), recursive = TRUE, showWarnings = FALSE)
  saveRDS(paka, paka_fit_path)
}
```

## Convergence

Both models converged without issues detected in the gradient or Hessian, and
all parameters --- including `log_sigma_r` in the random-effects model --- were
identifiable according to `check_estimability()` at fit time. Standard-error
calculations completed quickly for both, and substantially faster than the
reference SS3 fit (~6 minutes), though the random-effects model was slower than
its fixed-effects counterpart.

```{r}
#| label: tbl-convergence
#| echo: true
#| message: false
n_random <- length(parameters$init_rdev_a) + length(parameters$rdev_y)

conv_tbl <- data.frame(
  Metric = c("Positive-definite Hessian",
             "Max |gradient|",
             "Runtime (seconds)",
             "N parameters (fixed)",
             "N parameters (random)",
             "sigma_R"),
  opal = c(
    as.character(paka$pdh),
    paka$mgc_sci,
    round(as.numeric(paka$runtime, units = "secs"), 1),
    length(paka$par),
    0L,
    paste0(round(exp(as.numeric(parameters$log_sigma_r)), 2), " (fixed)")
  ),
  opal_re = c(
    as.character(paka$pdh_re),
    paka$mgc_sci_re,
    round(as.numeric(paka$runtime_re, units = "secs"), 1),
    length(paka$par_re),
    n_random,
    round(exp(unname(paka$par_re["log_sigma_r"])), 2)
  ),
  stringsAsFactors = FALSE
)
colnames(conv_tbl) <- c("Metric", "opal", "opal (RE)")

knitr::kable(conv_tbl, booktabs = TRUE, align = c("l", "l", "l"))
```

# Cross-platform comparison

## Estimated spawning biomass

The figure compares `opal`'s estimated spawning-biomass trajectory (fixed and
random effects, with 95% confidence intervals) against the SS3 operating-model
truth (grey dashed) and the SS3 estimation model (navy dotted). `opal`
reproduces the SS3 simulated trajectory, demonstrating that the internal
population-dynamics engine is sound; the fixed- and random-effects estimates
are close to one another.

```{r}
#| label: fig-paka-biomass
#| echo: true
#| message: false
#| fig-cap: "'Opakapaka spawning biomass: opal fixed-effects (teal) and random-effects (blue) estimates with 95% intervals, against the SS3 operating-model truth (grey dashed) and SS3 estimation model (navy dotted)."
#| fig-height: 4.5
#| fig-width: 8
data(opaka_truth)

sbio_from_sdr <- function(sdr, label) {
  v  <- sdr$value[names(sdr$value) == "spawning_biomass_y"][seq_along(calendar_years)]
  se <- sdr$sd[   names(sdr$value) == "spawning_biomass_y"][seq_along(calendar_years)]
  data.frame(model = label, year = calendar_years, value = v, se = se,
             lo95 = v - 1.96 * se, hi95 = v + 1.96 * se)
}

paka_fe <- sbio_from_sdr(paka$sdreport,    "opal")
paka_re <- sbio_from_sdr(paka$sdreport_re, "opal (RE)")

om <- subset(opaka_truth$om$sprseries, Yr %in% calendar_years)
em <- subset(opaka_truth$em$sprseries, Yr %in% calendar_years)

ybmax <- max(paka_fe$hi95, paka_re$hi95, om$SSB, em$SSB, na.rm = TRUE)

ggplot() +
  geom_ribbon(data = paka_fe, aes(year, ymin = lo95, ymax = hi95, fill = model), alpha = 0.25) +
  geom_ribbon(data = paka_re, aes(year, ymin = lo95, ymax = hi95, fill = model), alpha = 0.25) +
  geom_line(data = paka_fe, aes(year, value, colour = "opal"),      linewidth = 1) +
  geom_line(data = paka_re, aes(year, value, colour = "opal (RE)"), linewidth = 1) +
  geom_line(data = om, aes(Yr, SSB, colour = "SS3 OM"), linewidth = 1, linetype = "dashed") +
  geom_line(data = em, aes(Yr, SSB, colour = "SS3 EM"), linewidth = 1, linetype = "dotted") +
  scale_colour_opal("Model", breaks = c("opal", "opal (RE)", "SS3 OM", "SS3 EM")) +
  scale_fill_opal("Model",   breaks = c("opal", "opal (RE)", "SS3 OM", "SS3 EM")) +
  labs(x = "Year", y = "Spawning biomass", colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax * 1.2)) +
  theme_opal()
```

## Estimated selectivity

Estimated selectivity-at-age for the three fleets under the fixed-effects
model (from the cached `report()`). Comm and ResFish are informed directly by
the length composition data; Non_comm retains its fixed SS3 selectivity.

```{r}
#| label: fig-paka-selex
#| echo: true
#| message: false
#| fig-height: 4
#| fig-width: 10
#| fig-cap: Estimated selectivity-at-age by fleet (fixed-effects model).
sel_fya <- paka$report$sel_fya

sel_df <- expand.grid(fishery = seq_len(data$n_fishery), age = seq_len(data$n_age))
sel_df$selectivity <- mapply(function(f, a) sel_fya[f, 1, a], sel_df$fishery, sel_df$age)
sel_df$age_real <- data$age_a[sel_df$age]
sel_df$fleet    <- fleet_names[sel_df$fishery]

ggplot(sel_df, aes(age_real, selectivity)) +
  geom_line(colour = opal_cols["teal"], linewidth = 1) +
  facet_wrap(~fleet, ncol = 3) +
  labs(x = "Age", y = "Selectivity") +
  ylim(0, 1) +
  theme_opal()
```

## Length composition fits

Observed length compositions (grey bars) with fixed-effects model predictions
(teal), for a decadal subset of Comm years and all ResFish years. Predictions
are taken from the cached `report()$lf_pred`.

```{r}
#| label: fig-paka-lf-fit
#| echo: true
#| message: false
#| fig-height: 6
#| fig-width: 9
#| fig-cap: Length-composition fits for the commercial (decadal subset) and research (all years) fleets.
lf_pred <- paka$report$lf_pred
names(lf_pred) <- data_lf$lf_fishery_f
for (k in seq_along(lf_pred)) colnames(lf_pred[[k]]) <- data_lf$len_mid

lf_year_list <- split(data_lf$lf_year, data_lf$lf_fishery)

df_pred <- map_dfr(seq_along(lf_pred), function(i) {
  m   <- lf_pred[[i]]
  yrs <- lf_year_list[[i]]
  as.data.frame.table(m, responseName = "pred", stringsAsFactors = FALSE) %>%
    rename(id = Var1, length = Var2) %>%
    mutate(fishery = as.character(names(lf_pred)[i]),
           id      = as.integer(factor(id, levels = unique(id))),
           year    = yrs[id],
           length  = as.numeric(length))
})

df_obs <- data.frame(data_lf$lf_obs_in)
names(df_obs) <- data_lf$len_mid
df_obs <- df_obs %>%
  mutate(year = data_lf$lf_year, fishery = as.character(data_lf$lf_fishery)) %>%
  group_by(fishery) %>% mutate(id = row_number()) %>% ungroup() %>%
  pivot_longer(cols = -c(year, fishery, id), names_to = "length", values_to = "obs") %>%
  mutate(length = as.numeric(length))

df_lf <- left_join(df_obs, df_pred, by = c("id", "fishery", "length")) %>%
  select(-year.y) %>% rename(year = year.x) %>%
  mutate(cal_year = 1948L + year)

plot_lf <- function(fleet_id, ncol, subtitle) {
  yrs <- if (fleet_id == "1") {
    seq(min(data_lf$lf_year[data_lf$lf_fishery == 1]),
        max(data_lf$lf_year[data_lf$lf_fishery == 1]), by = 10L)
  } else {
    sort(unique(data_lf$lf_year[data_lf$lf_fishery == as.integer(fleet_id)]))
  }
  ggplot(df_lf %>% filter(fishery == fleet_id, year %in% yrs), aes(length)) +
    geom_col(aes(y = obs), fill = opal_cols["obs"], width = 2) +
    geom_line(aes(y = pred), colour = opal_cols["teal"], linewidth = 0.8) +
    facet_wrap(~cal_year, scales = "free_y", ncol = ncol) +
    labs(x = "Length (cm)", y = "Proportion", subtitle = subtitle) +
    theme_opal() +
    theme(strip.text = element_text(size = 7), axis.text = element_text(size = 6))
}

print(plot_lf("1", 5, "Comm (fleet 1, decadal subset)"))
print(plot_lf("3", 4, "ResFish (fleet 3, all years)"))
```

# Posterior-predictive checks

`RTMB`'s `simulate()` function generated the cached CPUE draws
(`paka$sims_cpue`) from the fitted model. These simulated datasets can be refit
within the model (simulation self-testing) or used to generate
posterior-predictive distributions for the observed data. Using the commercial
fishery index as an example, the observed CPUE tends to fall within the
posterior-predictive distribution (panels A and B). The PIT ECDF (panel C),
however, indicates the predictions are over-dispersed relative to the
observations. This is expected: the commercial index was generated with
negligible observation error, whereas `simulate()` draws incorporate the full
uncertainty (observation *and* process error) of the estimated model.

```{r}
#| label: fig-paka-cpue-ppc
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Posterior-predictive checks for the 'opakapaka commercial CPUE index (fishery 1). Simulated replicates in aqua, observations in dark teal. Left to right: replicate trajectories, density overlay, and PIT ECDF calibration check."
#| fig-height: 7
#| fig-width: 9
nsims_traj <- 50L
sims_cpue  <- paka$sims_cpue   # [obs x draws]

# bayesplot's 6 scheme slots run light -> dark:
opal_sim <- "#7DD2DB"   # simulated replicates (aqua)
opal_obs <- "#0E4049"   # observed (dark teal)
color_scheme_set(c(opal_sim, "#5EBFCB", "#27A7BE", "#1F8B9E", "#155F6B", opal_obs))

idx  <- which(data$cpue_data$fishery == 1)
y    <- log(data$cpue_data$value[idx])       # observed, log units
yrep <- t(as.matrix(sims_cpue[idx, ]))       # [draws x obs]

# guard against NA columns (years without CPUE) that break density estimation
ok   <- stats::complete.cases(t(yrep)) & !is.na(y)
y    <- y[ok]
yrep <- yrep[, ok, drop = FALSE]
stopifnot(ncol(yrep) == length(y))

set.seed(1)
draws <- sample(nrow(yrep), min(nsims_traj, nrow(yrep)))

p_dens <- ppc_dens_overlay(y, yrep[draws, , drop = FALSE]) +
  ggtitle("CPUE density overlay") + xlab("log(CPUE)") + ylab("Density") + theme_opal()

p_pit <- ppc_pit_ecdf(y, yrep) +
  ggtitle("PIT ECDF") + theme_opal()

mle_sims <- as.data.frame(sims_cpue[idx, ]) %>%
  pivot_longer(cols = everything(), names_to = "sim", values_to = "cpue") %>%
  filter(sim %in% sample(unique(sim), nsims_traj))
mle_sims$x <- sort(rep(data$cpue_data$year[idx], nsims_traj))

p_traj <- ggplot() +
  geom_line(data = mle_sims, aes(x = x, y = cpue, group = sim),
            colour = opal_sim, alpha = 0.4) +
  geom_line(data = data$cpue_data[idx, ], aes(x = year, y = log(value)),
            colour = opal_obs, linewidth = 1.5) +
  labs(x = "Year", y = "log(CPUE)") +
  ggtitle("CPUE trajectories") + theme_opal()

p_traj + p_dens + p_pit + plot_annotation(tag_levels = "A")
```

# Next steps

This vignette covers configuration, fitting, cross-platform benchmarking, and
posterior-predictive checking for the fixed- and random-effects `opal (with LF)`
models. The projection and Bayesian (MCMC) workflows built on the fitted
fixed-effects model --- constant-catch projection with MVN and MCMC uncertainty
propagation --- are carried forward in
`vignette("projections_opakapaka")`, which reuses the same model as its
starting point.
