---
title: "opal: the open population assessment library"
date: last-modified
date-format: "DD MMMM YYYY"
author:
  - name: "N. Ducharme-Barth"
    affiliations:
      - id: pifsc
        name: "NOAA Fisheries, Pacific Islands Fisheries Science Center, 1845 Wasp Boulevard, Bldg. 176, Honolulu HI 96818"
  - name: "D. Webber"
    affiliations:
      - id: quantifish
        name: "Quantifish Limited, Tauranga, New Zealand"
  - name: "P. Neubauer"
    affiliations:
      - id: dragonfly
        name: "Dragonfly Data Science, Wellington, New Zealand"
  - name: "K. Kim"
    affiliations:
      - id: spc
        name: "Oceanic Fisheries Programme, Pacific Community, Noumea, New Caledonia"
  - name: "A. Magnusson"
    affiliations:
      - ref: spc
  - name: "P. Hamer"
    affiliations:
      - ref: spc
meetingtype: "SCIENTIFIC COMMITTEE"
meetingnumber: "TWENTY-SECOND REGULAR SESSION"
meetinglocation: "Apia, Samoa"
meetingdates: "11-19 August 2026"
repnumber: WCPFC-SC22-2026/SA-WP-02
format:
  pdf:
    pdf-engine: pdflatex
    template-partials:
      - before-body.tex
    documentclass: SCreport
    fontsize: 11pt
    geometry:
      - margin=1in
    toc: true
    number-sections: true
    link-citations: true
    colorlinks: true
    linkcolor: SCblue
    filecolor: SCblue
    urlcolor: SCblue
    citecolor: SCblue
    toccolor: SCblue
    highlight-style: github
    code-block-bg: "#f6f8fa"
    code-block-border-left: "#c5c5c5"
    code-font: "Consolas, Monaco, 'Andale Mono', monospace"
    filters:
      - inline-code.lua
    header-includes: |
     \usepackage{draftwatermark}
     \SetWatermarkLightness{0.92}
     \usepackage{pdflscape}
     \providecommand{\citeproc}[2]{#2}
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    link-citations: true
    colorlinks: true
    highlight-style: github
    self-contained: true
execute:
  warning: false
  message: false
bibliography: opal.bib
csl: apa.csl

---

```{=html}
<style>
.draft-watermark {
  position: fixed;
  top: 50%;
  left: 50%;
  transform: translate(-50%, -50%) rotate(-45deg);
  font-size: 100px;
  font-weight: bold;
  color: #ccc;
  opacity: 0.5;
  z-index: -1000;
  pointer-events: none;
}
</style>
<div class="draft-watermark">DRAFT</div>
```

```{r}
#| label: setup-opal
#| include: false
# renv::restore(prompt = FALSE)   # optional: reproduce library from renv.lock

library(opal)    # pinned to commit 32239a3 via renv.lock
library(RTMB)
library(tidyverse)
library(SparseNUTS)
library(posterior)

theme_set(theme_bw())
```

```{r}
#| label: paka-lf-fit
#| echo: false
#| results: hide

## ---- Cache location ------------------------------------------------------
# Anchor to the document's own directory when available. Quarto's execution
# directory can vary, so check the cache locations that occur when rendering
# from either the document directory or the project root.
doc_dir <- tryCatch(
  dirname(knitr::current_input(dir = TRUE)),
  error = function(e) "."
)
paka_fit_candidates <- c(
  file.path(doc_dir, "cached", "opakapaka", "opal_lf_fit.rds"),
  file.path(doc_dir, "assets", "cached", "opakapaka", "opal_lf_fit.rds"),
  file.path(getwd(), "cached", "opakapaka", "opal_lf_fit.rds"),
  file.path(getwd(), "assets", "cached", "opakapaka", "opal_lf_fit.rds")
)
paka_fit_path <- paka_fit_candidates[file.exists(paka_fit_candidates)][1]
paka_cache_dir <- if (is.na(paka_fit_path)) {
  file.path(doc_dir, "cached", "opakapaka")
} else {
  dirname(paka_fit_path)
}
dir.create(paka_cache_dir, recursive = TRUE, showWarnings = FALSE)
if (is.na(paka_fit_path)) {
  paka_fit_path <- file.path(paka_cache_dir, "opal_lf_fit.rds")
}

## ---- Fit + diagnostics (fully cached) ------------------------------------
# On a cache hit nothing from opal/RTMB is rebuilt: the document consumes only
# the plain-R results bundle (`paka`). The model is built, fitted, and
# summarised solely when the cache is absent. (R ignores indentation, so the
# pipeline below is left un-indented inside the else for a smaller diff.)
if (file.exists(paka_fit_path)) {
  paka <- readRDS(paka_fit_path)
} else {

## ---- Data ----------------------------------------------------------------
data(opaka_data)
data <- opaka_data
if (is.null(data$n_index)) data$n_index <- 2L
data$priors <- NULL

# SS3 indexes ages 0--43; anchor the Schnute growth curve at biological age 0.
data$A1      <- 0
data$A2      <- data$A1 + (data$n_age - 1)
data$min_age <- 0
data$age_a   <- seq(data$min_age, by = 1, length.out = data$n_age)

# Sex ratio set to 1 at all ages (B0 is already female-only spawning biomass).
data$sex_ratio <- rep(1, data$n_age)

## ---- Length composition data ---------------------------------------------
data(opaka_lf)
lf_wide <- opaka_lf %>%
  tidyr::pivot_wider(
    id_cols     = c(fishery, year, month, ts),
    names_from  = bin,
    values_from = value,
    values_fill = 0
  ) %>%
  arrange(fishery, ts)

var_adjust_scalars <- 1 / rep(1, data$n_fishery)
data <- prep_lf_data(data, lf_wide,
                     lf_keep_fisheries = c(1, 3),   # Comm + ResFish
                     lf_var_adjust     = var_adjust_scalars,
                     lf_addtocomp      = 1e-10)

## ---- SS3-style recruitment bias-adjustment ramps -------------------------
init_cohort_years <- 1949 - (data$age_a - 1)
data$init_bias_adj_a <- get_bias_adj_vector(
  years            = init_cohort_years,
  do_rec_bias_ramp = 1,
  bias_years       = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj     = 0.869
)

calendar_years <- (1949 - 1) + data$years   # 1949--2023
data$bias_adj_y <- get_bias_adj_vector(
  years            = calendar_years,
  do_rec_bias_ramp = 1,
  bias_years       = c(1931, 1974, 2021.88, 2022.43),
  max_bias_adj     = 0.869
)

## ---- Parameters ----------------------------------------------------------
data(opaka_parameters)

# 1949 dev seeds the initial age-0 cohort; later years shift into rdev_y.
init_rdev_a <- rev(c(
  0.0206807, 0.00321579, 0.00372620, 0.0043174, 0.00500263, 0.00579674,
  0.00671727, 0.00778422, 0.00902081, 0.0104539, 0.0121143, 0.0140375,
  0.0162641, 0.0188404, 0.0218182, 0.0252561, 0.0292182, 0.0337726,
  0.0389898, 0.0449362, 0.0516662, 0.0592049, 0.0675236, 0.0764992,
  0.0858443, 0.0950047, 0.102706, 0.107687, 0.107712, 0.0998349,
  0.0808137, 0.0482339, 0.00234443, -0.0517563, -0.102395, -0.131458,
  -0.121101, -0.0714826, -0.00742166, -0.0109308, -0.0385029, 0.0366483,
  0.0165105, -0.166906
))
init_rdev_a[1] <- opaka_parameters$rdev_y[1]
rdev_y_shifted <- c(opaka_parameters$rdev_y[2:length(opaka_parameters$rdev_y)], 0.0)

parameters <- list(
  log_B0         = 9,
  log_h          = as.numeric(opaka_parameters$log_h),
  log_sigma_r    = as.numeric(opaka_parameters$log_sigma_r),
  log_cpue_q     = as.numeric(opaka_parameters$log_cpue_q),
  log_cpue_tau   = rep(log(0.1), data$n_index),
  log_cpue_omega = as.numeric(rep(opaka_parameters$log_cpue_omega, data$n_index)),
  log_lf_tau     = as.numeric(log(rep(0.1, data$n_fishery))),
  log_L1         = as.numeric(opaka_parameters$log_L1),
  log_L2         = as.numeric(opaka_parameters$log_L2),
  log_k          = as.numeric(opaka_parameters$log_k),
  log_CV1        = as.numeric(opaka_parameters$log_CV1),
  log_CV2        = as.numeric(opaka_parameters$log_CV2),
  par_sel        = as.matrix(opaka_parameters$par_sel),
  log_init_F_f   = c(log(0.0119122), rep(log(1e-8), data$n_fishery - 1)),
  rdev_y         = rdev_y_shifted,
  init_rdev_a    = init_rdev_a
)
parameters$par_sel <- convert_ss3_selex_to_rtmb(
  parameters$par_sel, data$sel_type_f, data$len_mid
)

## ---- Priors --------------------------------------------------------------
data$priors <- get_priors(parameters = parameters, data = data)

## ---- Base map ------------------------------------------------------------
map_sel_base    <- as.vector(matrix(NA, nrow(parameters$par_sel),
                                     ncol(parameters$par_sel)))
map_lf_tau_base <- rep(NA, length(parameters$log_lf_tau))

map <- list(
  log_h          = factor(NA),
  log_sigma_r    = factor(NA),
  log_cpue_tau   = as.factor(rep(NA, data$n_index)),
  log_cpue_omega = as.factor(rep(NA, data$n_index)),
  log_lf_tau     = factor(map_lf_tau_base),
  log_L1         = factor(NA),
  log_L2         = factor(NA),
  log_k          = factor(NA),
  log_CV1        = factor(NA),
  log_CV2        = factor(NA),
  par_sel        = factor(map_sel_base),
  log_init_F_f   = rep(factor(NA), data$n_fishery),
  init_rdev_a    = as.factor(seq_len(data$n_age)),
  rdev_y         = as.factor(seq_len(length(parameters$rdev_y)))
)

## ---- opal (with LF): data, map, AD object --------------------------------
# Turn on the multinomial length-composition likelihood and free selectivity
# for the two fleets with LF data (Comm: two logistic params; ResFish: three
# of the six double-normal params). CPUE remains in the likelihood.
data_lf <- data
data_lf$sel_fa_external <- NULL    # estimate selectivity, no external override
data_lf$lf_switch       <- 1L

map_sel <- matrix(NA, nrow(parameters$par_sel), ncol(parameters$par_sel))
map_sel[1, 1:2]       <- c(1, 2)         # Comm (fleet 1) logistic
map_sel[3, c(1, 4, 6)] <- c(3, 4, 5)     # ResFish (fleet 3) double-normal
map_sel <- as.vector(map_sel)

map_lf <- map
map_lf$par_sel <- factor(map_sel)

obj_lf <- MakeADFun(func = cmb(opal_model, data_lf),
                    parameters = parameters,
                    map        = map_lf)
stopifnot(is.finite(obj_lf$fn()))

## ---- Parameter bounds ----------------------------------------------------
Lwr_lf <- rep(-Inf, length(obj_lf$par))
Upr_lf <- rep( Inf, length(obj_lf$par))
Lwr_lf[grep("log_B0",      names(obj_lf$par))] <- log(1)
Upr_lf[grep("log_B0",      names(obj_lf$par))] <- log(exp(22))
Lwr_lf[grep("log_cpue_q",  names(obj_lf$par))] <- log(0.001)
Upr_lf[grep("log_cpue_q",  names(obj_lf$par))] <- log(10)
Lwr_lf[grep("rdev_y",      names(obj_lf$par))] <- -5
Upr_lf[grep("rdev_y",      names(obj_lf$par))] <-  5
Lwr_lf[grep("init_rdev_a", names(obj_lf$par))] <- -5
Upr_lf[grep("init_rdev_a", names(obj_lf$par))] <-  5
Lwr_lf[grep("par_sel",     names(obj_lf$par))] <- -7
Upr_lf[grep("par_sel",     names(obj_lf$par))] <-  7

## ---- Fit: three-pass nlminb ----------------------------------------------
control <- list(eval.max = 10000, iter.max = 10000)
opt_lf <- nlminb(start = obj_lf$par, objective = obj_lf$fn,
                 gradient = obj_lf$gr, lower = Lwr_lf, upper = Upr_lf,
                 control = control)
opt_lf <- nlminb(start = opt_lf$par, objective = obj_lf$fn,
                 gradient = obj_lf$gr, lower = Lwr_lf, upper = Upr_lf,
                 control = control)
opt_lf <- nlminb(start = opt_lf$par, objective = obj_lf$fn,
                 gradient = obj_lf$gr, lower = Lwr_lf, upper = Upr_lf,
                 control = control)
invisible(obj_lf$fn(opt_lf$par))   # set last.par to the MLE

## ---- Convergence diagnostics ---------------------------------------------
sdr_lf   <- RTMB::sdreport(obj_lf)
paka_mgc <- max(abs(obj_lf$gr(opt_lf$par)))
# Estimability is the convergence criterion (in place of opt_lf$convergence);
# capture the full return so it is cached and re-printed verbatim in BLOCK C.
estim    <- tryCatch(check_estimability(obj = obj_lf), error = function(e) e)

## ---- Estimated spawning biomass + SS3 OM / EM ----------------------------
rep_lf <- obj_lf$report()
data(opaka_truth)

## ---- Bayesian (MCMC) summaries -------------------------------------------
# The dense-metric NUTS fit is pre-computed and distributed with the package.
n_chains_mcmc <- 8L
mcmc_fit <- readRDS(system.file("extdata", "opaka_mcmc_fit.rds", package = "opal"))

post     <- SparseNUTS::extract_samples(mcmc_fit)
n_post   <- nrow(post); n_per <- n_post / n_chains_mcmc
chain_id <- rep(seq_len(n_chains_mcmc), each = n_per)
iter_id  <- rep(seq_len(n_per),         times = n_chains_mcmc)
mu_vec   <- obj_lf$env$last.par.best

# Diagnostics table: scalar parameters + three evenly-spaced recruitment deviates
rdev_cols  <- grep("rdev_y", colnames(post))
show_rdev  <- colnames(post)[rdev_cols[round(seq(1, length(rdev_cols), length.out = 3))]]
diag_cols  <- c("log_B0", "log_cpue_q[1]", "log_cpue_q[2]", show_rdev)
ddf <- as.data.frame(post[, diag_cols, drop = FALSE])
ddf$.chain <- chain_id; ddf$.iteration <- iter_id
mcmc_diag <- posterior::summarise_draws(
  posterior::as_draws_df(ddf),
  posterior::default_summary_measures(),
  posterior::default_convergence_measures())
# Match the MLE to each (possibly indexed) posterior parameter name: mu_vec has
# repeated names for vector parameters, so resolve "name[i]" positionally.
mle_for <- function(v) {
  if (grepl("\\[", v)) {
    base <- sub("\\[.*", "", v)
    idx  <- as.integer(sub(".*\\[(\\d+)\\].*", "\\1", v))
    unname(mu_vec[names(mu_vec) == base][idx])
  } else {
    unname(mu_vec[v])
  }
}
mcmc_diag$mle <- vapply(mcmc_diag$variable, mle_for, numeric(1))

# Trace data for the scalar parameters
scal       <- c("log_B0", "log_cpue_q[1]", "log_cpue_q[2]")
lcq        <- mu_vec[names(mu_vec) == "log_cpue_q"]
mcmc_trace <- do.call(rbind, lapply(scal, function(p)
  data.frame(iter = iter_id, chain = factor(chain_id), param = p, value = post[, p])))
mcmc_mle_ref <- data.frame(param = scal,
                           mle = c(mu_vec[["log_B0"]], lcq[1], lcq[2]))

## ---- Forward projections (MVN + MCMC), constant terminal catch -----------
sdr_sum_lf  <- summary(sdr_lf, "report")
sbio_rows   <- which(rownames(sdr_sum_lf) == "spawning_biomass_y")
sbio_mle_sd <- sdr_sum_lf[sbio_rows, "Estimate"]
sbio_se_sd  <- sdr_sum_lf[sbio_rows, "Std. Error"]
B0_est      <- rep_lf$B0
term_idx    <- data$n_year + 1L

# Fields required by the projection helpers
data_lf$first_yr         <- 1949L
data_lf$last_yr          <- 2023L
data_lf$removal_switch_f <- rep(0L, data_lf$n_fishery)

n_proj <- 10L; n_iter <- 200L

rdev_proj <- project_rec_devs(data = data_lf, obj = obj_lf,
  first_yr = data_lf$first_yr, n_proj = n_proj, n_iter = n_iter)
sel_proj  <- project_selectivity(data = data_lf, obj = obj_lf,
  first_yr = data_lf$last_yr - 9, n_proj = n_proj, n_iter = n_iter)
catch_ysf_proj <- array(0, dim = c(n_proj, data_lf$n_season, data_lf$n_fishery))
for (y in seq_len(n_proj)) catch_ysf_proj[y, , ] <- data_lf$catch_obs_ysf[data_lf$n_year, , ]

proj_mvn <- project_dynamics(data = data_lf, object = obj_lf, mcmc = NULL,
  n_proj = n_proj, n_iter = n_iter, rdev_y = rdev_proj$rdev_y,
  sel_fya = sel_proj, catch_ysf = catch_ysf_proj, return_hist = TRUE)

rdev_proj_mcmc <- project_rec_devs(data = data_lf, obj = obj_lf, mcmc = mcmc_fit,
  first_yr = data_lf$first_yr, n_proj = n_proj, n_iter = n_iter)
proj_mcmc <- project_dynamics(data = data_lf, object = obj_lf, mcmc = mcmc_fit,
  n_proj = n_proj, n_iter = n_iter, rdev_y = rdev_proj_mcmc$rdev_y,
  sel_fya = sel_proj, catch_ysf = catch_ysf_proj, return_hist = TRUE)

# Projected spawning biomass per iteration (proj_year 0 = terminal bridge state)
spa <- rep_lf$spawning_potential_a
sbio_iter <- function(proj_list, label) {
  do.call(rbind, lapply(seq_along(proj_list), function(i) {
    dyn <- proj_list[[i]]
    sb  <- sapply(1:(n_proj + 1), function(y) sum(dyn$number_ysa[y, 1, ] * spa))
    data.frame(iter = i, proj_year = 0:n_proj, sbio = sb, method = label)
  }))
}
df_all <- rbind(sbio_iter(proj_mvn$dyn, "MVN"), sbio_iter(proj_mcmc$dyn, "MCMC"))

# Historical window (last 60 yr) mapped to calendar years (t = n_year+1 -> 2024)
n_hist_show <- 60L
hist_idx    <- (term_idx - n_hist_show + 1):term_idx
hist_years  <- 1948L + hist_idx

df_hist_mvn <- data.frame(year = hist_years,
  med  = sbio_mle_sd[hist_idx],
  lo50 = sbio_mle_sd[hist_idx] - qnorm(0.75) * sbio_se_sd[hist_idx],
  hi50 = sbio_mle_sd[hist_idx] + qnorm(0.75) * sbio_se_sd[hist_idx],
  lo90 = sbio_mle_sd[hist_idx] - qnorm(0.95) * sbio_se_sd[hist_idx],
  hi90 = sbio_mle_sd[hist_idx] + qnorm(0.95) * sbio_se_sd[hist_idx],
  method = "MVN")
hist_ribbon <- function(hist_sbio, label, idx, years) {
  do.call(rbind, lapply(seq_along(idx), function(k) {
    v <- hist_sbio[, idx[k]]
    data.frame(year = years[k], med = median(v, na.rm = TRUE),
      lo50 = quantile(v, .25, na.rm = TRUE), hi50 = quantile(v, .75, na.rm = TRUE),
      lo90 = quantile(v, .05, na.rm = TRUE), hi90 = quantile(v, .95, na.rm = TRUE),
      method = label)
  }))
}
df_hist_mcmc     <- hist_ribbon(proj_mcmc$hist_sbio, "MCMC", hist_idx, hist_years)
proj_hist_ribbon <- rbind(df_hist_mvn, df_hist_mcmc)

proj_ribbon <- df_all |>
  dplyr::filter(!is.na(sbio)) |>
  dplyr::group_by(method, proj_year) |>
  dplyr::summarise(med  = median(sbio),
                   lo50 = quantile(sbio, .25), hi50 = quantile(sbio, .75),
                   lo90 = quantile(sbio, .05), hi90 = quantile(sbio, .95),
                   .groups = "drop") |>
  dplyr::mutate(year = 2024L + proj_year)

proj_hist_mle <- data.frame(year = hist_years, sbio = sbio_mle_sd[hist_idx])
proj_om <- subset(opaka_truth$om$sprseries, Yr %in% hist_years, select = c(Yr, SSB))
proj_em <- subset(opaka_truth$em$sprseries, Yr %in% hist_years, select = c(Yr, SSB))

## ---- Results bundle (everything the document consumes) -------------------
paka <- list(
  par     = opt_lf$par,
  pdh     = isTRUE(sdr_lf$pdHess),
  mgc     = paka_mgc,
  mgc_sci = formatC(paka_mgc, format = "e", digits = 2),
  estim   = estim,
  ssb     = data.frame(
    year     = calendar_years,
    estimate = rep_lf$spawning_biomass_y[1:data$n_year],
    model    = "opal (with LF)"
  ),
  om = subset(opaka_truth$om$sprseries, Yr %in% calendar_years,
              select = c(Yr, SSB)),
  em = subset(opaka_truth$em$sprseries, Yr %in% calendar_years,
              select = c(Yr, SSB))
)

## ---- Bayesian + projection outputs appended to the bundle ----------------
paka$mcmc_diag        <- mcmc_diag
paka$mcmc_trace       <- mcmc_trace
paka$mcmc_mle_ref     <- mcmc_mle_ref
paka$n_chains_mcmc    <- n_chains_mcmc
paka$proj_hist_ribbon <- proj_hist_ribbon
paka$proj_ribbon      <- proj_ribbon
paka$proj_hist_mle    <- proj_hist_mle
paka$proj_om          <- proj_om
paka$proj_em          <- proj_em
paka$n_proj           <- n_proj
paka$n_iter           <- n_iter
paka$n_hist_show      <- n_hist_show

saveRDS(paka, paka_fit_path)

}   # end cache-miss branch
```

```{r}
#| label: load-bet-cache
#| include: false

bet_fit_candidates <- c(
  file.path(doc_dir, "cached", "bet", "opal_bet_fit.rds"),
  file.path(doc_dir, "assets", "cached", "bet", "opal_bet_fit.rds"),
  file.path(getwd(), "cached", "bet", "opal_bet_fit.rds"),
  file.path(getwd(), "assets", "cached", "bet", "opal_bet_fit.rds")
)
bet_fit_path <- bet_fit_candidates[file.exists(bet_fit_candidates)][1]
if (is.na(bet_fit_path)) {
  stop("BET cached results not found")
}
bet <- readRDS(bet_fit_path)
bet_final_sb_b0 <- tail(bet$report$spawning_biomass_y, 1) / bet$report$B0
bet_final_dyn_dep <- tail(bet$report$dynamic_depletion_y, 1)
bet_objective_components <- bet$diagnostics$objective_components
if (is.null(bet_objective_components)) {
  stop("BET objective components are not available in the cached results")
}
bet_objective_component_total <- sum(bet_objective_components)
if (abs(bet_objective_component_total - bet$opt$objective) > 1e-6) {
  stop("BET objective components do not sum to the optimizer objective")
}
format_duration <- function(seconds) {
  if (is.null(seconds) || length(seconds) == 0 || !is.finite(seconds)) {
    return("not recorded")
  }
  if (seconds < 90) {
    paste0(round(seconds, 1), " s")
  } else {
    paste0(round(seconds / 60, 1), " min")
  }
}
bet_mle_runtime_text <- format_duration(bet$diagnostics$mle_runtime_seconds)
bet_mle_iterations <- if (is.null(bet$diagnostics$mle_runtime_iterations)) {
  bet$opt$iterations
} else {
  bet$diagnostics$mle_runtime_iterations
}
bet_mle_function_evals <- if (is.null(bet$diagnostics$mle_runtime_evaluations)) {
  bet$opt$evaluations[["function"]]
} else {
  bet$diagnostics$mle_runtime_evaluations[["function"]]
}
bet_mle_gradient_evals <- if (is.null(bet$diagnostics$mle_runtime_evaluations)) {
  bet$opt$evaluations[["gradient"]]
} else {
  bet$diagnostics$mle_runtime_evaluations[["gradient"]]
}

bet_harmonic_mean <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (!length(x)) return(NA_real_)
  length(x) / sum(1 / x)
}
bet_sdnr_summary <- function(x, conf = 0.95) {
  x <- x[is.finite(x)]
  df <- length(x) - 1L
  est <- sd(x)
  if (df < 1L) {
    return(tibble(osa_sdnr = est, osa_sdnr_lci = NA_real_, osa_sdnr_hci = NA_real_))
  }
  alpha <- (1 - conf) / 2
  tibble(
    osa_sdnr = est,
    osa_sdnr_lci = sqrt(df * est^2 / qchisq(1 - alpha, df)),
    osa_sdnr_hci = sqrt(df * est^2 / qchisq(alpha, df))
  )
}
bet_normalise_composition <- function(x) {
  total <- sum(x, na.rm = TRUE)
  if (!is.finite(total) || total <= 0) return(rep(NA_real_, length(x)))
  x / total
}
bet_comp_residual_rows <- bind_rows(
  bet$diagnostics$lf_residuals,
  bet$diagnostics$wf_residuals
) %>%
  mutate(
    fishery_num = as.integer(sub("Fishery ", "", Fishery)),
    composition = factor(composition, levels = c("Length", "Weight"))
  )
bet_comp_cell_summary <- bet_comp_residual_rows %>%
  filter(is.finite(residual)) %>%
  group_by(composition, Fishery, fishery_num) %>%
  group_modify(~bind_cols(
    tibble(
      n_cells = nrow(.x),
      mean = mean(.x$residual),
      mar = median(abs(.x$residual)),
      max_abs = max(abs(.x$residual))
    ),
    bet_sdnr_summary(.x$residual)
  )) %>%
  ungroup() %>%
  select(
    composition, Fishery, fishery_num,
    n_cells,
    mean, osa_sdnr, osa_sdnr_lci, osa_sdnr_hci, mar, max_abs
  )
bet_comp_row_diagnostics <- bet_comp_residual_rows %>%
  group_by(composition, Fishery, fishery_num, row) %>%
  group_modify(~{
    obs <- bet_normalise_composition(.x$observed)
    pred <- bet_normalise_composition(.x$predicted)
    bin <- .x$bin
    n_eff <- .x$n_eff[1]
    if (anyNA(obs) || anyNA(pred) || !is.finite(n_eff) || n_eff <= 0) {
      return(tibble(
        n_eff = n_eff,
        mean_residual = NA_real_,
        mi_eff_n = NA_real_
      ))
    }
    obs_mean <- sum(obs * bin)
    pred_mean <- sum(pred * bin)
    var_mean <- (sum(pred * bin^2) - pred_mean^2) / n_eff
    mean_residual <- if (is.finite(var_mean) && var_mean > 0) {
      (obs_mean - pred_mean) / sqrt(var_mean)
    } else {
      NA_real_
    }
    denominator <- sum((obs - pred)^2)
    numerator <- sum(pred * (1 - pred))
    mi_eff_n <- if (is.finite(denominator) && denominator > 0) {
      numerator / denominator
    } else {
      Inf
    }
    tibble(
      n_eff = n_eff,
      mean_residual = mean_residual,
      mi_eff_n = mi_eff_n
    )
  }) %>%
  ungroup()
bet_comp_mean_summary <- bet_comp_row_diagnostics %>%
  group_by(composition, Fishery, fishery_num) %>%
  summarise(
    n_rows = n(),
    francis_sdnr = sd(mean_residual[is.finite(mean_residual)]),
    francis_n_multiplier = ifelse(
      is.finite(francis_sdnr) && francis_sdnr > 0,
      1 / francis_sdnr^2,
      NA_real_
    ),
    mean_input_n = mean(n_eff, na.rm = TRUE),
    median_input_n = median(n_eff, na.rm = TRUE),
    mi_harmonic_n = bet_harmonic_mean(mi_eff_n),
    mi_median_n = median(mi_eff_n[is.finite(mi_eff_n)], na.rm = TRUE),
    mi_over_mean_input_n = mi_harmonic_n / mean_input_n,
    .groups = "drop"
  )
bet_comp_diagnostic_summary <- bet_comp_cell_summary %>%
  left_join(
    bet_comp_mean_summary,
    by = c("composition", "Fishery", "fishery_num")
  ) %>%
  arrange(composition, fishery_num)
bet_comp_residual_source <- bet$diagnostics$residual_source
if (is.null(bet_comp_residual_source) || !length(bet_comp_residual_source)) {
  bet_comp_residual_source <- "compResidual::resMulti"
}
bet_comp_mean_input_n_range <- range(
  bet_comp_diagnostic_summary$mean_input_n,
  na.rm = TRUE
)
```



\newpage

# Introduction {#sec-introduction}

[`opal`](https://n-ducharmebarth-noaa.github.io/opal/) --- the **o**pen **p**opulation **a**ssessment **l**ibrary --- is an open-source, modular R package for fisheries stock assessment. The package is built on RTMB [@kristensen_rtmb_2023], an R interface to the Template Model Builder framework [@kristensen_tmb_2016], which provides automatic differentiation of the model objective function and supports gradient-based optimization and the Laplace approximation for models that include random effects. Writing the model in R rather than C++ keeps the entire model definition within a single language environment, which lowers the barrier to inspecting, modifying, and extending the underlying code.

`opal` builds directly upon the `sbt` R package developed by [Quantifish](https://www.quantifish.co.nz/) for the assessment of southern bluefin tuna under the Commission for the Conservation of Southern Bluefin Tuna [CCSBT\; @webber_sbt_2026], and draws on design influences from a number of existing assessment platforms, including the Stochastic Population over Regional Components model [SPoRC\; @cheng_span_2026], the Woods Hole Assessment Model [WHAM\; @stock_woods_2021; @miller_woods_2020], Stock Synthesis [SS3\; @methot_stock_2013], and MULTIFAN-CL [@fournier_multifan-cl_1998].

The intent of `opal` is to provide assessment scientists with a flexible modelling framework that supports the estimation of random effects; facilitates enhanced model diagnostics, including one-step-ahead (OSA) residuals, simulation self-testing, and posterior predictive checks; and integrates directly into a Bayesian estimation framework. These capabilities are described in @sec-features.

The package source code, function documentation, and worked examples are publicly available ([website](https://n-ducharmebarth-noaa.github.io/opal/), [code repository](https://github.com/n-ducharmebarth-noaa/opal), & [documentation repository](https://github.com/n-ducharmebarth-noaa/opal-documentation)), and the package is distributed under an open-source license. This paper describes the structure of the model (@sec-model-structure), its principal features (@sec-features), and two case studies that illustrate its application to a real-data tuna assessment and a simulated single-species assessment (@sec-case-studies).

# Model structure {#sec-model-structure}

`opal` estimates age-structured population dynamics and accommodates length-based processes. The population is tracked as numbers-at-age. Biological quantities --- weight, maturity, fecundity, natural mortality, and sex ratio --- may be supplied on either an age basis or a length basis: a vector defined over age classes is used directly, while a vector defined over length bins is converted to age through a probability-of-length-at-age (PLA) matrix. Selectivity is likewise defined on a length basis and mapped to age through the PLA. This structure allows population demographics to be specified in terms of either age or length, and allows the model to fit multiple data types: relative abundance indices (CPUE), length composition, and weight composition. Where time-varying or hierarchically structured processes are required, parameters may be specified as random effects and integrated out using the Laplace approximation provided by RTMB [@kristensen_tmb_2016].

The model is implemented as a single objective function, `opal_model()`, that returns the negative log-likelihood (NLL) for a given set of parameters and data. Rather than encoding the population dynamics and observation model as a single monolithic routine, `opal_model()` orchestrates a sequence of standalone module functions, each responsible for one component of the calculation. The principal modules, in the order they are evaluated, are: mean length-at-age (`get_growth()`); the standard deviation of length-at-age (`get_sd_at_age()`); the PLA matrix (`get_pla()`); weight-, maturity-, and fecundity-at-age (`get_weight_at_length()`, `get_maturity_at_age()`, and a resolver, `resolve_bio_vector()`, that accepts either age- or length-basis inputs); selectivity-at-age (`get_selectivity()`, with the parametric forms `sel_logistic()` and `sel_double_normal()`); initial equilibrium numbers-at-age (`get_initial_numbers()`); the forward population dynamics and harvest (the dynamics routine and `get_harvest_rate()`); recruitment (`get_recruitment()`); the data likelihoods (`get_cpue_like()`, `get_length_like()`, `get_weight_like()`); and the prior and penalty terms (`get_recruitment_prior()`, `evaluate_priors()`). The modular design that this structure reflects is described further in @sec-future-development.

## Population dynamics {#sec-population-dynamics}

The population is structured by year, season, and age, and may be configured for any number of fisheries.

### Growth and the probability of length-at-age {#sec-growth-pla}

Mean length-at-age, $\mu_a$, is computed from a Schnute parameterization of the von Bertalanffy growth function, anchored by the lengths $L_1$ and $L_2$ at two reference ages $A_1$ and $A_2$, which matches the SS3 growth pattern with a linearly interpolated coefficient of variation [@methot_stock_2013]:

$$
\mu_a = L_1 + (L_2 - L_1)\,\frac{1 - e^{-k(a - A_1)}}{1 - e^{-k(A_2 - A_1)}}.
$$

*Module: `get_growth()`*

The coefficient of variation is interpolated linearly with mean length between the two reference ages, and the standard deviation of length-at-age is its product with the mean:

$$
\mathrm{CV}_a = \mathrm{CV}_1 + \frac{\mu_a - L_1}{L_2 - L_1}\,(\mathrm{CV}_2 - \mathrm{CV}_1),
\qquad
\sigma_a = \mu_a\,\mathrm{CV}_a.
$$

*Module: `get_sd_at_age()`*

From $\mu_a$ and $\sigma_a$, the probability-of-length-at-age (PLA) matrix gives the probability that an individual of age $a$ falls within length bin $\ell$, assuming length-at-age is normally distributed. For contiguous bins with edges $\{e_\ell\}$, this is the difference of the normal cumulative distribution function $\Phi$ evaluated at the bin edges:

$$
P_{\ell a} = \Phi\!\left(\frac{e_{\ell+1} - \mu_a}{\sigma_a}\right) - \Phi\!\left(\frac{e_{\ell} - \mu_a}{\sigma_a}\right),
\qquad \sum_\ell P_{\ell a} = 1 .
$$

*Module: `get_pla()`*

The PLA is computed once per function evaluation and reused for the conversion of weight, maturity, fecundity, natural mortality, and selectivity between the length and age bases.

### Biology and spawning potential {#sec-biology}

Quantities defined on a length basis are mapped to age through the PLA. For a length-basis vector $x_\ell$, the age-basis vector is $x_a = \sum_\ell P_{\ell a}\, x_\ell$ (equivalently $\mathbf{x}_a = \mathbf{P}^{\top}\mathbf{x}_\ell$); a vector supplied on an age basis is used directly. Weight-at-length follows an allometric length-weight relationship, $w_\ell = \alpha_{\mathrm{LW}}\, m_\ell^{\beta_{\mathrm{LW}}}$, where $m_\ell$ is the bin midpoint, and is converted to weight-at-age through the PLA. Spawning potential at age is the product of sex ratio (the fraction mature-female-equivalent at age), maturity, and fecundity:

$$
\psi_a = r_a\, m_a\, f_a ,
$$

*Modules: `get_weight_at_length()`, `get_maturity_at_age()`, `resolve_bio_vector()`*

where $r_a$, $m_a$, and $f_a$ are sex ratio, maturity, and fecundity at age, respectively. Alternatively, a pre-computed spawning-potential vector may be supplied directly. All derived biology arrays are passed explicitly to the dynamics routine rather than read from the data object, so that automatic-differentiation gradients propagate correctly if any of these quantities are estimated (for example, growth parameters estimated through the PLA, or natural mortality specified as a function of length).

### Initial conditions {#sec-initial-conditions}

The population is initialized at the unfished equilibrium. Unfished survivorship-per-recruit accumulates natural mortality across ages with a plus-group at the oldest age, $\phi_a^{0} = \phi_{a-1}^{0}\,e^{-M_{a-1}}$ with $\phi_{n}^{0}$ divided by $1 - e^{-M_{n}}$, and the unfished spawners-per-recruit is $\Phi_0 = \sum_a \psi_a\, \phi_a^{0}$. Unfished recruitment and the Beverton-Holt parameters follow from unfished spawning biomass $B_0$ and steepness $h$:

$$
R_0 = \frac{B_0}{\Phi_0},
\qquad
a_{\mathrm{BH}} = \frac{4 h R_0}{5h - 1},
\qquad
b_{\mathrm{BH}} = \frac{B_0 (1 - h)}{5h - 1}.
$$

*Module: `get_initial_numbers()`*

When initial fishing mortality is specified, survivorship is recomputed with total mortality $Z_a = M_a + \sum_f F_f^{\mathrm{init}} s_{f,a}$, giving a fished spawners-per-recruit $\Phi_{\mathrm{eq}}$ and an equilibrium recruitment $R_{\mathrm{eq}} = a_{\mathrm{BH}} - b_{\mathrm{BH}}/\Phi_{\mathrm{eq}}$. Initial numbers-at-age are $N_a^{\mathrm{init}} = R_{\mathrm{eq}}\,\phi_a$, optionally modified by estimated initial age-structure deviations with the lognormal bias correction described below.

### Harvest and the forward projection {#sec-harvest}

The population is projected forward over years and seasons. Within each season, survival from natural mortality is $S_a = e^{-M_a / n_{\mathrm{season}}}$. Removals are taken as a harvest fraction rather than a continuous fishing-mortality rate. For each fishery with observed catch, a fully selected harvest rate is obtained by dividing the observed catch by the available biomass (or available numbers, depending on the catch units of the fishery),

$$
F_{f} = \frac{C_{f}}{\sum_a N_a\, s_{f,a}\, w_{f,a} + \varepsilon},
\qquad
h_{f,a} = F_{f}\, s_{f,a},
$$

*Module: `get_harvest_rate()`*

where $s_{f,a}$ is selectivity-at-age, $w_{f,a}$ is weight-at-age, and $\varepsilon$ is a small constant guarding the denominator. The total harvest rate at age is the sum over fisheries, $h_a = \sum_f h_{f,a}$, and predicted catch-at-age is $\hat{C}_{f,a} = h_{f,a} N_a$. To keep the combined harvest fraction feasible, the quantity $1 - \sum_f F_f$ is passed through a differentiable lower-bound function (`posfun`) that returns the argument where it exceeds a small threshold and otherwise smoothly approaches that threshold, adding the resulting penalty to the objective function.

Surviving numbers are advanced by age and season, with recruitment entering the first age class at the start of each year, and the plus-group accumulating survivors from the oldest two age classes. Spawning biomass is $\mathrm{SB}_y = \sum_a N_{y,a}\, \psi_a$. The routine returns numbers-at-age, predicted catch-at-age by fishery, harvest rates, and the spawning-biomass time series for both the exploited and the dynamic unfished populations; their ratio defines dynamic depletion, $\mathrm{SB}_y / \mathrm{SB}_y^{0}$.

### Recruitment {#sec-recruitment}

Recruitment follows a Beverton-Holt stock-recruitment relationship with annual log-normal deviations $\varepsilon_y$:

$$
R_y = \frac{a_{\mathrm{BH}}\, \mathrm{SB}_y}{b_{\mathrm{BH}} + \mathrm{SB}_y}\; \exp\!\left(\varepsilon_y - \tfrac{1}{2}\, b_y\, \sigma_r^2\right),
$$

*Module: `get_recruitment()`*

where $\sigma_r$ is the standard deviation of the deviations and $b_y$ is a year-specific bias-adjustment scalar. The lognormal bias-correction term $\tfrac{1}{2} b_y \sigma_r^2$ optionally can be modulated by a bias-adjustment ramp that reduces the correction in years where recruitment is poorly informed by the data, following the approach used in SS3 [@methot_stock_2013]. Recruitment deviations may be treated as fixed effects (penalized maximum likelihood) or as random effects.

## Likelihood components {#sec-likelihood-components}

The statistical model is written as a joint posterior density for the unknown parameters and latent quantities, conditional on the observed data. In implementation, `opal` evaluates the corresponding negative log posterior as the sum of the data log-likelihood contributions, recruitment-deviation density, parameter prior densities, and a soft harvest-rate constraint. Each data likelihood is controlled by a switch in the data object, so that any combination of CPUE, length composition, and weight composition observations may be included or excluded without altering the model code.

### Abundance indices {#sec-abundance-indices}

Relative abundance is represented as an observation model for the logarithm of each standardized index. The CPUE component (`get_cpue_like()`) supports one or more indices, each with its own catchability $q_j$, additional observation standard deviation $\tau_j$, a power parameter $\omega_j$ relating the index to vulnerable abundance, and a cumulative effort-creep term $\kappa_{ij}$. For observation $i$ in index $j$, the latent expected index is formed from the vulnerable abundance $\tilde{N}_i$ associated with that observation,

$$
\log \hat{I}_{ij} = \log q_j + \omega_j\, \log \tilde{N}_i + \log \kappa_{ij}.
$$

*Module: `get_cpue_like()`*

The expected log index is mean-centred within each index before comparison with the observations. Observation and process variability are combined in quadrature,

$$
\sigma_{ij}^2 = \sigma_{\mathrm{obs},ij}^2 + \tau_j^2,
$$

where $\sigma_{\mathrm{obs},ij}$ is the input standard error. The resulting observation model is

$$
\log I_{ij} \mid \hat{I}_{ij}, \sigma_{ij}^2
\sim \mathrm{Normal}\!\left(\log \hat{I}_{ij},\, \sigma_{ij}^2\right),
$$

and the CPUE contribution to the objective is the negative log density of these observations under this normal model.

### Composition data {#sec-composition-data}

Length and weight composition observations are represented as draws from probability vectors predicted by the population and observation model. Predicted compositions are obtained by converting predicted catch-at-age $\hat{C}_{f,a}$ to a length basis through the PLA (and, for weight composition, an additional length-to-weight rebinning matrix $\mathbf{G}$):

$$
\hat{p}_{f,\ell} \,\propto\, \sum_a P_{\ell a}\, \hat{C}_{f,a}
\qquad\text{(length)},
\qquad
\hat{p}_{f,k} \,\propto\, \sum_{\ell, a} G_{k\ell}\, P_{\ell a}\, \hat{C}_{f,a}
\qquad\text{(weight)} .
$$

*Modules: `get_length_like()`, `get_weight_like()`*

Predicted proportions are aggregated into the active bin range for the fishery, with the tails optionally accumulated into the minimum and maximum bins, then a small robustness constant is added and the vector is renormalized to sum to one. For observation $i$ of fishery $f$, let $\hat{\mathbf{p}}_{if}$ denote this predicted composition, $n_i$ the observed or effective sample size, $\mathbf{c}_{if}$ the observed count vector, and $\mathbf{o}_{if}$ the observed proportion vector. The available observation models are

$$
\mathbf{c}_{if} \mid \hat{\mathbf{p}}_{if}, n_i
\sim \mathrm{Multinomial}\!\left(n_i,\, \hat{\mathbf{p}}_{if}\right),
$$

for the multinomial option,

$$
\mathbf{o}_{if} \mid \hat{\mathbf{p}}_{if}, n_i, \tau_f
\sim \mathrm{Dirichlet}\!\left(\hat{\mathbf{p}}_{if}\, n_i\, e^{\tau_f}\right),
$$

for the Dirichlet option, and

$$
\mathbf{c}_{if} \mid \hat{\mathbf{p}}_{if}, n_i, \tau_f
\sim \mathrm{Dirichlet\text{-}Multinomial}\!\left(n_i,\, \hat{\mathbf{p}}_{if}\, e^{\tau_f}\right),
$$

for the Dirichlet-multinomial option. The parameter $\tau_f$ is a fishery-specific log-scale variance-adjustment parameter that controls the effective sample size or overdispersion. The composition contribution to the objective is the negative log density of the selected observation model. It is evaluated as a batched matrix operation over the observations of each fishery rather than a per-observation loop, which reduces the size of the automatic-differentiation tape.

### Priors and penalties {#sec-priors-penalties}

Prior information is incorporated through a flexible prior specification (`evaluate_priors()`) that applies parametric prior densities to named parameters. Recruitment deviations are treated in the same distributional style. By default,

$$
\varepsilon_y \mid \sigma_r \sim \mathrm{Normal}(0, \sigma_r^2),
$$

with the lognormal recruitment bias adjustment applied in the recruitment equation (@sec-recruitment). Initial age-structure deviations, when estimated, use the analogous normal density. These terms can be interpreted as penalties in an MLE fit or as prior/random-effect contributions in the joint posterior. This shared formulation allows the same model code to be optimized with `nlminb` or sampled as a Bayesian model (@sec-bayesian-inference).

# Features {#sec-features}

### Bayesian inference {#sec-bayesian-inference}

Prior information can be incorporated either through a penalized maximum likelihood estimation (MLE) approach, in which priors enter the objective function as penalty terms and the model is optimized with `nlminb`, or through direct Bayesian inference. Bayesian estimation is performed with `SparseNUTS` [@monnahan_sparsenuts_2026], which implements a sparse-preconditioned variant of the No-U-Turn Sampler that exploits the sparse precision structure of TMB models to improve sampling efficiency for high-dimensional, correlated, and hierarchically structured posteriors [@monnahan_leveraging_2026]. Since the `opal` objective function and its component modules are constructed in RTMB, the same model definition is used for both the MLE and the Bayesian workflows. Convergence is assessed with the usual Markov chain Monte Carlo (MCMC) diagnostics --- split-$\hat{R}$ and bulk and tail effective sample sizes --- and the resulting posterior draws are carried directly into the forward projections described below. For the 'opakapaka case study (@sec-case-paka), a dense-metric sampler run for eight chains mixed well across the full parameter vector (@tbl-paka-mcmc-diag, @fig-paka-mcmc-trace).

### Projections {#sec-projections}

`opal` supports catch-based forward projection of the population from an estimated set of parameters. A projection is specified by a number of projection years, a future catch scenario, and a number of uncertainty iterations, and is run with `project_dynamics()` together with the helper functions `project_rec_devs()` and `project_selectivity()`. Future recruitment deviations are simulated from an ARIMA model fitted to the estimated recruitment-deviation series (or, under the Bayesian workflow, to each posterior draw's own deviation history), and future selectivity is resampled from the most recent estimated years, so that both process components carry forward the variability seen in the historical period.

Projections are conditioned on the specified future catch and propagate parameter uncertainty into the projected quantities. Two approaches to characterizing that uncertainty are available: resampling from a multivariate normal (MVN) approximation to the joint parameter distribution at the mode, and propagation of a posterior sample obtained by MCMC (@sec-bayesian-inference). Under the MVN approach the scalar parameters are drawn from the Hessian-based covariance while the recruitment and initial-age deviations are set to their conditional expectation given those draws. This preserves the estimated correlation between population scale and the deviations and avoids a Jensen-inequality inflation of mean biomass, so that the uncertainty in the terminal (projection-start) state coincides with the delta-method uncertainty of the historical trajectory. Both approaches yield distributions of projected spawning biomass and depletion under the specified catch scenario (@fig-paka-projection).

### Model diagnostics {#sec-model-diagnostics}

`opal` provides a set of built-in diagnostics intended to support model evaluation and validation. These include estimability checks and parameter-correlation summaries derived from the fitted model; simulation self-testing, in which data are simulated from the model and re-estimated to confirm that parameters can be recovered; posterior-predictive checking where the observations are compared to the simulated distributions of the predictions given the model and one-step-ahead (OSA) residuals, computed through RTMB's `oneStepPredict()`, which provide residuals that are independent and identically distributed under a correctly specified model and are therefore more readily interpretable than ordinary residuals for non-Gaussian observations [@dunn_randomized_1996; @thygesen_validation_2017; @kristensen_tmb_2016]. The simulation and posterior-predictive capabilities also support the projection and Bayesian workflows described above.

# Case studies {#sec-case-studies}

Two case studies are distributed with the package as worked examples. The first applies `opal` to real data from an existing WCPFC tuna assessment; the second applies it to simulated data for which the underlying population dynamics are known. Detailed configuration, code, and outputs for each are provided in the package vignettes.

## 2020 western and central Pacific Ocean bigeye tuna {#sec-case-bet}

The first case study uses the data from a fleets-as-areas model described in the appendix of the 2020 assessment of bigeye tuna (*Thunnus obesus*) in the western and central Pacific Ocean [WCPO\;@ducharme-barth_stock_2020]. The bundled data object contains the biological parameters, catch and standardized CPUE observations, length and weight composition data, configured at a quarterly time step across the fisheries defined in the fleets-as-areas assessment, with selectivity specified by fishery as either logistic or double-normal forms.

The current diagnostic configuration uses the quarterly standardized CPUE series as four seasonal indices, fits length compositions from fisheries 8--14, and fits weight compositions from fisheries 1, 2, 3, 4, 6, 7, and 15. Fishery 8 length compositions are retained but strongly downweighted, and weight-composition fishery 5 remains inactive after diagnostic runs failed to converge cleanly. Recruitment deviations are estimated across the model period, catchability is estimated separately by index, and selected selectivity parameters are estimated for the fisheries with active composition data. Growth, natural mortality, steepness, recruitment variability, effort creep, and the remaining selectivity parameters are fixed at the bundled reference values.

The fit converged according to `nlminb`, with the constrained KKT-style gradient close to zero despite a larger raw maximum gradient caused by active bounds (@tbl-bet-fit-summary). Fitting this diagnostic configuration from the configured starting values to the maximum-likelihood estimate took `r bet_mle_runtime_text` on the workstation used to build this document, requiring `r bet_mle_iterations` optimizer iterations, `r bet_mle_function_evals` function evaluations, and `r bet_mle_gradient_evals` gradient evaluations. The additive objective decomposition is dominated numerically by the composition likelihoods, especially the weight-composition component, while the CPUE component contributes the abundance trend information and the recruitment-deviation density regularizes annual departures from the stock-recruitment curve. The estimated unfished spawning biomass is approximately `r format(round(bet$report$B0), big.mark = ",")` mt, and the final-year relative spawning biomass is `r round(bet_final_sb_b0, 3)` when scaled by $SB_0$ (@tbl-bet-depletion). The dynamic depletion ratio is slightly higher (`r round(bet_final_dyn_dep, 3)`) because it uses the time-varying unfished spawning biomass denominator rather than the static $SB_0$ denominator.

Diagnostic plots in the `opal` BET vignette compare observed and predicted CPUE by seasonal index, observed and predicted length and weight compositions, fitted recruitment deviations, and estimated selectivity-at-length and selectivity-at-age curves. The CPUE fit in @fig-bet-cpue-fit tracks the broad decline in standardized abundance from the early 1950s through the low-abundance period in the 2000s, while allowing the four quarterly indices to retain their own scale and observation uncertainty; simulated CPUE series from the fitted model show the amount of variation expected under the fitted lognormal observation model. The CPUE OSA residuals in @fig-bet-osa-residuals are centred close to zero and fall within approximately $\pm 2$ residual units, with SDNRs below one for each index month (@tbl-bet-cpue-sdnr). These residuals use the randomized quantile-residual logic underlying OSA diagnostics for latent-variable models, so approximate standard-normal behaviour is expected when the observation model is adequate [@dunn_randomized_1996; @thygesen_validation_2017]. The mean length- and weight-composition diagnostics in @fig-bet-lf-diagnostics and @fig-bet-wf-diagnostics aggregate observations, predictions, and simulations over all years for each active composition fishery; the simulated lines are generated on the same mean-composition scale as the red fitted line, so their average is centred on the fitted composition. To show the unaggregated temporal structure, @fig-bet-lf12-year-composition and @fig-bet-wf15-year-composition give year-specific examples for length-composition fishery 12 and weight-composition fishery 15. Example LF and WF composition OSA residual diagnostics are shown for fishery 12 and fishery 15 in @fig-bet-lf-composition-residuals and @fig-bet-wf-composition-residuals; these two panels are examples rather than the full residual set. The by-fishery Q-Q diagnostics in @fig-bet-composition-residual-qq summarise all active composition residuals against a fixed standard-normal reference line, rather than a fitted Q-Q line, so below-one dispersion remains visible. Following the SBT diagnostic summaries, @tbl-bet-composition-sdnr separates the bin-level OSA residual SDNR from the Francis mean-composition SDNR and McAllister-Ianelli effective-sample-size diagnostics. The OSA residuals are generated with `r bet_comp_residual_source` for the active multinomial composition likelihoods, and the near-one bin-level SDNRs are conditional on the variance-adjusted input sample sizes used in the likelihood; the mean input $N$ by fishery ranges from `r signif(bet_comp_mean_input_n_range[1], 3)` to `r signif(bet_comp_mean_input_n_range[2], 3)` after downweighting. The close Q-Q alignment is therefore a check of the randomized bin-level residual distribution for the weighted pseudo-data likelihood, not evidence that the raw composition data are perfectly fitted.

The selectivity diagnostics in @tbl-bet-selectivity-issues and @fig-bet-selectivity-length show where the fitted `opal` selectivity curves differ from the SS3 reference curves and where estimated terms are constrained by bounds. Most flagged issues are bound-related rather than unconstrained interior failures, which is consistent with the small KKT-style gradient in @tbl-bet-fit-summary despite the larger raw maximum gradient. The recruitment deviations in @fig-bet-recruitment-deviations show the seasonal pattern of estimated process variation. The biomass and depletion trajectories in @fig-bet-biomass show the long-term decline in spawning biomass relative to unfished conditions and the partial separation between $SB/SB_0$ and dynamic depletion. As with the 'opakapaka example, these results are intended to demonstrate the end-to-end workflow rather than to replace a full production assessment model.

## Hawai'i 'Opakapaka {#sec-case-paka}

The second case study uses simulated data for Hawai'i 'opakapaka (*Pristipomoides filamentosus*), a deepwater snapper, for which the true population dynamics are known. The reference model was constructed in SS3 [v3.30.19\; @methot_stock_2013] as part of an associated simulation study [@oshima_opakapaka_2026], which provides both an operating model that generates the data and an estimation model fitted to those data. The `opal` model covers the historical period 1949--2023 with three fleets --- a commercial fishery, a non-commercial fishery, and a research-fishery survey --- at an annual time step, and fits to two CPUE indices, generated with negligible observation error, together with length composition data from the commercial and survey fleets.

The case study is configured to estimate population scale, annual recruitment deviations, initial age structure, and selectivity parameters for the fleets that have length composition data, with the remaining life-history and fishery parameters fixed at the values used to generate the simulated data. Because the operating model is built in SS3, the case study also serves as a cross-platform comparison: it provides a basis for evaluating the correspondence between the `opal` and SS3 population-dynamics calculations under identical inputs, and for comparing the spawning-biomass trajectory and parameter estimates from `opal` against the operating-model truth and the SS3 estimation model.

Fitting the length composition likelihood and freeing selectivity for the commercial and survey fleets, the resulting estimated spawning-biomass trajectory is compared against the SS3 operating and estimation models in
@fig-paka-lf-biomass. Model convergence is summarised by @tbl-paka-lf-convergence.

The same model was also fitted with the Bayesian workflow (@sec-bayesian-inference): a dense-metric NUTS sampler was run for eight chains, which were well mixed with split-$\hat{R}$ close to one and large effective sample sizes for the scalar parameters and recruitment deviations (@tbl-paka-mcmc-diag, @fig-paka-mcmc-trace). The stock was then projected ten years forward under a constant catch held at the terminal year's level, propagating parameter uncertainty through both the MVN and MCMC approaches (@sec-projections). The two approaches agree closely at the projection-start state, and the projection intervals widen only gradually because 'opakapaka is long-lived and matures slowly, so the existing adult biomass declines nearly deterministically under constant catch while future recruitment and selectivity uncertainty accumulate over the projection horizon (@fig-paka-projection).

# Discussion {#sec-discussion}

`opal` provides an age-structured, length-based stock assessment framework implemented in RTMB, with a modular internal structure, support for multiple data types and likelihood forms, random-effects estimation, catch-based projections, Bayesian inference, and a set of built-in diagnostics. The two case studies demonstrate its application to both a real-data tuna assessment configured as an age-structured production model and a simulated single-species assessment that fits to index and length composition data and supports comparison against a known operating model.

## Future development {#sec-future-development}

`opal` is constructed as a set of standalone, individually documented module functions that are combined within the `opal_model()` objective function. Population-model components --- growth, selectivity, recruitment, and the data likelihoods --- are implemented as separate exported functions rather than embedded in a single routine, so that components can be modified, replaced, or extended independently, and so that individual modules can be tested in isolation. The collection of module functions is registered through a single accessor (`bet_globals()`) for use when the model is passed to the Bayesian sampler. This modular organization is intended to make the codebase tractable to maintain and to lower the barrier to community contribution.

Software development follows a defined contribution workflow. The package uses a two-branch structure: a `dev` branch for active development and a stable `main` branch for tested, release-ready code. New features are developed on branches forked from `dev` and are expected to include roxygen function documentation and minimal, self-contained `testthat` unit tests. Contributions are submitted as pull requests against `dev`, and are incorporated into `main` only after they pass the package checks --- that is, after all tests pass and the package and its vignettes can be built.

Integrated testing operates at several levels. Unit tests verify the behaviour of individual modules; for example, the growth tests confirm that `get_growth()` returns the reference lengths at the reference ages and is monotonic, and that the weight, maturity, and PLA conversions behave as expected. Integration tests verify that the modules combine correctly within the full `opal_model()` objective function and that the objective and its gradient are finite when particular data and likelihood components are active. A numerical-regression baseline (`opal_baseline`), produced and checked through a dedicated vignette, stores a complete reference evaluation of the bigeye case-study model --- the objective value, the gradient, the report output, and timing --- so that a refactored build can be compared against the reference to detect unintended numerical changes and performance regressions. Continuous-integration checks (`R-CMD-check`) are run automatically on the repository.

Several extensions are planned. Priorities identified for ongoing work include refinement of the case studies; code optimization, particularly of the selectivity and size-composition likelihood calculations; and additional functionality such as biological and fishery reference points, alternative selectivity forms, and conditional age-at-length data. Within the WCPFC context, planned work also includes a side-by-side comparison of MULTIFAN-CL [@fournier_multifan-cl_1998], SS3 [@methot_stock_2013], and `opal` for a WCPO bigeye tuna fleets-as-areas model, an Indian Ocean yellowfin tuna fleets-as-areas case study based on simulated data, and potential application to the 2027 North Pacific blue shark stock assessment.

# Acknowledgements {#sec-acknowledgements}

Development of opal is intended to be an open and collaborative effort, inclusive to the broader fisheries assessment community. To that end we are grateful to the many people that contributed to the project through informal discussions, and participation in periodic update meetings. While not an exhaustive list, this community includes: N. Davies (Te Takina Ltd), S. Hoyle (Hoyle Consulting), M. Maunder (Inter-American Tropical Tuna Commission), R. Bi (IATTC), F. Carvalho (PIFSC), G. Pilling (SPC), J. Hampton (SPC *emeritus*), and many others including participants in the 2025 CAPAM mini workshop on developing the next generation tuna stock assessment model as well as the 2026 SPC Pre-Assessment Workshop.

# Declaration of generative AI use {#sec-ai-declaration}

Generative AI tools were used to assist with code development, refactoring, and documentation, and with drafting and editing text in this document. All analytical choices, model specifications, diagnostic interpretations, and conclusions are those of the authors, who reviewed and verified all AI-assisted output.

# References {#sec-references}

::: {#refs}
:::

\newpage

# Tables {#sec-tables}

```{r}
#| label: tbl-bet-fit-summary
#| tbl-cap: "Optimization and objective-component summary for the fitted WCPO bigeye tuna diagnostic model. The Max KKT gradient is the maximum projected gradient after accounting for active parameter bounds. Objective components were computed as AD-objective switch increments at the fitted parameter vector and sum to the optimizer objective."
#| echo: false

format_bet_value <- function(x, digits = 2) {
  format(round(x, digits), big.mark = ",", nsmall = digits)
}

bet_fit_tbl <- data.frame(
  Metric = c(
    "nlminb convergence",
    "Objective",
    "Max |gradient|",
    "Max KKT gradient",
    "B0",
    "Prior contribution",
    "Recruitment-deviation contribution",
    "CPUE contribution",
    "Length-composition contribution",
    "Weight-composition contribution",
    "Harvest penalty",
    "Objective-component total"
  ),
  Value = c(
    bet$opt$convergence,
    format_bet_value(bet$opt$objective),
    round(bet$raw_grad, 4),
    round(bet$kkt_grad, 4),
    format(round(bet$report$B0), big.mark = ","),
    format_bet_value(bet_objective_components["Prior contribution"]),
    format_bet_value(bet_objective_components["Recruitment-deviation contribution"]),
    format_bet_value(bet_objective_components["CPUE contribution"]),
    format_bet_value(bet_objective_components["Length-composition contribution"]),
    format_bet_value(bet_objective_components["Weight-composition contribution"]),
    format_bet_value(bet_objective_components["Harvest penalty"]),
    format_bet_value(bet_objective_component_total)
  ),
  check.names = FALSE
)

knitr::kable(bet_fit_tbl, booktabs = TRUE, align = c("l", "l"))
```

```{r}
#| label: tbl-bet-depletion
#| tbl-cap: "Final-year relative spawning-biomass diagnostics for the fitted WCPO bigeye tuna diagnostic model."
#| echo: false

bet_depletion_tbl <- data.frame(
  Metric = c("SB/SB0", "SB/SB_F0"),
  Value = round(c(bet_final_sb_b0, bet_final_dyn_dep), 3),
  check.names = FALSE
)

knitr::kable(bet_depletion_tbl, booktabs = TRUE, align = c("l", "r"))
```

```{r}
#| label: tbl-bet-cpue-sdnr
#| tbl-cap: "Standard deviation of normalized residuals (SDNR) and median absolute residual (MAR) for the CPUE OSA residuals in the fitted WCPO bigeye tuna diagnostic model."
#| echo: false

bet_cpue_sdnr_tbl <- bet$diagnostics$cpue_sdnr %>%
  transmute(
    `Index month` = index,
    n,
    SDNR = round(SDNR, 3),
    `95% LCI` = round(LCI, 3),
    `95% HCI` = round(HCI, 3),
    MAR = round(MAR, 3)
  )

knitr::kable(bet_cpue_sdnr_tbl, booktabs = TRUE, align = c("l", "r", "r", "r", "r", "r"))
```

```{r}
#| label: tbl-bet-composition-sdnr
#| tbl-cap: "Composition residual diagnostics by composition type and fishery for the fitted WCPO bigeye tuna diagnostic model. OSA SDNR is the standard deviation of finite bin-level OSA residuals returned by compResidual::resMulti(); the 95% interval uses the same chi-square SDNR interval as the SBT diagnostics. Francis and McAllister-Ianelli columns follow the SBT composition-diagnostic summaries."
#| echo: false

bet_comp_sdnr_tbl <- bet_comp_diagnostic_summary %>%
  transmute(
    Composition = composition,
    Fishery,
    Rows = n_rows,
    Cells = n_cells,
    `OSA SDNR` = round(osa_sdnr, 3),
    `OSA 95% CI` = sprintf("%.3f-%.3f", osa_sdnr_lci, osa_sdnr_hci),
    MAR = round(mar, 3),
    `Francis SDNR` = round(francis_sdnr, 3),
    `N multiplier` = round(francis_n_multiplier, 2),
    `Mean input N` = round(mean_input_n, 2),
    `MI harmonic N` = round(mi_harmonic_n, 1),
    `MI/input N` = round(mi_over_mean_input_n, 2)
  )

bet_comp_sdnr_kable <- knitr::kable(
  bet_comp_sdnr_tbl,
  format = if (knitr::is_latex_output()) "latex" else "html",
  booktabs = TRUE,
  align = c("l", "l", rep("r", 10))
)
if (knitr::is_latex_output()) {
  knitr::asis_output(paste0(
    "\\begingroup\\scriptsize\\setlength{\\tabcolsep}{3pt}\n",
    bet_comp_sdnr_kable,
    "\n\\endgroup"
  ))
} else {
  bet_comp_sdnr_kable
}
```

```{r}
#| label: tbl-bet-selectivity-issues
#| tbl-cap: "Selectivity-parameter diagnostics by fishery for the fitted WCPO bigeye tuna diagnostic model. Flags identify estimated selectivity terms that are on parameter bounds; these bound checks help explain large raw gradients when the constrained KKT gradient is small."
#| echo: false

bet_sel_raw <- bet$diagnostics$selectivity_raw
sel_short_component <- c(
  "peak", "top", "asc. width", "desc. width", "initial sel.", "final sel."
)
bet_sel_map <- bet_sel_raw$map_sel
bet_sel_pos <- which(names(bet_sel_raw$opt_par) == "par_sel")
bet_sel_level_pos <- setNames(bet_sel_pos, as.character(seq_along(bet_sel_pos)))

bet_sel_diag <- map_dfr(seq_len(nrow(bet_sel_map)), function(f) {
  map_dfr(seq_len(ncol(bet_sel_map)), function(p) {
    level <- as.character(bet_sel_map[f, p])
    if (is.na(level)) return(tibble())
    idx <- bet_sel_level_pos[[level]]
    estimate <- bet_sel_raw$opt_par[idx]
    lower <- bet_sel_raw$bounds$lower[idx]
    upper <- bet_sel_raw$bounds$upper[idx]
    bound_status <- case_when(
      is.finite(lower) && estimate <= lower + 1e-7 ~ "lower",
      is.finite(upper) && estimate >= upper - 1e-7 ~ "upper",
      TRUE ~ "free"
    )
    tibble(
      fishery = f,
      Form = if_else(bet_sel_raw$sel_type_f[f] == 1L, "logistic", "double-normal"),
      term = sel_short_component[p],
      bound_status = bound_status
    )
  })
})

bet_selectivity_issues_tbl <- map_dfr(
  sort(as.integer(names(bet_sel_raw$sel_specs))),
  function(f) {
    flags <- bet_sel_diag %>%
      filter(fishery == f, bound_status != "free")
    tibble(
      Fishery = f,
      Form = unique(bet_sel_diag$Form[bet_sel_diag$fishery == f]),
      `Estimated terms` = paste(sel_short_component[bet_sel_raw$sel_specs[[as.character(f)]]],
                                collapse = ", "),
      `Flagged selectivity issue(s)` = if (nrow(flags) == 0L) {
        "None flagged"
      } else {
        paste(paste0(flags$term, " at ", flags$bound_status, " bound"),
              collapse = "; ")
      }
    )
  }
)

knitr::kable(
  bet_selectivity_issues_tbl,
  booktabs = TRUE,
  align = c("r", "l", "l", "l")
)
```

```{r}
#| label: tbl-paka-lf-convergence
#| tbl-cap: "opal (with LF) convergence diagnostics: positive-definite Hessian, maximum absolute gradient (mgc), minimum Hessian eigenvalue, and parameter estimability."
#| echo: false

est <- paka$estim

if (inherits(est, "condition")) {
  # check_estimability() failed and the condition was cached in Block B.
  min_eig_chr  <- "\u2014"
  estim_status <- paste0("check failed: ", conditionMessage(est))
} else {
  n_bad       <- length(est$WhichBad)
  min_eig_chr <- formatC(min(est$Eigen$values), format = "e", digits = 2)
  estim_status <- if (n_bad == 0) {
    "0 (all parameters estimable)"
  } else {
    bad_names <- rownames(est$Hess)[est$WhichBad]
    paste0(n_bad, " (", paste(bad_names, collapse = ", "), ")")
  }
}

conv_tbl <- data.frame(
  Diagnostic = c("Positive-definite Hessian",
                 "Max |gradient| (mgc)",
                 "Min Hessian eigenvalue",
                 "Inestimable parameters"),
  Value = c(as.character(paka$pdh),
            paka$mgc_sci,
            min_eig_chr,
            estim_status),
  check.names = FALSE
)

knitr::kable(conv_tbl, booktabs = TRUE, align = c("l", "l"))
```

```{r}
#| label: tbl-paka-mcmc-diag
#| tbl-cap: "Posterior summaries and MCMC convergence diagnostics for the 'opakapaka opal (with LF) model: the MLE estimate, posterior mean and standard deviation, 5th and 95th percentiles, bulk and tail effective sample sizes (ESS), and split-Rhat, for the scalar parameters (unfished spawning biomass and the two CPUE catchabilities) and three representative recruitment deviations."
#| echo: false

mdiag <- paka$mcmc_diag
knitr::kable(
  mdiag[, c("variable", "mle", "mean", "sd", "q5", "q95",
            "ess_bulk", "ess_tail", "rhat")],
  col.names = c("Parameter", "MLE", "Mean", "SD", "q5", "q95",
                "ESS bulk", "ESS tail", "Rhat"),
  digits = 3, booktabs = TRUE, align = c("l", rep("r", 8))
)
```

\newpage

# Figures {#sec-figures}


```{r}
#| echo: false
#| label: fig-bet-cpue-fit
#| fig-cap: "Observed, predicted, and simulated standardized CPUE for the fitted WCPO bigeye tuna diagnostic model, shown by index month. Vertical error bars show one fitted standard deviation on the log scale; grey lines show fitted-model simulations."
#| fig-height: 4.5
#| fig-width: 8

bet_cpue <- bet$data$cpue_data %>%
  mutate(
    index = factor(month,
                   levels = sort(unique(month)),
                   labels = month.name[sort(unique(month))]),
    pred = bet$report$cpue_pred,
    sigma = bet$report$cpue_sigma,
    lower = exp(log(value) - sigma),
    upper = exp(log(value) + sigma)
  )

bet_cpue_sim <- bet$report$cpue_sim %>%
  mutate(
    index = factor(month,
                   levels = sort(unique(month)),
                   labels = month.name[sort(unique(month))]),
    simulated = exp(simulated_log)
  )

ggplot(bet_cpue, aes(x = year)) +
  geom_line(data = bet_cpue_sim,
            aes(y = simulated, group = sim_id),
            colour = "grey45", alpha = 0.42, linewidth = 0.32) +
  geom_line(aes(y = pred), colour = "red3", linewidth = 0.8) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.35,
                colour = "grey20", alpha = 0.85, linewidth = 0.3) +
  geom_point(aes(y = value), shape = 21, size = 1.25,
             fill = "white", colour = "black", stroke = 0.25) +
  facet_wrap(~ index, scales = "free_y", ncol = 2) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Year", y = "Standardized CPUE") +
  theme_minimal()
```

```{r}
#| echo: false
#| label: fig-bet-osa-residuals
#| fig-cap: "One-step-ahead residual diagnostics for the CPUE component of the fitted WCPO bigeye tuna diagnostic model, shown by index month."
#| fig-height: 5
#| fig-width: 8

bet_osa <- bet$data$cpue_data %>%
  mutate(
    residual = bet$report$osa_cpue_residual,
    index = factor(month,
                   levels = sort(unique(month)),
                   labels = month.name[sort(unique(month))])
  )

bet_osa_qq <- bet_osa %>%
  group_by(index) %>%
  arrange(residual, .by_group = TRUE) %>%
  mutate(
    q_theoretical = qnorm(ppoints(n())),
    panel = "Normal Q-Q"
  ) %>%
  ungroup() %>%
  transmute(index, panel, x = q_theoretical, y = residual)

bet_osa_time <- bet_osa %>%
  transmute(index, panel = "Residual by year", x = year, y = residual)

ggplot(bind_rows(bet_osa_qq, bet_osa_time), aes(x = x, y = y)) +
  geom_hline(yintercept = 0, colour = "grey55", linewidth = 0.35) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed",
             colour = "grey55", linewidth = 0.35) +
  geom_abline(data = data.frame(panel = "Normal Q-Q"),
              aes(intercept = 0, slope = 1),
              inherit.aes = FALSE, colour = "grey55", linewidth = 0.35) +
  geom_segment(data = bet_osa_time,
               aes(x = x, xend = x, y = 0, yend = y),
               inherit.aes = FALSE, colour = "grey65", linewidth = 0.25) +
  geom_point(size = 0.9, alpha = 0.75, colour = "black") +
  facet_grid(index ~ panel, scales = "free_x") +
  labs(x = NULL, y = "OSA residual") +
  theme_minimal()
```

```{r}
#| echo: false
#| label: fig-bet-lf-diagnostics
#| fig-cap: "Mean observed length-composition proportions for WCPO bigeye tuna fisheries 8-14, aggregated over all years, with fitted predictions and simulated replicate lines shown for the fisheries included in the diagnostic length-composition likelihood."
#| fig-height: 7
#| fig-width: 8

bet_lf_pred <- bet$report$lf_pred
names(bet_lf_pred) <- bet$data$lf_fishery_f
for (k in seq_along(bet_lf_pred)) {
  colnames(bet_lf_pred[[k]]) <- bet$data$len_mid
}

bet_lf_pred_df <- map_dfr(seq_along(bet_lf_pred), function(i) {
  as.data.frame.table(bet_lf_pred[[i]], responseName = "proportion",
                      stringsAsFactors = FALSE) %>%
    rename(length = Var2) %>%
    mutate(
      fishery = paste("Fishery", names(bet_lf_pred)[i]),
      length = as.numeric(length),
      series = "Predicted"
    )
})

bet_lf_sim_df <- map_dfr(seq_along(bet$report$lf_sim_mean), function(s) {
  map_dfr(seq_along(bet$report$lf_sim_mean[[s]]), function(i) {
    tibble(
      fishery = paste("Fishery", bet$data$lf_fishery_f[i]),
      length = bet$data$len_mid,
      proportion = as.numeric(bet$report$lf_sim_mean[[s]][[i]]),
      sim_id = s
    )
  })
})

bet_lf_obs_df <- bet$data$wcpo_bet_lf %>%
  group_by(fishery, year, month, ts, bin) %>%
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  group_by(fishery, year, month, ts) %>%
  mutate(proportion = value / sum(value, na.rm = TRUE)) %>%
  ungroup() %>%
  transmute(
    fishery = paste("Fishery", fishery),
    length = bin,
    proportion,
    series = "Observed"
  )

bet_lf_mean <- bind_rows(bet_lf_obs_df, bet_lf_pred_df) %>%
  group_by(fishery, length, series) %>%
  summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop")

bet_fishery_levels <- paste("Fishery", sort(unique(bet$data$wcpo_bet_lf$fishery)))
bet_lf_mean <- bet_lf_mean %>%
  mutate(fishery = factor(fishery, levels = bet_fishery_levels))
bet_lf_sim_df <- bet_lf_sim_df %>%
  mutate(fishery = factor(fishery, levels = bet_fishery_levels))

ggplot() +
  geom_col(data = filter(bet_lf_mean, series == "Observed"),
           aes(x = length, y = proportion), fill = "grey75", width = 1.8) +
  geom_line(data = bet_lf_sim_df,
            aes(x = length, y = proportion, group = sim_id),
            colour = "steelblue", alpha = 0.35, linewidth = 0.35) +
  geom_line(data = filter(bet_lf_mean, series == "Predicted"),
            aes(x = length, y = proportion), colour = "red3", linewidth = 0.8) +
  facet_wrap(~ fishery, ncol = 2, scales = "free_y") +
  labs(x = "Length (cm)", y = "Mean proportion") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()
```

```{r}
#| echo: false
#| label: fig-bet-wf-diagnostics
#| fig-cap: "Mean observed, predicted, and simulated weight-composition proportions for active WCPO bigeye tuna weight-composition fisheries, aggregated over all years."
#| fig-height: 7
#| fig-width: 8

bet_wf_pred <- bet$report$wf_pred
names(bet_wf_pred) <- bet$data$wf_fishery_f
for (k in seq_along(bet_wf_pred)) {
  colnames(bet_wf_pred[[k]]) <- bet$data$wt_mid
}

bet_wf_pred_df <- map_dfr(seq_along(bet_wf_pred), function(i) {
  as.data.frame.table(bet_wf_pred[[i]], responseName = "proportion",
                      stringsAsFactors = FALSE) %>%
    rename(weight = Var2) %>%
    mutate(
      fishery = paste("Fishery", names(bet_wf_pred)[i]),
      weight = as.numeric(weight),
      series = "Predicted"
    )
})

bet_wf_sim_df <- map_dfr(seq_along(bet$report$wf_sim_mean), function(s) {
  map_dfr(seq_along(bet$report$wf_sim_mean[[s]]), function(i) {
    tibble(
      fishery = paste("Fishery", bet$data$wf_fishery_f[i]),
      weight = bet$data$wt_mid,
      proportion = as.numeric(bet$report$wf_sim_mean[[s]][[i]]),
      sim_id = s
    )
  })
})

bet_wf_obs_df <- bet$data$wcpo_bet_wf %>%
  filter(fishery %in% bet$data$wf_fishery_f) %>%
  group_by(fishery, year, month, ts, bin) %>%
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  group_by(fishery, year, month, ts) %>%
  mutate(proportion = value / sum(value, na.rm = TRUE)) %>%
  ungroup() %>%
  transmute(
    fishery = paste("Fishery", fishery),
    weight = bin,
    proportion,
    series = "Observed"
  )

bet_wf_mean <- bind_rows(bet_wf_obs_df, bet_wf_pred_df) %>%
  group_by(fishery, weight, series) %>%
  summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop")

bet_wf_fishery_levels <- paste("Fishery", sort(unique(bet$data$wf_fishery_f)))
bet_wf_mean <- bet_wf_mean %>%
  mutate(fishery = factor(fishery, levels = bet_wf_fishery_levels))
bet_wf_sim_df <- bet_wf_sim_df %>%
  mutate(fishery = factor(fishery, levels = bet_wf_fishery_levels))

ggplot() +
  geom_col(data = filter(bet_wf_mean, series == "Observed"),
           aes(x = weight, y = proportion), fill = "grey75", width = 1.8) +
  geom_line(data = bet_wf_sim_df,
            aes(x = weight, y = proportion, group = sim_id),
            colour = "steelblue", alpha = 0.35, linewidth = 0.35) +
  geom_line(data = filter(bet_wf_mean, series == "Predicted"),
            aes(x = weight, y = proportion), colour = "red3", linewidth = 0.8) +
  facet_wrap(~ fishery, ncol = 2, scales = "free_y") +
  labs(x = "Weight (kg)", y = "Mean proportion") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()
```

```{r}
#| echo: false
#| include: false
#| label: prep-bet-year-composition

select_panel_years <- function(years, max_years) {
  years <- sort(unique(years))
  if (length(years) <= max_years) return(years)
  years[unique(round(seq(1, length(years), length.out = max_years)))]
}

simulate_year_composition <- function(pred_year, obs_year, bin_col, nsim = 12L) {
  years <- sort(unique(pred_year$year))
  map_dfr(seq_len(nsim), function(s) {
    map_dfr(years, function(y) {
      pdat <- pred_year %>% filter(year == y) %>% arrange(.data[[bin_col]])
      p <- pdat$proportion
      p[!is.finite(p) | p < 0] <- 0
      p <- if (sum(p) > 0) p / sum(p) else rep(1 / length(p), length(p))
      n_eff <- obs_year %>%
        filter(year == y) %>%
        summarise(n = sum(value, na.rm = TRUE), .groups = "drop") %>%
        pull(n)
      n_eff <- max(50L, as.integer(round(n_eff)))
      tibble(
        year = y,
        !!bin_col := pdat[[bin_col]],
        proportion = as.numeric(rmultinom(1L, size = n_eff, prob = p)) / n_eff,
        sim_id = s
      )
    })
  })
}

make_year_comp <- function(kind, fishery, nsim = 12L) {
  if (kind == "lf") {
    pred_list <- bet$report$lf_pred
    fishery_f <- bet$data$lf_fishery_f
    n_eff_rows <- split(bet$data$lf_n, bet$data$lf_fishery)[[as.character(fishery)]]
    bins <- bet$data$len_mid
    obs <- bet$data$wcpo_bet_lf %>% filter(fishery == !!fishery)
    bin_col <- "length"
  } else {
    pred_list <- bet$report$wf_pred
    fishery_f <- bet$data$wf_fishery_f
    n_eff_rows <- split(bet$data$wf_n, bet$data$wf_fishery)[[as.character(fishery)]]
    bins <- bet$data$wt_mid
    obs <- bet$data$wcpo_bet_wf %>% filter(fishery == !!fishery)
    bin_col <- "weight"
  }

  i <- which(fishery_f == fishery)
  pred <- as.matrix(pred_list[[i]])
  obs_rows <- obs %>%
    distinct(year, month, ts) %>%
    arrange(ts, year, month)
  if (nrow(obs_rows) != nrow(pred)) {
    stop("Observed rows do not align with predicted composition rows")
  }
  if (length(n_eff_rows) != nrow(pred)) {
    stop("Effective N rows do not align with predicted composition rows")
  }
  obs_rows <- obs_rows %>%
    mutate(effective_n = as.numeric(n_eff_rows))

  pred_year <- as.data.frame(pred) %>%
    setNames(as.character(bins)) %>%
    mutate(year = obs_rows$year) %>%
    pivot_longer(-year, names_to = bin_col, values_to = "proportion") %>%
    mutate(!!bin_col := as.numeric(.data[[bin_col]])) %>%
    group_by(year, .data[[bin_col]]) %>%
    summarise(proportion = mean(proportion, na.rm = TRUE), .groups = "drop")

  obs_year <- obs %>%
    group_by(year, bin) %>%
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
    group_by(year) %>%
    mutate(proportion = value / sum(value, na.rm = TRUE)) %>%
    ungroup() %>%
    rename(!!bin_col := bin)

  effective_n <- obs_rows %>%
    group_by(year) %>%
    summarise(effective_n = sum(effective_n, na.rm = TRUE), .groups = "drop") %>%
    mutate(label = sprintf("Neff = %.1f", effective_n))

  list(
    observed = obs_year,
    predicted = pred_year,
    simulated = simulate_year_composition(pred_year, obs_year, bin_col, nsim = nsim),
    effective_n = effective_n,
    bin_col = bin_col
  )
}
```

```{r}
#| echo: false
#| label: fig-bet-lf12-year-composition
#| fig-cap: "Year-specific observed, predicted, and simulated length-composition proportions for selected years of WCPO bigeye tuna fishery 12. Unlike Figure 3, this plot does not aggregate over all years; each panel has its own fitted and simulated composition. Panel text gives the variance-adjusted effective N used by the likelihood for that year."
#| fig-height: 7.2
#| fig-width: 8

set.seed(12012)
bet_lf12_year <- make_year_comp("lf", 12, nsim = 12L)
bet_lf12_panel_years <- select_panel_years(bet_lf12_year$predicted$year, 16L)

ggplot() +
  geom_col(data = filter(bet_lf12_year$observed, year %in% bet_lf12_panel_years),
           aes(x = length, y = proportion), fill = "grey75", width = 1.8) +
  geom_line(data = filter(bet_lf12_year$predicted, year %in% bet_lf12_panel_years),
            aes(x = length, y = proportion), colour = "red3", linewidth = 0.55) +
  geom_line(data = filter(bet_lf12_year$simulated, year %in% bet_lf12_panel_years),
            aes(x = length, y = proportion, group = sim_id),
            colour = "steelblue", alpha = 0.55, linewidth = 0.3) +
  geom_line(data = filter(bet_lf12_year$predicted, year %in% bet_lf12_panel_years),
            aes(x = length, y = proportion), colour = "red3", linewidth = 0.4) +
  geom_text(data = filter(bet_lf12_year$effective_n, year %in% bet_lf12_panel_years),
            aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.05, vjust = 1.3,
            size = 2.1, colour = "grey25") +
  facet_wrap(~ year, ncol = 4, scales = "free_y") +
  labs(x = "Length (cm)", y = "Proportion") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  theme(strip.text = element_text(size = 7),
        axis.text = element_text(size = 6))
```

```{r}
#| echo: false
#| label: fig-bet-wf15-year-composition
#| fig-cap: "Year-specific observed, predicted, and simulated weight-composition proportions for selected years of WCPO bigeye tuna fishery 15. Unlike Figure 4, this plot does not aggregate over all years; each panel has its own fitted and simulated composition. Panel text gives the variance-adjusted effective N used by the likelihood for that year."
#| fig-height: 7.2
#| fig-width: 8

set.seed(15015)
bet_wf15_year <- make_year_comp("wf", 15, nsim = 12L)
bet_wf15_panel_years <- select_panel_years(bet_wf15_year$predicted$year, 20L)

ggplot() +
  geom_col(data = filter(bet_wf15_year$observed, year %in% bet_wf15_panel_years),
           aes(x = weight, y = proportion), fill = "grey75", width = 1.8) +
  geom_line(data = filter(bet_wf15_year$predicted, year %in% bet_wf15_panel_years),
            aes(x = weight, y = proportion), colour = "red3", linewidth = 0.5) +
  geom_line(data = filter(bet_wf15_year$simulated, year %in% bet_wf15_panel_years),
            aes(x = weight, y = proportion, group = sim_id),
            colour = "steelblue", alpha = 0.52, linewidth = 0.28) +
  geom_line(data = filter(bet_wf15_year$predicted, year %in% bet_wf15_panel_years),
            aes(x = weight, y = proportion), colour = "red3", linewidth = 0.35) +
  geom_text(data = filter(bet_wf15_year$effective_n, year %in% bet_wf15_panel_years),
            aes(x = Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = 1.05, vjust = 1.3,
            size = 2.1, colour = "grey25") +
  facet_wrap(~ year, ncol = 4, scales = "free_y") +
  labs(x = "Weight (kg)", y = "Proportion") +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal() +
  theme(strip.text = element_text(size = 7),
        axis.text = element_text(size = 6))
```

```{r}
#| echo: false
#| include: false
#| label: prep-bet-composition-residuals

bet_lf_residuals <- bet$diagnostics$lf_residuals %>%
  filter(is.finite(residual)) %>%
  transmute(
    type,
    Fishery,
    year,
    bin,
    residual,
    sign
  )

bet_wf_residuals <- bet$diagnostics$wf_residuals %>%
  filter(is.finite(residual)) %>%
  transmute(
    type,
    Fishery,
    year,
    bin,
    residual,
    sign
  )

bet_comp_residuals <- bind_rows(bet_lf_residuals, bet_wf_residuals) %>%
  mutate(
    type = factor(type, levels = c("Length composition", "Weight composition")),
    Fishery = factor(Fishery, levels = paste("Fishery", sort(unique(as.integer(sub("Fishery ", "", Fishery))))))
  )

plot_bet_comp_residuals <- function(plot_data, y_label) {
  plot_summary <- plot_data %>%
    group_by(Fishery, year, bin) %>%
    summarise(residual = mean(residual, na.rm = TRUE), .groups = "drop") %>%
    mutate(residual = pmax(pmin(residual, 4), -4))

  ggplot(plot_summary, aes(x = year, y = bin, fill = residual)) +
    geom_raster(interpolate = FALSE) +
    facet_wrap(~ Fishery, ncol = 2, scales = "free_y") +
    scale_fill_gradient2(
      low = "#4575b4", mid = "white", high = "#d73027",
      midpoint = 0, limits = c(-4, 4), oob = scales::squish,
      name = "Mean OSA residual"
    ) +
    labs(x = "Year", y = y_label) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

```{r}
#| echo: false
#| label: fig-bet-lf-composition-residuals
#| fig-cap: "Length-composition OSA residual diagnostics for fishery 12 in the fitted WCPO bigeye tuna diagnostic model. Colours show the mean OSA residual in each year-length cell, with blue indicating negative residuals and red indicating positive residuals."
#| fig-height: 4.8
#| fig-width: 8

plot_bet_comp_residuals(
  filter(bet_comp_residuals, type == "Length composition", Fishery == "Fishery 12"),
  y_label = "Length (cm)"
)
```

```{r}
#| echo: false
#| label: fig-bet-wf-composition-residuals
#| fig-cap: "Weight-composition OSA residual diagnostics for fishery 15 in the fitted WCPO bigeye tuna diagnostic model. Colours show the mean OSA residual in each year-weight cell, with blue indicating negative residuals and red indicating positive residuals."
#| fig-height: 4.8
#| fig-width: 8

plot_bet_comp_residuals(
  filter(bet_comp_residuals, type == "Weight composition", Fishery == "Fishery 15"),
  y_label = "Weight (kg)"
)
```

```{r}
#| echo: false
#| label: fig-bet-composition-residual-qq
#| fig-cap: "Normal Q-Q diagnostics for length- and weight-composition OSA residuals in the fitted WCPO bigeye tuna diagnostic model, shown separately by composition type and fishery. The red line is the fixed standard-normal one-to-one reference; it is not fitted separately within panels."
#| fig-height: 8
#| fig-width: 8

bet_comp_qq <- bet_comp_residuals %>%
  mutate(
    fishery_number = as.integer(sub("Fishery ", "", as.character(Fishery))),
    sdnr = ave(residual, type, Fishery, FUN = sd),
    panel = sprintf("%s\n%s\nSDNR = %.2f", type, Fishery, sdnr)
  ) %>%
  group_by(type, Fishery) %>%
  arrange(residual, .by_group = TRUE) %>%
  mutate(theoretical = qnorm(ppoints(n()))) %>%
  ungroup() %>%
  arrange(type, fishery_number) %>%
  mutate(panel = factor(panel, levels = unique(panel)))

bet_comp_qq_limit <- max(abs(c(bet_comp_qq$theoretical, bet_comp_qq$residual)), na.rm = TRUE)
bet_comp_qq_limit <- ceiling(min(max(bet_comp_qq_limit, 4), 5.5) * 2) / 2

ggplot(bet_comp_qq, aes(x = theoretical, y = residual)) +
  geom_abline(intercept = 0, slope = 1, colour = "red3", linetype = "dashed", linewidth = 0.35) +
  geom_point(size = 0.28, alpha = 0.25) +
  facet_wrap(~ panel, ncol = 4) +
  coord_equal(xlim = c(-bet_comp_qq_limit, bet_comp_qq_limit),
              ylim = c(-bet_comp_qq_limit, bet_comp_qq_limit)) +
  labs(x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal() +
  theme(strip.text = element_text(size = 6.5))
```

```{r}
#| echo: false
#| label: fig-bet-recruitment-deviations
#| fig-cap: "Estimated recruitment deviations for the fitted WCPO bigeye tuna diagnostic model, shown by seasonal index month."
#| fig-height: 5
#| fig-width: 8

bet_rec_raw <- bet$diagnostics$recruitment_raw
bet_rec_lookup <- bet_rec_raw$cpue_lookup %>%
  arrange(ts)
bet_rec_months <- sort(unique(bet_rec_lookup$month))
bet_rec_season_lookup <- tibble(
  month = bet_rec_months,
  season = seq_along(bet_rec_months),
  season_label = paste0(month.name[month])
)
bet_rec_df <- bet_rec_lookup %>%
  left_join(bet_rec_season_lookup, by = "month") %>%
  mutate(
    rdev = bet_rec_raw$rdev_y,
    estimated = if (length(bet_rec_raw$map_rdev) == length(bet_rec_raw$rdev_y)) {
      !is.na(bet_rec_raw$map_rdev)
    } else {
      TRUE
    },
    season_label = factor(season_label, levels = month.name[bet_rec_months])
  )

ggplot(bet_rec_df, aes(x = year, y = rdev)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey40") +
  geom_line(colour = "grey35", linewidth = 0.45) +
  geom_point(aes(colour = estimated), size = 1.1) +
  facet_wrap(~ season_label, ncol = 2) +
  scale_colour_manual(
    values = c(`FALSE` = "grey40", `TRUE` = "firebrick"),
    labels = c(`FALSE` = "Fixed", `TRUE` = "Estimated"),
    guide = "none"
  ) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  labs(x = "Year", y = "Recruitment deviation") +
  theme_minimal()
```

```{r}
#| echo: false
#| label: fig-bet-selectivity-length
#| fig-cap: "Selectivity-at-length curves for WCPO bigeye tuna fisheries. Blue lines show the fitted opal selectivity curves; dashed black lines show the SS3 reference selectivity curves used to initialise the diagnostic model."
#| fig-height: 8
#| fig-width: 8

sel_logistic_plain <- function(len, par) {
  mu <- mean(len)
  sd_len <- sd(len)
  a50 <- mu + par[1] * sd_len
  w95 <- exp(par[2]) * sd_len
  1 / (1 + exp(-log(19) * (len - a50) / w95))
}

sel_double_normal_plain <- function(len, par) {
  mu <- mean(len)
  sd_len <- sd(len)
  peak <- mu + par[1] * sd_len
  upselex <- exp(par[3]) * sd_len^2
  downselex <- exp(par[4]) * sd_len^2
  point1 <- 1 / (1 + exp(-par[5]))
  point2 <- 1 / (1 + exp(-par[6]))
  j2 <- length(len)
  bin_width <- len[2] - len[1]
  peak2 <- peak + bin_width + (0.99 * len[j2] - peak - bin_width) /
    (1 + exp(-par[2]))
  t1min <- exp(-(len[1] - peak)^2 / upselex)
  t2min <- exp(-(len[j2] - peak2)^2 / downselex)
  t1 <- len - peak
  t2 <- len - peak2
  join1 <- 1 / (1 + exp(-(20 / (1 + abs(t1))) * t1))
  join2 <- 1 / (1 + exp(-(20 / (1 + abs(t2))) * t2))
  asc <- point1 + (1 - point1) * (exp(-t1^2 / upselex) - t1min) / (1 - t1min)
  dsc <- 1 + (point2 - 1) * (exp(-t2^2 / downselex) - 1) / (t2min - 1)
  asc * (1 - join1) + join1 * (1 - join2 + dsc * join2)
}

bet_selectivity_curve <- function(par_sel, series) {
  map_dfr(seq_len(nrow(par_sel)), function(f) {
    sel <- if (bet_sel_raw$sel_type_f[f] == 1L) {
      sel_logistic_plain(bet_sel_raw$len_mid, par_sel[f, ])
    } else {
      sel_double_normal_plain(bet_sel_raw$len_mid, par_sel[f, ])
    }
    tibble(
      Fishery = paste("Fishery", f),
      length = bet_sel_raw$len_mid,
      selectivity = as.numeric(sel),
      series = series
    )
  })
}

bet_selectivity_df <- bind_rows(
  bet_selectivity_curve(bet_sel_raw$par_est, "opal estimate"),
  bet_selectivity_curve(bet_sel_raw$par_ref, "SS3 reference")
) %>%
  mutate(
    Fishery = factor(Fishery, levels = paste("Fishery", seq_len(nrow(bet_sel_raw$par_est)))),
    series = factor(series, levels = c("SS3 reference", "opal estimate"))
  )

ggplot(bet_selectivity_df, aes(x = length, y = selectivity)) +
  geom_line(data = filter(bet_selectivity_df, series == "SS3 reference"),
            colour = "black", linetype = "dashed", linewidth = 0.45) +
  geom_line(data = filter(bet_selectivity_df, series == "opal estimate"),
            colour = "cyan4", linewidth = 0.7) +
  facet_wrap(~ Fishery, ncol = 3) +
  labs(x = "Length (cm)", y = "Selectivity") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_minimal()
```

```{r}
#| echo: false
#| label: fig-bet-biomass
#| fig-cap: "Estimated spawning biomass and relative biomass trajectories for the fitted WCPO bigeye tuna diagnostic model. Shaded ribbons show approximate 95% intervals; report-level standard errors are used where finite, with a 10% CV display ribbon used where those standard errors are unavailable."
#| fig-height: 5
#| fig-width: 8

bet_biomass_unc <- bet$diagnostics$biomass_uncertainty
if (is.null(bet_biomass_unc)) {
  stop("BET biomass uncertainty is not available in the cached results")
}
bet_display_se <- function(value, se) {
  ifelse(is.finite(se), se, abs(value) * 0.10)
}
bet_biomass <- bind_rows(
  bet_biomass_unc %>%
    mutate(se = bet_display_se(spawning_biomass, spawning_biomass_se)) %>%
    transmute(
      year,
      quantity = "Spawning biomass (mt)",
      value = spawning_biomass,
      lower = pmax(0, spawning_biomass - 1.96 * se),
      upper = spawning_biomass + 1.96 * se
    ),
  bet_biomass_unc %>%
    mutate(se = bet_display_se(sb_sb0, sb_sb0_se)) %>%
    transmute(
      year,
      quantity = "SB/SB0",
      value = sb_sb0,
      lower = pmax(0, sb_sb0 - 1.96 * se),
      upper = sb_sb0 + 1.96 * se
    ),
  bet_biomass_unc %>%
    mutate(se = bet_display_se(dynamic_depletion, dynamic_depletion_se)) %>%
    transmute(
      year,
      quantity = "SB/SB_F0",
      value = dynamic_depletion,
      lower = pmax(0, dynamic_depletion - 1.96 * se),
      upper = dynamic_depletion + 1.96 * se
    )
) %>%
  mutate(
    quantity = factor(
      quantity,
      levels = c("Spawning biomass (mt)", "SB/SB0", "SB/SB_F0")
    )
  )

ggplot(bet_biomass, aes(x = year, y = value)) +
  geom_ribbon(aes(ymin = lower, ymax = upper),
              fill = "cyan4", alpha = 0.18, colour = NA) +
  geom_line(colour = "cyan4", linewidth = 0.9) +
  facet_wrap(~ quantity, ncol = 1, scales = "free_y") +
  labs(x = "Year", y = NULL) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  theme_minimal()
```

```{r}
#| echo: false
#| label: fig-paka-lf-biomass
#| fig-cap: "Estimated spawning-biomass trajectory for the 'opakapaka opal (with LF) model compared with the SS3 operating model (OM) and estimation model (EM), 1949--2023."
#| fig-height: 4.5
#| fig-width: 8

ybmax_lf <- max(paka$ssb$estimate, paka$om$SSB, paka$em$SSB, na.rm = TRUE)

ggplot() +
  geom_line(data = paka$ssb,
            aes(x = year, y = estimate, colour = "opal (with LF)"),
            linewidth = 1) +
  geom_line(data = paka$om,
            aes(x = Yr, y = SSB, colour = "SS3 OM"), linewidth = 1) +
  geom_line(data = paka$em,
            aes(x = Yr, y = SSB, colour = "SS3 EM"), linewidth = 1) +
  scale_colour_manual(values = c(
    "opal (with LF)" = "cyan4",
    "SS3 OM"         = "black",
    "SS3 EM"         = "red"
  )) +
  labs(x = "Year", y = "Spawning biomass", colour = NULL) +
  scale_y_continuous(limits = c(0, ybmax_lf * 1.2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| echo: false
#| label: fig-paka-mcmc-trace
#| fig-cap: "MCMC traces for the scalar parameters of the 'opakapaka opal (with LF) model --- unfished spawning biomass and the two CPUE catchabilities, on the log scale --- across eight chains (colours). Red dashed lines mark the MLE. Overlapping, stationary chains indicate good mixing."
#| fig-height: 5
#| fig-width: 8

ggplot(paka$mcmc_trace, aes(iter, value, colour = chain)) +
  geom_line(linewidth = 0.3, alpha = 0.7) +
  geom_hline(data = paka$mcmc_mle_ref, aes(yintercept = mle),
             linetype = "dashed", colour = "red", linewidth = 0.5) +
  facet_wrap(~ param, scales = "free_y", ncol = 1) +
  labs(x = "Post-warmup iteration", y = "Value", colour = "Chain") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| echo: false
#| label: fig-paka-projection
#| fig-cap: "'Opakapaka spawning biomass: historical uncertainty and a 10-year constant-catch projection (200 draws per method); the last 60 years are shown. Grey solid = estimated MLE; black dashed = SS3 operating model; red = SS3 estimation model. Coloured ribbons give the 50% (darker) and 90% (lighter) intervals for the MVN and MCMC approaches, and the orange line is the MCMC historical median. The vertical dashed line marks the end of the historical period (2024)."
#| fig-height: 4.5
#| fig-width: 8

pal_proj <- c(MVN = "steelblue", MCMC = "darkorange")

ggplot() +
  geom_ribbon(data = paka$proj_hist_ribbon,
              aes(year, ymin = lo90, ymax = hi90, fill = method), alpha = 0.12) +
  geom_ribbon(data = paka$proj_hist_ribbon,
              aes(year, ymin = lo50, ymax = hi50, fill = method), alpha = 0.25) +
  geom_line(data = subset(paka$proj_hist_ribbon, method == "MCMC"),
            aes(year, med, colour = method), linewidth = 0.75) +
  geom_line(data = paka$proj_hist_mle, aes(year, sbio),
            colour = "grey20", linewidth = 0.9) +
  geom_line(data = paka$proj_om, aes(Yr, SSB),
            colour = "black", linewidth = 0.8, linetype = "dashed") +
  geom_line(data = paka$proj_em, aes(Yr, SSB),
            colour = "red", linewidth = 0.8) +
  geom_vline(xintercept = 2024, linetype = "dashed", colour = "grey50") +
  geom_ribbon(data = paka$proj_ribbon,
              aes(year, ymin = lo90, ymax = hi90, fill = method), alpha = 0.15) +
  geom_ribbon(data = paka$proj_ribbon,
              aes(year, ymin = lo50, ymax = hi50, fill = method), alpha = 0.30) +
  geom_line(data = paka$proj_ribbon, aes(year, med, colour = method),
            linewidth = 1) +
  scale_fill_manual(values = pal_proj) +
  scale_colour_manual(values = pal_proj) +
  labs(x = "Year", y = "Spawning biomass", colour = "Method", fill = "Method") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

