---
title: "Retrospective analysis and the significance of Mohn's rho"
format:
	html:
		embed-resources: false
		html-math-method: mathml
		code-fold: true
		code-summary: "Show code"
execute:
	eval: true
	warning: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

A retrospective analysis refits an assessment model after sequentially removing
terminal data. Mohn's $\rho$ summarises the average relative change between the
terminal estimate in each peel and the estimate from the full fit. This
self-contained vignette reconstructs the retained WCPO bigeye tuna (BET) fit,
then builds its retrospective peel series.

The common rule of thumb $(-0.15, 0.20)$ is shown only as context. A Mohn's
$\rho$ statistic has a model- and data-dependent sampling distribution and is
not itself a measure of estimation bias.

# Setup

```{r}
#| label: load-pkg
#| echo: true
#| message: false
library(opal)
library(RTMB)
library(tidyverse)

theme_set(theme_bw())

find_repo_root <- function() {
	candidates <- normalizePath(c(".", ".."), mustWork = FALSE)
	hits <- candidates[file.exists(file.path(candidates, "assets", "cached", "bet",
																					 "opal_bet_fit.rds"))]
	if (!length(hits)) {
		stop("Cannot locate assets/cached/bet/opal_bet_fit.rds. Render from the ",
				 "opal-documentation repository or its assets directory.")
	}
	hits[[1]]
}

repo_dir <- find_repo_root()
cache_dir <- file.path(repo_dir, "assets", "cached", "bet", "retro")
cache_file <- file.path(cache_dir, "bet_retro.rds")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
retro_cache <- if (file.exists(cache_file)) readRDS(cache_file) else list()
cache_hit <- function(key) !is.null(retro_cache[[key]])
```

::: {.callout-note}
## Caching

The cache stores only plain-R outputs. RTMB AD objects are rebuilt in every R
session and are never serialized.
:::

## Reference configuration

```{r}
#| label: config
#| echo: true
bet_config <- list(
	lf_fisheries = c(8L, 9L, 10L, 11L, 12L, 13L, 14L),
	wf_fisheries = c(1L, 2L, 3L, 4L, 6L, 7L, 15L),
	lf_var_adjust = 400,
	lf_var_override = c(`8` = 25000, `10` = 1600),
	wf_var_adjust = 10000,
	wf_var_override = c(`3` = 40000, `6` = 40000, `7` = 40000),
	lf_switch = 1L,
	wf_switch = 1L,
	cpue_switch = 1L,
	cpue_tau = 1e-8,
	n_peel = 1L,
	quarters_per_year = 4L
)
```

## Reconstructing the retained fit

The promoted fit cache contains the exact full-period data used by the retained
fit, together with the optimum and diagnostics. It does not contain RTMB AD
objects, which are rebuilt below. The embedded source composition data are used
only to rebuild the derived LF/WF structures for shortened peels.

```{r}
#| label: reference-inputs
#| echo: true
#| message: false
base_fit_path <- file.path(repo_dir, "assets", "cached", "bet", "opal_bet_fit.rds")
base_fit <- readRDS(base_fit_path)
if (is.null(base_fit$opt$par) || is.null(base_fit$diagnostics$selectivity_raw)) {
	stop("Retained BET cache is missing opt$par or selectivity diagnostics.")
}

data(wcpo_bet_parameters, package = "opal")
base_data <- base_fit$data
if (is.null(base_data$wcpo_bet_lf) || is.null(base_data$wcpo_bet_wf)) {
	stop("Retained BET cache is missing embedded LF/WF source data.")
}

seed_active_parameters <- function(parameters, map, opt_par) {
	for (name in unique(names(opt_par))) {
		if (!name %in% names(parameters)) {
			stop("Retained active parameter is absent from bet_parameters: ", name)
		}
		values <- unname(opt_par[names(opt_par) == name])
		map_value <- map[[name]]
		if (is.null(map_value)) {
			if (length(parameters[[name]]) != length(values)) {
				stop("Retained active parameter has an incompatible length: ", name)
			}
			parameters[[name]] <- values
			next
		}
		if (length(parameters[[name]]) != length(map_value)) {
			stop("Parameter template and map have incompatible lengths: ", name)
		}

		active <- !is.na(map_value)
		level_id <- as.integer(map_value)
		active_levels <- sort(unique(level_id[active]))
		if (length(values) != length(active_levels)) {
			stop("Retained active parameter and map have incompatible lengths: ", name)
		}
		for (i in seq_along(active_levels)) {
			matches <- !is.na(level_id) & level_id == active_levels[[i]]
			parameters[[name]][matches] <- values[[i]]
		}
	}
	parameters
}

retained_opt_values <- function(name) {
	values <- unname(base_fit$opt$par[names(base_fit$opt$par) == name])
	if (!length(values)) stop("Retained optimum is missing active parameter: ", name)
	values
}

make_lf_wide <- function(lf) {
	lf %>%
		pivot_wider(id_cols = c(fishery, year, month, ts), names_from = bin,
								values_from = value, values_fill = 0) %>%
		arrange(fishery, ts)
}

make_wf_wide <- function(wf) {
	wf %>%
		pivot_wider(id_cols = c(fishery, year, month, ts), names_from = bin,
								values_from = value, values_fill = 0) %>%
		arrange(fishery, ts)
}

lf_wide <- make_lf_wide(base_data$wcpo_bet_lf)
wf_wide <- make_wf_wide(base_data$wcpo_bet_wf)

if (!identical(as.numeric(base_data$lf_switch), as.numeric(bet_config$lf_switch)) ||
		!identical(as.numeric(base_data$wf_switch), as.numeric(bet_config$wf_switch)) ||
		!identical(as.numeric(base_data$cpue_switch), as.numeric(bet_config$cpue_switch))) {
	stop("bet_config likelihood switches do not match the retained baseline data.")
}

bet_parameters <- list(
	log_B0 = retained_opt_values("log_B0"),
	log_h = as.numeric(wcpo_bet_parameters$log_h),
	log_sigma_r = as.numeric(wcpo_bet_parameters$log_sigma_r),
	log_cpue_q = retained_opt_values("log_cpue_q"),
	cpue_creep = rep(as.numeric(wcpo_bet_parameters$cpue_creep), base_data$n_index),
	log_cpue_tau = rep(log(bet_config$cpue_tau), base_data$n_index),
	log_cpue_omega = rep(as.numeric(wcpo_bet_parameters$log_cpue_omega),
								 base_data$n_index),
	log_lf_tau = log(rep(0.1, base_data$n_fishery)),
	log_wf_tau = rep(0, base_data$n_fishery),
	log_L1 = as.numeric(wcpo_bet_parameters$log_L1),
	log_L2 = as.numeric(wcpo_bet_parameters$log_L2),
	log_k = as.numeric(wcpo_bet_parameters$log_k),
	log_CV1 = as.numeric(wcpo_bet_parameters$log_CV1),
	log_CV2 = as.numeric(wcpo_bet_parameters$log_CV2),
	par_sel = base_fit$diagnostics$selectivity_raw$par_est,
	rdev_y = retained_opt_values("rdev_y")
)

map_sel <- base_fit$diagnostics$selectivity_raw$map_sel
bet_map <- list(
	log_h = factor(NA),
	log_sigma_r = factor(NA),
	log_cpue_q = factor(seq_len(base_data$n_index)),
	cpue_creep = factor(rep(NA, base_data$n_index)),
	log_cpue_tau = factor(rep(NA, base_data$n_index)),
	log_cpue_omega = factor(rep(NA, base_data$n_index)),
	log_lf_tau = factor(rep(NA_integer_, base_data$n_fishery)),
	log_wf_tau = factor(rep(NA_integer_, base_data$n_fishery)),
	log_L1 = factor(NA),
	log_L2 = factor(NA),
	log_k = factor(NA),
	log_CV1 = factor(NA),
	log_CV2 = factor(NA),
	par_sel = factor(map_sel),
	rdev_y = factor(base_fit$diagnostics$recruitment_raw$map_rdev)
)
bet_parameters <- seed_active_parameters(bet_parameters, bet_map, base_fit$opt$par)
if (length(bet_parameters$rdev_y) != base_data$n_year ||
		length(bet_map$rdev_y) != base_data$n_year) {
	stop("Cached recruitment deviations or their map do not match n_year.")
}
obj_reference <- MakeADFun(
	func = cmb(opal_model, base_data), parameters = bet_parameters, map = bet_map
)
obj_reference$env$tracemgc <- FALSE
bet_reference_par <- base_fit$opt$par
bounds <- base_fit$diagnostics$selectivity_raw$bounds
if (is.null(bounds$lower) || is.null(bounds$upper) ||
		length(bounds$lower) != length(bet_reference_par) ||
		length(bounds$upper) != length(bet_reference_par)) {
	stop("Retained BET cache is missing compatible optimizer bounds.")
}
names(bounds$lower) <- names(bet_reference_par)
names(bounds$upper) <- names(bet_reference_par)

if (!identical(names(obj_reference$par), names(bet_reference_par))) {
	stop("Reconstructed active parameters do not match the retained BET fit. ",
			 "Check bet_config and the selected cache.")
}
if (!isTRUE(all.equal(unname(obj_reference$par), unname(bet_reference_par),
									tolerance = 0))) {
	stop("Reconstructed bet_parameters are not seeded from the retained optimum.")
}
if (!is.finite(obj_reference$fn(bet_reference_par))) {
	stop("Retained BET parameters have a non-finite objective under this configuration.")
}

cat("Model time steps (quarters):", base_data$n_year, "\n")
cat("Peel depth:", bet_config$n_peel, "years =",
		bet_config$n_peel * bet_config$quarters_per_year, "quarters\n")
```

# Building peels

Composition structures are rebuilt from source rather than filtered in place.
The preflight identifies all numeric fields whose first dimension equals
`n_year`; only unhandled fields can emit a leak warning for a real peel.

```{r}
#| label: peel-helper
#| echo: true
#| message: false
truncate_leading <- function(x, keep) {
	d <- dim(x)
	if (is.null(d)) return(x[seq_len(keep)])
	idx <- rep(list(quote(expr = )), length(d))
	idx[[1]] <- seq_len(keep)
	do.call("[", c(list(x), idx, list(drop = FALSE)))
}

time_indexed_fields <- function(data) {
	n_year <- data$n_year
	names(Filter(function(x) {
		is.numeric(x) && length(x) > 0L &&
			identical(as.integer(if (is.null(dim(x))) length(x) else dim(x)[1]),
								as.integer(n_year))
	}, data))
}

known_time_fields <- c(
	"years", "catch_obs_ysf", "alk_ysal", "dl_yal", "length_mu_ysa",
	"af_sliced_ysfa", "bias_adj_y"
)
preflight_time_fields <- time_indexed_fields(base_data)
message("n_year-indexed fields in the reconstructed input: ",
				paste(preflight_time_fields, collapse = ", "))

peel_bet_inputs <- function(data, parameters, map, lf_wide, wf_wide,
														n_drop, config) {
	stopifnot(n_drop >= 0L, n_drop < data$n_year)
	keep <- data$n_year - n_drop
	d <- data
	handled <- intersect(known_time_fields, names(d))
	for (nm in handled) d[[nm]] <- truncate_leading(d[[nm]], keep)

	d$cpue_data <- dplyr::filter(data$cpue_data, ts <= keep)
	d$n_year <- keep
	if (!is.null(data$last_yr)) {
		d$last_yr <- data$last_yr - n_drop / config$quarters_per_year
	}

	lf_var <- rep(config$lf_var_adjust, d$n_fishery)
	lf_var[as.integer(names(config$lf_var_override))] <-
		as.numeric(config$lf_var_override)
	wf_var <- rep(config$wf_var_adjust, d$n_fishery)
	wf_var[as.integer(names(config$wf_var_override))] <-
		as.numeric(config$wf_var_override)
	d <- prep_lf_data(
		d, dplyr::filter(lf_wide, ts <= keep),
		lf_keep_fisheries = config$lf_fisheries,
		lf_var_adjust = lf_var, lf_switch = config$lf_switch
	)
	d$wt_bin_start <- 1L
	d$wt_bin_width <- 1L
	d$n_wt <- 200L
	d <- prep_wf_data(
		d, dplyr::filter(wf_wide, ts <= keep),
		wf_keep_fisheries = config$wf_fisheries,
		wf_var_adjust = wf_var, wf_switch = config$wf_switch
	)
	d$lf_switch <- config$lf_switch
	d$wf_switch <- config$wf_switch
	d$cpue_switch <- config$cpue_switch

	if (n_drop > 0L) {
		leaks <- setdiff(time_indexed_fields(d), handled)
		if (length(leaks)) {
			warning("Possible un-truncated year-indexed field(s): ",
							paste(leaks, collapse = ", "), call. = FALSE)
		}
	}

	p <- parameters
	p$rdev_y <- parameters$rdev_y[seq_len(keep)]
	m <- map
	if (!is.null(m$rdev_y)) m$rdev_y <- droplevels(factor(m$rdev_y[seq_len(keep)]))

	for (nm in handled) {
		dx <- dim(d[[nm]])
		if (!is.null(dx) && dx[1] != keep) stop("Failed to truncate ", nm, ".")
	}
	if (any(d$cpue_data$ts > keep) || length(p$rdev_y) != keep ||
			(!is.null(m$rdev_y) && length(m$rdev_y) != keep)) {
		stop("Peel validation failed for time-indexed data or recruitment deviations.")
	}
	list(data = d, parameters = p, map = m, keep = keep, n_drop = n_drop)
}
```

## Fitting and running peels

```{r}
#| label: fit-peels
#| echo: true
#| message: false
align_active_vector <- function(target, source, label) {
	target_keys <- paste(names(target), ave(seq_along(target), names(target), FUN = seq_along))
	source_keys <- paste(names(source), ave(seq_along(source), names(source), FUN = seq_along))
	pos <- match(target_keys, source_keys)
	if (anyNA(pos)) stop("Cannot align ", label, " to this peel's active parameters.")
	unname(source[pos])
}

summarise_estimability <- function(obj, par) {
	hessian <- tryCatch(
		obj$he(par),
		error = function(e) e
	)
	if (inherits(hessian, "error")) {
		return(list(
			status = "hessian_error",
			message = conditionMessage(hessian),
			min_eigen = NA_real_,
			n_bad_directions = NA_integer_,
			bad_params = tibble(Param = character(), MLE = numeric(),
										Param_check = character())
		))
	}

	check <- tryCatch(
		opal::check_estimability(obj, h = hessian),
		error = function(e) e
	)
	if (inherits(check, "error")) {
		return(list(
			status = "check_error",
			message = conditionMessage(check),
			min_eigen = NA_real_,
			n_bad_directions = NA_integer_,
			bad_params = tibble(Param = character(), MLE = numeric(),
										Param_check = character())
		))
	}

	bad_params <- as_tibble(check$BadParams) %>%
		select(any_of(c("Param", "MLE", "Param_check")))
	if (!all(c("Param", "MLE", "Param_check") %in% names(bad_params))) {
		bad_params <- tibble(Param = character(), MLE = numeric(),
								 Param_check = character())
	}
	list(
		status = "ok",
		message = NA_character_,
		min_eigen = min(Re(check$Eigen$values)),
		n_bad_directions = length(check$WhichBad),
		bad_params = bad_params
	)
}

fit_peel <- function(inputs, start_par, bounds, n_restart = 4L, grad_tol = 1e-2) {
  obj <- MakeADFun(func = cmb(opal_model, inputs$data),
                   parameters = inputs$parameters, map = inputs$map)
  obj$env$tracemgc <- FALSE
  par0 <- align_active_vector(obj$par, start_par, "starting values")
  Lwr <- align_active_vector(obj$par, bounds$lower, "lower bounds")
  Upr <- align_active_vector(obj$par, bounds$upper, "upper bounds")
  if (length(par0) != length(Lwr) || length(par0) != length(Upr)) {
    stop("Peel parameter and bound layouts differ.")
  }
  opt <- list(par = par0)
  for (i in seq_len(n_restart)) {
    opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
                  lower = Lwr, upper = Upr,
                  control = list(eval.max = 10000, iter.max = 10000))
    if (max(abs(obj$gr(opt$par))) < grad_tol) break
  }
  invisible(obj$fn(opt$par))
  obj$env$last.par.best <- opt$par
  grad <- as.numeric(obj$gr(opt$par))
  kkt <- abs(grad)
  at_l <- is.finite(Lwr) & opt$par <= Lwr + 1e-7
  at_u <- is.finite(Upr) & opt$par >= Upr - 1e-7
  kkt[at_l] <- pmax(0, -grad[at_l])
  kkt[at_u] <- pmax(0, grad[at_u])

  at_bound <- at_l | at_u

  bound_report <- data.frame(
    parameter   = names(obj$par)[at_bound],
    estimate    = opt$par[at_bound],
    bound_hit   = ifelse(at_l[at_bound], "Lower", "Upper"),
    bound_value = ifelse(at_l[at_bound], Lwr[at_bound], Upr[at_bound]),
    stringsAsFactors = FALSE
  )

  rep <- obj$report(opt$par)
  par_mle <- obj$env$parList(opt$par)
  sdr <- tryCatch(RTMB::sdreport(obj), error = function(e) NULL)
  estimability <- summarise_estimability(obj, opt$par)
  cpue_pred <- as.numeric(rep$cpue_pred)
  if (length(cpue_pred) != nrow(inputs$data$cpue_data)) {
    stop("CPUE predictions do not align with the peel CPUE observations.")
  }
  list(
    n_drop = inputs$n_drop, keep = inputs$keep,
    objective = as.numeric(opt$objective), convergence = opt$convergence,
    max_grad = max(abs(grad)), kkt_grad = max(kkt), n_at_bound = sum(at_l | at_u),
    par = opt$par,
    Lwr = Lwr,
    Upr = Upr,
    spawning_biomass_y = as.numeric(rep$spawning_biomass_y),
    spawning_biomass0_y = as.numeric(rep$spawning_biomass0_y),
    static_depletion_y = as.numeric(rep$static_depletion_y),
    dynamic_depletion_y = as.numeric(rep$dynamic_depletion_y),
    cpue_data = inputs$data$cpue_data,
    cpue_pred = cpue_pred,
    par_sel = par_mle$par_sel,
    sd_summary = if (is.null(sdr)) NULL else summary(sdr, "report"),
    pdHess = if (is.null(sdr)) NA else sdr$pdHess,
    estimability = estimability,
    bound_report = bound_report
  )
}

turn_off_bounded_parameters <- function(parameters, map, opt_par, Lwr, Upr, tol = 1e-7) {
  at_l <- is.finite(Lwr) & opt_par <= Lwr + tol
  at_u <- is.finite(Upr) & opt_par >= Upr - tol
  at_bound_idx <- which(at_l | at_u)
  
  if (length(at_bound_idx) == 0) {
    return(list(parameters = parameters, map = map, at_bound_idx = integer(0), report = data.frame()))
  }
  
  new_parameters <- parameters
  new_map <- map
  report_rows <- list()
  par_names <- names(opt_par)
  
  for (k in at_bound_idx) {
    pname <- par_names[k]
    val <- opt_par[k]
    bound_type <- if (at_l[k]) "Lower" else "Upper"
    bound_val  <- if (at_l[k]) Lwr[k] else Upr[k]
    
    instance_idx <- sum(par_names[1:k] == pname)
    
    if (is.null(new_map[[pname]])) {
      new_map[[pname]] <- factor(seq_along(new_parameters[[pname]]))
      matches <- instance_idx
    } else {
      m_fac <- map[[pname]]
      level_id <- as.integer(m_fac)
      active_levels <- sort(unique(level_id[!is.na(level_id)]))
      target_level <- active_levels[instance_idx]
      matches <- which(!is.na(level_id) & level_id == target_level)
    }
    
    new_parameters[[pname]][matches] <- val
    
    m_vec <- as.character(new_map[[pname]])
    m_vec[matches] <- NA_character_
    new_map[[pname]] <- factor(m_vec)
    
    report_rows[[length(report_rows) + 1]] <- data.frame(
      parameter = pname,
      element_indices = paste(matches, collapse = ","),
      estimate = val,
      bound_hit = bound_type,
      bound_value = bound_val,
      stringsAsFactors = FALSE
    )
  }
  
  for (nm in names(new_map)) {
    if (!is.null(new_map[[nm]]) && is.factor(new_map[[nm]])) {
      new_map[[nm]] <- droplevels(new_map[[nm]])
    }
  }
  
  list(
    parameters = new_parameters,
    map = new_map,
    at_bound_idx = at_bound_idx,
    report = do.call(rbind, report_rows)
  )
}

cache_provenance <- list(
  schema_version = 9L, # Increment schema version to force re-evaluation
  reference_cache = normalizePath(base_fit_path),
  config = bet_config,
  start_values = "base_fit$opt$par",
  estimability = "opal::check_estimability(obj, h = obj$he(par))"
)

cache_matches <- cache_hit("peels") && identical(retro_cache$provenance, cache_provenance)

if (!cache_matches) {
  if (cache_hit("peels")) message("Retrospective cache provenance changed; recomputing peels.")
  
  # --- 1. Evaluate Peel 0 (Reference Model) first ---
  message("Pre-fitting Peel 0 (Reference Model) to evaluate parameter bounds...")
  inp0 <- peel_bet_inputs(base_data, bet_parameters, bet_map, lf_wide, wf_wide,
                          n_drop = 0L, config = bet_config)
  fit0_initial <- fit_peel(inp0, start_par = bet_reference_par, bounds = bounds)
  
  # --- 2. Fix parameters hit on bounds in Peel 0 ---
  fix_info <- turn_off_bounded_parameters(
    parameters = bet_parameters,
    map        = bet_map,
    opt_par    = fit0_initial$par,
    Lwr        = fit0_initial$Lwr,
    Upr        = fit0_initial$Upr
  )
  
  bet_parameters <- fix_info$parameters
  bet_map        <- fix_info$map
  
  if (length(fix_info$at_bound_idx) > 0) {
    message("Turning off ", length(fix_info$at_bound_idx), " bounded parameter(s) from Peel 0.")
    keep_idx <- setdiff(seq_along(fit0_initial$par), fix_info$at_bound_idx)
    
    bet_reference_par <- fit0_initial$par[keep_idx]
    bounds$lower      <- fit0_initial$Lwr[keep_idx]
    bounds$upper      <- fit0_initial$Upr[keep_idx]
    names(bounds$lower) <- names(bet_reference_par)
    names(bounds$upper) <- names(bet_reference_par)
  }
  
  # --- 3. Fit all retrospective peels with updated active map ---
  peel_depths <- (0:bet_config$n_peel) * bet_config$quarters_per_year
  peels <- purrr::map(peel_depths, function(n_drop) {
    message("Peel with ", n_drop / bet_config$quarters_per_year, " calendar years removed")
    inp <- peel_bet_inputs(base_data, bet_parameters, bet_map, lf_wide, wf_wide,
                           n_drop = n_drop, config = bet_config)
    fit_peel(inp, start_par = bet_reference_par, bounds = bounds)
  })
  names(peels) <- paste0("peel_", 0:bet_config$n_peel)
  
  retro_cache$peels <- peels
  retro_cache$provenance <- cache_provenance
  saveRDS(retro_cache, cache_file)
}
peels <- retro_cache$peels
```

```{r}
#| label: tbl-peel-convergence
#| echo: true
#| tbl-cap: Convergence and estimability diagnostics by retrospective peel. Estimability is evaluated only for active parameters and should be interpreted with the KKT gradient and Hessian status.
purrr::map_dfr(peels, function(p) {
	est <- p$estimability
	bad_names <- if (identical(est$status, "ok")) {
		bad <- est$bad_params %>% filter(Param_check == "Bad") %>% pull(Param)
		if (length(bad)) paste(unique(bad), collapse = ", ") else "None"
	} else {
		NA_character_
	}
	tibble(peel = p$n_drop / bet_config$quarters_per_year,
				 time_steps = p$keep, objective = p$objective,
				 max_grad = p$max_grad, kkt_grad = p$kkt_grad,
				 n_at_bound = p$n_at_bound, pdHess = p$pdHess,
				 estimability = est$status,
				 min_eigen = est$min_eigen,
				 n_bad_directions = est$n_bad_directions,
				 bad_parameters = bad_names,
				 estimability_message = est$message)
}) %>%
	mutate(retained = kkt_grad < 1e-2 & (is.na(pdHess) | pdHess),
				 across(c(objective, max_grad, kkt_grad, min_eigen), ~ signif(.x, 4))) %>%
	knitr::kable()
```

`check_estimability()` examines the Hessian only for parameters active in the
peel's RTMB objective. A `Bad` result therefore identifies an active parameter
that contributes materially to a near-null Hessian direction; it is not a
convergence verdict. Read it alongside the KKT gradient, `pdHess`, and any
estimability diagnostic error shown in the table.

# Peel time series

The plot below overlays all peels within each metric facet. Pointwise Wald 95%
intervals come from `sdreport()` and use the same colour as their peel. Intervals
are clipped to the feasible scale: zero for spawning biomass and $[0, 1]$ for
depletion. A peel without a usable `sdreport()` result retains its estimate but
has no uncertainty ribbon.

```{r}
#| label: prepare-peel-time-series
#| echo: true
report_series <- function(peel_fit, peel_label, quantity, metric) {
	n <- length(peel_fit[[quantity]])
	se <- rep(NA_real_, n)
	if (!is.null(peel_fit$sd_summary) && "Std. Error" %in% colnames(peel_fit$sd_summary)) {
		rows <- which(rownames(peel_fit$sd_summary) == quantity)
		if (length(rows)) {
			se[seq_len(min(n, length(rows)))] <-
				peel_fit$sd_summary[rows[seq_len(min(n, length(rows)))], "Std. Error"]
		}
	}
	tibble(
		peel = peel_label,
		calendar_year = as.numeric(base_data$years[seq_len(n)]),
		metric = metric,
		estimate = as.numeric(peel_fit[[quantity]]),
		se = as.numeric(se)
	)
}

peel_labels <- paste0("Peel ", purrr::map_int(peels, "n_drop") / bet_config$quarters_per_year)
peel_time_series <- purrr::imap_dfr(peels, function(peel_fit, peel_name) {
	peel_label <- paste0("Peel ", peel_fit$n_drop / bet_config$quarters_per_year)
	bind_rows(
		report_series(peel_fit, peel_label, "spawning_biomass_y", "Spawning biomass"),
		report_series(peel_fit, peel_label, "static_depletion_y", "Static depletion"),
		report_series(peel_fit, peel_label, "dynamic_depletion_y", "Dynamic depletion")
	)
}) %>%
	mutate(
		peel = factor(peel, levels = peel_labels),
		lower = estimate - qnorm(0.975) * se,
		upper = estimate + qnorm(0.975) * se,
		lower = pmax(0, lower),
		upper = if_else(metric == "Spawning biomass", upper, pmin(1, upper))
	)
```

```{r}
#| label: fig-peel-time-series
#| fig-cap: Retrospective time series by metric. Colour identifies the number of calendar years removed; ribbons are pointwise 95% uncertainty intervals.
#| fig-width: 12
#| fig-height: 10
peel_time_series %>%
	ggplot(aes(calendar_year, estimate, group = peel)) +
	geom_ribbon(aes(ymin = lower, ymax = upper, fill = peel), alpha = 0.25, na.rm = TRUE) +
	geom_line(linewidth = 1.15, aes(colour = peel)) +
	facet_wrap(~ metric, scales = "free_y", ncol = 1) +
	scale_colour_viridis_d(option = "turbo", end = 0.9) +
	scale_fill_viridis_d(option = "turbo", end = 0.9) +
	labs(x = "Calendar year", y = NULL, colour = "Peel (years removed)", fill = "Peel (years removed)") +
	guides(fill = "none") +
	theme(legend.position = "bottom")
```

# Index fit by peel

The index-fit plot uses the log-scale RMSE, $
\sqrt{\operatorname{mean}[(\log I_\mathrm{obs} - \log I_\mathrm{pred})^2]}$,
calculated separately for each retained index observation in a peel.

```{r}
#| label: prepare-peel-index-fit
#| echo: true
peel_cpue_fit <- purrr::imap_dfr(peels, function(peel_fit, peel_name) {
	# CPUE predictions follow the baseline rows retained by this peel exactly.
	cpue <- base_data$cpue_data %>% filter(ts <= peel_fit$keep)
	if (!all(c("year", "month", "value") %in% names(cpue))) {
		stop("Baseline CPUE data are missing year, month, or value.")
	}
	if (is.null(peel_fit$cpue_pred)) {
		stop("Cached peel lacks CPUE predictions; rerun the fit-peels chunk to refresh the cache.")
	}
	if (nrow(cpue) != length(peel_fit$cpue_pred)) {
		stop("Baseline CPUE rows and cached peel predictions have incompatible lengths.")
	}
	cpue %>%
		transmute(
			peel = paste0("Peel ", peel_fit$n_drop / bet_config$quarters_per_year),
			calendar_year = as.numeric(year) + (as.numeric(month) - 1) / 12,
			index = factor(month, levels = sort(unique(base_data$cpue_data$month)),
				labels = month.abb[sort(unique(base_data$cpue_data$month))]),
			observed = as.numeric(value),
			predicted = as.numeric(peel_fit$cpue_pred)
		)
}) %>%
	mutate(peel = factor(peel, levels = peel_labels))

cpue_observed <- peel_cpue_fit %>%
	select(index, calendar_year, observed) %>%
	distinct()
cpue_rmse <- peel_cpue_fit %>%
	filter(observed > 0, predicted > 0) %>%
	group_by(index, peel) %>%
	summarise(rmse = sqrt(mean((log(observed) - log(predicted))^2)), .groups = "drop") %>%
	group_by(index) %>%
	arrange(peel, .by_group = TRUE) %>%
	mutate(label_rank = row_number(), label = paste0(as.character(peel), ": ", sprintf("%.3f", rmse))) %>%
	ungroup()
```

```{r}
#| label: fig-peel-index-fit
#| fig-cap: Observed and predicted CPUE by index and retrospective peel. Labels give the log-scale RMSE for each peel within an index.
#| fig-width: 12
#| fig-height: 7
ggplot(peel_cpue_fit, aes(calendar_year, predicted, colour = peel, group = peel)) +
	geom_point(data = cpue_observed, aes(y = observed), inherit.aes = FALSE,
				 colour = "grey15", size = 1.1, alpha = 0.7) +
	geom_line(linewidth = 0.7) +
	geom_text(data = cpue_rmse, aes(x = Inf, y = Inf, label = label, colour = peel,
											group = peel, vjust = 1.1 + 1.15 * label_rank),
				 hjust = 1.05, size = 2.8, show.legend = FALSE) +
	facet_wrap(~ index, scales = "free_y") +
	scale_y_log10() +
	scale_colour_viridis_d(option = "turbo", end = 0.9) +
	labs(x = "Calendar year", y = "Standardized CPUE", colour = "Peel (years removed)") +
	theme(legend.position = "bottom")
```

# Selectivity by peel

```{r}
#| label: prepare-peel-selectivity
#| echo: true
sel_logistic <- function(len, par) {
	mean_length <- mean(len)
	sd_length <- sd(len)
	a50 <- mean_length + par[1] * sd_length
	w95 <- exp(par[2]) * sd_length
	1 / (1 + exp(-log(19) * (len - a50) / w95))
}

sel_double_normal <- function(len, par) {
	mean_length <- mean(len)
	sd_length <- sd(len)
	peak <- mean_length + par[1] * sd_length
	upper_scale <- exp(par[3]) * sd_length^2
	lower_scale <- exp(par[4]) * sd_length^2
	upper_point <- plogis(par[5])
	lower_point <- plogis(par[6])
	last <- length(len)
	bin_width <- len[2] - len[1]
	second_peak <- peak + bin_width + (0.99 * len[last] - peak - bin_width) /
		(1 + exp(-par[2]))
	upper_min <- exp(-(len[1] - peak)^2 / upper_scale)
	lower_min <- exp(-(len[last] - second_peak)^2 / lower_scale)
	upper_distance <- len - peak
	lower_distance <- len - second_peak
	join_upper <- plogis((20 / (1 + abs(upper_distance))) * upper_distance)
	join_lower <- plogis((20 / (1 + abs(lower_distance))) * lower_distance)
	ascending <- upper_point + (1 - upper_point) *
		(exp(-upper_distance^2 / upper_scale) - upper_min) / (1 - upper_min)
	descending <- 1 + (lower_point - 1) *
		(exp(-lower_distance^2 / lower_scale) - 1) / (lower_min - 1)
	ascending * (1 - join_upper) + join_upper * (1 - join_lower + descending * join_lower)
}

sel_type_f <- base_fit$diagnostics$selectivity_raw$sel_type_f
if (is.null(sel_type_f)) sel_type_f <- base_data$sel_type_f
if (is.null(sel_type_f)) stop("Selectivity types are unavailable in the retained fit cache.")

peel_selectivity <- purrr::imap_dfr(peels, function(peel_fit, peel_name) {
	par_sel <- peel_fit$par_sel
	if (is.null(dim(par_sel))) stop("Cached peel selectivity parameters are not a matrix.")
	if (nrow(par_sel) != length(sel_type_f)) {
		stop("Cached peel selectivity parameters and selectivity types are incompatible.")
	}
	purrr::map_dfr(seq_len(nrow(par_sel)), function(fishery) {
		selectivity <- if (sel_type_f[[fishery]] == 1L) {
			sel_logistic(base_data$len_mid, par_sel[fishery, ])
		} else {
			sel_double_normal(base_data$len_mid, par_sel[fishery, ])
		}
		tibble(
			peel = paste0("Peel ", peel_fit$n_drop / bet_config$quarters_per_year),
			fishery = paste("Fishery", fishery),
			length = base_data$len_mid,
			selectivity = pmin(1, pmax(0, as.numeric(selectivity)))
		)
	})
}) %>%
	mutate(
		peel = factor(peel, levels = peel_labels),
		fishery = factor(fishery, levels = paste("Fishery", seq_along(sel_type_f)))
	)
```

```{r}
#| label: fig-peel-selectivity
#| fig-cap: Estimated selectivity-at-length curves for each retrospective peel. Colours identify the number of calendar years removed.
#| fig-width: 12
#| fig-height: 14
ggplot(peel_selectivity, aes(length, selectivity, colour = peel, group = peel)) +
	geom_line(linewidth = 0.65) +
	facet_wrap(~ fishery, ncol = 3) +
	coord_cartesian(ylim = c(0, 1)) +
	scale_colour_viridis_d(option = "turbo", end = 0.9) +
	labs(x = "Length", y = "Selectivity", colour = "Peel (years removed)") +
	theme(legend.position = "bottom")
```

# Mohn's rho

```{r}
#| label: mohn-rho
#| echo: true
qpy <- bet_config$quarters_per_year
mohn_rho <- function(peels, quantity, unit = c("quarter", "year"), drop_last = 0L) {
	unit <- match.arg(unit)
	full <- peels[[1]][[quantity]]
	use <- peels[-1]
	if (drop_last > 0L) use <- use[seq_len(length(use) - drop_last)]
	terms <- purrr::map_dfr(use, function(p) {
		idx <- if (unit == "quarter") p$keep else seq(p$keep - qpy + 1L, p$keep)
		x_peel <- mean(p[[quantity]][idx])
		x_full <- mean(full[idx])
		tibble(peel = p$n_drop / qpy, x_peel = x_peel, x_full = x_full,
					 rel_diff = (x_peel - x_full) / x_full)
	})
	list(rho = mean(terms$rel_diff), terms = terms)
}

quantities <- c("spawning_biomass_y", "dynamic_depletion_y", "static_depletion_y")
rho_table <- tidyr::expand_grid(quantity = quantities,
																unit = c("quarter", "year"),
																drop_last = c(0L, 1L)) %>%
	mutate(res = purrr::pmap(list(quantity, unit, drop_last),
													 ~ mohn_rho(peels, ..1, ..2, ..3)),
				 rho = purrr::map_dbl(res, "rho"),
				 min_diff = purrr::map_dbl(res, ~ min(.x$terms$rel_diff)),
				 max_diff = purrr::map_dbl(res, ~ max(.x$terms$rel_diff))) %>%
	select(-res)

rho_table %>%
	mutate(unit = recode(unit, quarter = "Terminal quarter", year = "Terminal-year mean"),
				 peels_used = if_else(drop_last == 0L, "all", "final peel dropped"),
				 across(c(rho, min_diff, max_diff), ~ round(.x, 3)),
				 outside_rule = rho < -0.15 | rho > 0.20) %>%
	select(quantity, unit, peels_used, rho, min_diff, max_diff, outside_rule) %>%
	knitr::kable()
```

# Interpretation

A retrospective pattern indicates that post-sample data update an estimate in
a way that is inconsistent with the fitted model conditioned on earlier data.
It does not identify the direction or magnitude of estimation error, and does
not justify adjusting $\rho$.

The post-sample bootstrap significance test is intentionally not executed here.
Each replicate requires all retrospective refits and should be run separately
after the deterministic peel series has been reviewed.
