---
title: "Retrospective analysis and the significance of Mohn's rho"
format:
	html:
		embed-resources: false
		html-math-method: mathml
		code-fold: true
		code-summary: "Show code"
execute:
	eval: true
	warning: false
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

A retrospective analysis refits an assessment model after sequentially removing
terminal data. Mohn's $\rho$ summarises the average relative change between the
terminal estimate in each peel and the estimate from the full fit. This
self-contained vignette reconstructs the retained WCPO bigeye tuna (BET) fit,
then builds its retrospective peel series.

The common rule of thumb $(-0.15, 0.20)$ is shown only as context. A Mohn's
$\rho$ statistic has a model- and data-dependent sampling distribution and is
not itself a measure of estimation bias.

# Setup

```{r}
#| label: load-pkg
#| echo: true
#| message: false
library(opal)
library(RTMB)
library(tidyverse)

theme_set(theme_bw())

find_repo_root <- function() {
	candidates <- normalizePath(c(".", ".."), mustWork = FALSE)
	hits <- candidates[file.exists(file.path(candidates, "assets", "cached", "bet",
																					 "opal_bet_fit.rds"))]
	if (!length(hits)) {
		stop("Cannot locate assets/cached/bet/opal_bet_fit.rds. Render from the ",
				 "opal-documentation repository or its assets directory.")
	}
	hits[[1]]
}

repo_dir <- find_repo_root()
cache_dir <- file.path(repo_dir, "assets", "cached", "bet", "retro")
cache_file <- file.path(cache_dir, "bet_retro.rds")
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
retro_cache <- if (file.exists(cache_file)) readRDS(cache_file) else list()
```

::: {.callout-note}
## Caching

The cache stores only plain-R outputs. RTMB AD objects are rebuilt in every R
session and are never serialized.
:::

## Reference configuration

```{r}
#| label: config
#| echo: true
bet_config <- list(
	lf_fisheries = c(8L, 9L, 10L, 11L, 12L, 13L, 14L),
	wf_fisheries = c(1L, 2L, 3L, 4L, 6L, 7L, 15L),
	lf_var_adjust = 400,
	lf_var_override = c(`8` = 25000, `10` = 1600),
	wf_var_adjust = 10000,
	wf_var_override = c(`3` = 40000, `6` = 40000, `7` = 40000),
	lf_switch = 1L,
	wf_switch = 1L,
	cpue_switch = 1L,
	cpue_tau = 1e-8,
	n_peel = 6L,
	quarters_per_year = 4L,
	fix_bounded_at = "estimate"
)
```

## Reconstructing the retained fit

The promoted fit cache contains the exact full-period data used by the retained
fit, together with the optimum and diagnostics. It does not contain RTMB AD
objects, which are rebuilt below. The embedded source composition data are used
only to rebuild the derived LF/WF structures for shortened peels.

```{r}
#| label: reference-inputs
#| echo: true
#| message: false
base_fit_path <- file.path(repo_dir, "assets", "cached", "bet", "opal_bet_fit.rds")
base_fit <- readRDS(base_fit_path)
if (is.null(base_fit$opt$par) || is.null(base_fit$diagnostics$selectivity_raw)) {
	stop("Retained BET cache is missing opt$par or selectivity diagnostics.")
}

data(wcpo_bet_parameters, package = "opal")
base_data <- base_fit$data
if (is.null(base_data$wcpo_bet_lf) || is.null(base_data$wcpo_bet_wf)) {
	stop("Retained BET cache is missing embedded LF/WF source data.")
}

seed_active_parameters <- function(parameters, map, opt_par) {
	for (name in unique(names(opt_par))) {
		if (!name %in% names(parameters)) {
			stop("Retained active parameter is absent from bet_parameters: ", name)
		}
		values <- unname(opt_par[names(opt_par) == name])
		map_value <- map[[name]]
		if (is.null(map_value)) {
			if (length(parameters[[name]]) != length(values)) {
				stop("Retained active parameter has an incompatible length: ", name)
			}
			parameters[[name]] <- values
			next
		}
		if (length(parameters[[name]]) != length(map_value)) {
			stop("Parameter template and map have incompatible lengths: ", name)
		}

		active <- !is.na(map_value)
		level_id <- as.integer(map_value)
		active_levels <- sort(unique(level_id[active]))
		if (length(values) != length(active_levels)) {
			stop("Retained active parameter and map have incompatible lengths: ", name)
		}
		for (i in seq_along(active_levels)) {
			matches <- !is.na(level_id) & level_id == active_levels[[i]]
			parameters[[name]][matches] <- values[[i]]
		}
	}
	parameters
}

retained_opt_values <- function(name) {
	values <- unname(base_fit$opt$par[names(base_fit$opt$par) == name])
	if (!length(values)) stop("Retained optimum is missing active parameter: ", name)
	values
}

make_lf_wide <- function(lf) {
	lf %>%
		pivot_wider(id_cols = c(fishery, year, month, ts), names_from = bin,
								values_from = value, values_fill = 0) %>%
		arrange(fishery, ts)
}

make_wf_wide <- function(wf) {
	wf %>%
		pivot_wider(id_cols = c(fishery, year, month, ts), names_from = bin,
								values_from = value, values_fill = 0) %>%
		arrange(fishery, ts)
}

lf_wide <- make_lf_wide(base_data$wcpo_bet_lf)
wf_wide <- make_wf_wide(base_data$wcpo_bet_wf)

if (!identical(as.numeric(base_data$lf_switch), as.numeric(bet_config$lf_switch)) ||
		!identical(as.numeric(base_data$wf_switch), as.numeric(bet_config$wf_switch)) ||
		!identical(as.numeric(base_data$cpue_switch), as.numeric(bet_config$cpue_switch))) {
	stop("bet_config likelihood switches do not match the retained baseline data.")
}

bet_parameters <- list(
	log_B0 = retained_opt_values("log_B0"),
	log_h = as.numeric(wcpo_bet_parameters$log_h),
	log_sigma_r = as.numeric(wcpo_bet_parameters$log_sigma_r),
	log_cpue_q = retained_opt_values("log_cpue_q"),
	cpue_creep = rep(as.numeric(wcpo_bet_parameters$cpue_creep), base_data$n_index),
	log_cpue_tau = rep(log(bet_config$cpue_tau), base_data$n_index),
	log_cpue_omega = rep(as.numeric(wcpo_bet_parameters$log_cpue_omega),
								 base_data$n_index),
	log_lf_tau = log(rep(0.1, base_data$n_fishery)),
	log_wf_tau = rep(0, base_data$n_fishery),
	log_L1 = as.numeric(wcpo_bet_parameters$log_L1),
	log_L2 = as.numeric(wcpo_bet_parameters$log_L2),
	log_k = as.numeric(wcpo_bet_parameters$log_k),
	log_CV1 = as.numeric(wcpo_bet_parameters$log_CV1),
	log_CV2 = as.numeric(wcpo_bet_parameters$log_CV2),
	par_sel = base_fit$diagnostics$selectivity_raw$par_est,
	rdev_y = retained_opt_values("rdev_y")
)

map_sel <- base_fit$diagnostics$selectivity_raw$map_sel
bet_map <- list(
	log_h = factor(NA),
	log_sigma_r = factor(NA),
	log_cpue_q = factor(seq_len(base_data$n_index)),
	cpue_creep = factor(rep(NA, base_data$n_index)),
	log_cpue_tau = factor(rep(NA, base_data$n_index)),
	log_cpue_omega = factor(rep(NA, base_data$n_index)),
	log_lf_tau = factor(rep(NA_integer_, base_data$n_fishery)),
	log_wf_tau = factor(rep(NA_integer_, base_data$n_fishery)),
	log_L1 = factor(NA),
	log_L2 = factor(NA),
	log_k = factor(NA),
	log_CV1 = factor(NA),
	log_CV2 = factor(NA),
	par_sel = factor(map_sel),
	rdev_y = factor(base_fit$diagnostics$recruitment_raw$map_rdev)
)
bet_parameters_template <- bet_parameters
bet_parameters <- seed_active_parameters(bet_parameters, bet_map, base_fit$opt$par)
if (length(bet_parameters$rdev_y) != base_data$n_year ||
		length(bet_map$rdev_y) != base_data$n_year) {
	stop("Cached recruitment deviations or their map do not match n_year.")
}
obj_reference <- MakeADFun(
	func = cmb(opal_model, base_data), parameters = bet_parameters, map = bet_map
)
obj_reference$env$tracemgc <- FALSE
bet_reference_par <- base_fit$opt$par
bounds <- base_fit$diagnostics$selectivity_raw$bounds
if (is.null(bounds$lower) || is.null(bounds$upper) ||
		length(bounds$lower) != length(bet_reference_par) ||
		length(bounds$upper) != length(bet_reference_par)) {
	stop("Retained BET cache is missing compatible optimizer bounds.")
}
names(bounds$lower) <- names(bet_reference_par)
names(bounds$upper) <- names(bet_reference_par)

if (!identical(names(obj_reference$par), names(bet_reference_par))) {
	stop("Reconstructed active parameters do not match the retained BET fit. ",
			 "Check bet_config and the selected cache.")
}
if (!isTRUE(all.equal(unname(obj_reference$par), unname(bet_reference_par),
									tolerance = 0))) {
	stop("Reconstructed bet_parameters are not seeded from the retained optimum.")
}
if (!is.finite(obj_reference$fn(bet_reference_par))) {
	stop("Retained BET parameters have a non-finite objective under this configuration.")
}

cat("Model time steps (quarters):", base_data$n_year, "\n")
cat("Peel depth:", bet_config$n_peel, "years =",
		bet_config$n_peel * bet_config$quarters_per_year, "quarters\n")
```

# Building peels

Composition structures are rebuilt from source rather than filtered in place.
The preflight identifies all numeric fields whose first dimension equals
`n_year`; only unhandled fields can emit a leak warning for a real peel.

```{r}
#| label: peel-helper
#| echo: true
#| message: false
truncate_leading <- function(x, keep) {
	d <- dim(x)
	if (is.null(d)) return(x[seq_len(keep)])
	idx <- rep(list(quote(expr = )), length(d))
	idx[[1]] <- seq_len(keep)
	do.call("[", c(list(x), idx, list(drop = FALSE)))
}

time_indexed_fields <- function(data) {
	n_year <- data$n_year
	names(Filter(function(x) {
		is.numeric(x) && length(x) > 0L &&
			identical(as.integer(if (is.null(dim(x))) length(x) else dim(x)[1]),
								as.integer(n_year))
	}, data))
}

known_time_fields <- c(
	"years", "catch_obs_ysf", "alk_ysal", "dl_yal", "length_mu_ysa",
	"af_sliced_ysfa", "bias_adj_y"
)
preflight_time_fields <- time_indexed_fields(base_data)
message("n_year-indexed fields in the reconstructed input: ",
				paste(preflight_time_fields, collapse = ", "))

peel_bet_inputs <- function(data, parameters, map, lf_wide, wf_wide,
														n_drop, config) {
	stopifnot(n_drop >= 0L, n_drop < data$n_year)
	keep <- data$n_year - n_drop
	d <- data
	handled <- intersect(known_time_fields, names(d))
	for (nm in handled) d[[nm]] <- truncate_leading(d[[nm]], keep)

	d$cpue_data <- dplyr::filter(data$cpue_data, ts <= keep)
	d$n_year <- keep
	if (!is.null(data$last_yr)) {
		d$last_yr <- data$last_yr - n_drop / config$quarters_per_year
	}

	lf_var <- rep(config$lf_var_adjust, d$n_fishery)
	lf_var[as.integer(names(config$lf_var_override))] <-
		as.numeric(config$lf_var_override)
	wf_var <- rep(config$wf_var_adjust, d$n_fishery)
	wf_var[as.integer(names(config$wf_var_override))] <-
		as.numeric(config$wf_var_override)
	d <- prep_lf_data(
		d, dplyr::filter(lf_wide, ts <= keep),
		lf_keep_fisheries = config$lf_fisheries,
		lf_var_adjust = lf_var, lf_switch = config$lf_switch
	)
	d$wt_bin_start <- 1L
	d$wt_bin_width <- 1L
	d$n_wt <- 200L
	d <- prep_wf_data(
		d, dplyr::filter(wf_wide, ts <= keep),
		wf_keep_fisheries = config$wf_fisheries,
		wf_var_adjust = wf_var, wf_switch = config$wf_switch
	)
	d$lf_switch <- config$lf_switch
	d$wf_switch <- config$wf_switch
	d$cpue_switch <- config$cpue_switch

	if (n_drop > 0L) {
		leaks <- setdiff(time_indexed_fields(d), handled)
		if (length(leaks)) {
			warning("Possible un-truncated year-indexed field(s): ",
							paste(leaks, collapse = ", "), call. = FALSE)
		}
	}

	p <- parameters
	p$rdev_y <- parameters$rdev_y[seq_len(keep)]
	m <- map
	if (!is.null(m$rdev_y)) m$rdev_y <- droplevels(factor(m$rdev_y[seq_len(keep)]))

	for (nm in handled) {
		dx <- dim(d[[nm]])
		if (!is.null(dx) && dx[1] != keep) stop("Failed to truncate ", nm, ".")
	}
	if (any(d$cpue_data$ts > keep) || length(p$rdev_y) != keep ||
			(!is.null(m$rdev_y) && length(m$rdev_y) != keep)) {
		stop("Peel validation failed for time-indexed data or recruitment deviations.")
	}
	list(data = d, parameters = p, map = m, keep = keep, n_drop = n_drop)
}
```

## Aligning parameters across layouts

Peels and the bound-fixing step both change the *active* parameter layout: a peel
shortens `rdev_y`, and fixing a bounded parameter removes it. Matching starting
values or bounds by position within a parameter name is only valid when the two
layouts are identical, so both operations are keyed instead to the element of the
parameter template that each active slot controls (`par_sel[17]`, `rdev_y[233]`).
Those keys are stable under both operations.

The correspondence between an active slot and the template element it controls is
*derived* from RTMB rather than assumed. `parList()` expands a reduced vector back
into the full parameter list, so expanding a probe vector `1, 2, ..., n` reveals
which elements each slot drives: the element holding the value `k` is controlled by
slot `k`. A second expansion of an all-zero vector separates active elements from
fixed ones, whose values do not respond to the probe.

This removes any dependence on how TMB collapses a mapped parameter, which is not
reliably first-appearance order and which an earlier version of this chunk got
wrong.

```{r}
#| label: par-keys
#| echo: true
#| message: false
key_parameter <- function(keys) sub("\\[[0-9]+\\]$", "", keys)
key_element <- function(keys) as.integer(sub("^.*\\[([0-9]+)\\]$", "\\1", keys))

# parList()'s signature has varied across TMB versions. Use whichever call form
# actually responds to the supplied vector.
expand_probe <- function(obj, values) {
	forms <- list(
		function(v) obj$env$parList(v),
		function(v) obj$env$parList(par = v),
		function(v) obj$env$parList(x = v, par = v)
	)
	for (form in forms) {
		base <- tryCatch(form(rep(0, length(values))), error = function(e) NULL)
		probe <- tryCatch(form(values), error = function(e) NULL)
		if (is.null(base) || is.null(probe)) next
		if (!identical(unlist(base), unlist(probe))) {
			return(list(base = base, probe = probe))
		}
	}
	stop("obj$env$parList() did not respond to the probe vector, so the active ",
			 "parameter layout cannot be derived.")
}

# Map each active slot to the template element it controls, e.g. "par_sel[17]".
# Where several elements share a slot, the lowest element index becomes the key.
active_par_keys <- function(obj) {
	if (length(obj$env$random)) {
		stop("Key derivation assumes a fixed-effects model; obj has random effects.")
	}
	n <- length(obj$par)
	expanded <- expand_probe(obj, as.numeric(seq_len(n)))
	keys <- rep(NA_character_, n)
	for (nm in names(expanded$probe)) {
		base <- as.numeric(expanded$base[[nm]])
		probe <- as.numeric(expanded$probe[[nm]])
		if (!length(probe)) next
		# Fixed elements hold their template value under both probes.
		for (el in which(base == 0 & probe != 0)) {
			slot <- round(probe[[el]])
			if (abs(probe[[el]] - slot) > 1e-8 || slot < 1 || slot > n) next
			slot <- as.integer(slot)
			if (is.na(keys[[slot]])) keys[[slot]] <- sprintf("%s[%d]", nm, el)
		}
	}
	if (anyNA(keys)) {
		stop("Could not identify template elements for ", sum(is.na(keys)), " of ",
				 n, " active parameters.")
	}
	if (!identical(key_parameter(keys), names(obj$par))) {
		stop("Derived keys do not match names(obj$par); the parList() expansion is ",
				 "not consistent with the active parameter names.")
	}
	keys
}

align_by_key <- function(keys, keyed_values, label) {
	pos <- match(keys, names(keyed_values))
	if (anyNA(pos)) {
		stop("Cannot align ", label, " for: ",
				 paste(keys[is.na(pos)], collapse = ", "))
	}
	stats::setNames(as.numeric(keyed_values[pos]), keys)
}
```

```{r}
#| label: reference-keys
#| echo: true
#| message: false
reference_keys <- active_par_keys(obj_reference)

# Master keyed stores. Every later layout is a subset of these keys.
keyed_start <- stats::setNames(as.numeric(bet_reference_par), reference_keys)
keyed_lower <- stats::setNames(as.numeric(bounds$lower), reference_keys)
keyed_upper <- stats::setNames(as.numeric(bounds$upper), reference_keys)

stopifnot(length(reference_keys) == length(bet_reference_par),
					!anyDuplicated(reference_keys))
```

```{r}
#| label: tbl-key-derivation
#| echo: true
#| tbl-cap: Derived active parameter layout. `slots` counts the active parameters contributed by each entry of the parameter list; the first and last keys show the template elements at either end of that block.
tibble(key = reference_keys, slot = seq_along(reference_keys)) %>%
	mutate(parameter = key_parameter(key)) %>%
	group_by(parameter) %>%
	summarise(slots = n(), first_slot = min(slot), last_slot = max(slot),
						first_key = key[which.min(slot)], last_key = key[which.max(slot)],
						.groups = "drop") %>%
	arrange(first_slot) %>%
	knitr::kable()
```

## Fitting and running peels

```{r}
#| label: peel-fitting-functions
#| echo: true
#| message: false
summarise_estimability <- function(obj, par) {
	hessian <- tryCatch(obj$he(par), error = function(e) e)
	empty <- tibble(Param = character(), MLE = numeric(), Param_check = character())
	if (inherits(hessian, "error")) {
		return(list(status = "hessian_error", message = conditionMessage(hessian),
								min_eigen = NA_real_, n_bad_directions = NA_integer_,
								bad_params = empty))
	}
	check <- tryCatch(opal::check_estimability(obj, h = hessian),
										error = function(e) e)
	if (inherits(check, "error")) {
		return(list(status = "check_error", message = conditionMessage(check),
								min_eigen = NA_real_, n_bad_directions = NA_integer_,
								bad_params = empty))
	}
	bad_params <- as_tibble(check$BadParams) %>%
		select(any_of(c("Param", "MLE", "Param_check")))
	if (!all(c("Param", "MLE", "Param_check") %in% names(bad_params))) {
		bad_params <- empty
	}
	list(status = "ok", message = NA_character_,
			 min_eigen = min(Re(check$Eigen$values)),
			 n_bad_directions = length(check$WhichBad),
			 bad_params = bad_params)
}

bounded_parameters <- function(keys, par, lower, upper, tol = 1e-7) {
	at_lower <- is.finite(lower) & par <= lower + tol
	at_upper <- is.finite(upper) & par >= upper - tol
	hit <- which(at_lower | at_upper)
	tibble(
		key = keys[hit],
		parameter = key_parameter(keys[hit]),
		element = key_element(keys[hit]),
		estimate = as.numeric(par[hit]),
		bound = if_else(at_lower[hit], "lower", "upper"),
		bound_value = if_else(at_lower[hit], as.numeric(lower[hit]),
													as.numeric(upper[hit]))
	)
}

fit_peel <- function(inputs, keyed_start, keyed_lower, keyed_upper,
										 n_restart = 4L, grad_tol = 1e-2) {
	obj <- MakeADFun(func = cmb(opal_model, inputs$data),
									 parameters = inputs$parameters, map = inputs$map)
	obj$env$tracemgc <- FALSE
	keys <- active_par_keys(obj)

	par0 <- align_by_key(keys, keyed_start, "starting values")
	lower <- align_by_key(keys, keyed_lower, "lower bounds")
	upper <- align_by_key(keys, keyed_upper, "upper bounds")

	opt <- list(par = unname(par0))
	for (i in seq_len(n_restart)) {
		opt <- nlminb(start = opt$par, objective = obj$fn, gradient = obj$gr,
									lower = unname(lower), upper = unname(upper),
									control = list(eval.max = 10000, iter.max = 10000))
		if (max(abs(obj$gr(opt$par))) < grad_tol) break
	}
	invisible(obj$fn(opt$par))
	obj$env$last.par.best <- opt$par

	grad <- as.numeric(obj$gr(opt$par))
	at_lower <- is.finite(lower) & opt$par <= lower + 1e-7
	at_upper <- is.finite(upper) & opt$par >= upper - 1e-7
	kkt <- abs(grad)
	kkt[at_lower] <- pmax(0, -grad[at_lower])
	kkt[at_upper] <- pmax(0, grad[at_upper])

	rep <- obj$report(opt$par)
	par_mle <- obj$env$parList(opt$par)
	sdr <- tryCatch(RTMB::sdreport(obj), error = function(e) NULL)
	cpue_pred <- as.numeric(rep$cpue_pred)
	if (length(cpue_pred) != nrow(inputs$data$cpue_data)) {
		stop("CPUE predictions do not align with the peel CPUE observations.")
	}

	list(
		n_drop = inputs$n_drop, keep = inputs$keep,
		objective = as.numeric(opt$objective), convergence = opt$convergence,
		max_grad = max(abs(grad)), kkt_grad = max(kkt),
		n_at_bound = sum(at_lower | at_upper),
		keys = keys,
		par = stats::setNames(as.numeric(opt$par), keys),
		bound_report = bounded_parameters(keys, opt$par, lower, upper),
		spawning_biomass_y = as.numeric(rep$spawning_biomass_y),
		spawning_biomass0_y = as.numeric(rep$spawning_biomass0_y),
		static_depletion_y = as.numeric(rep$static_depletion_y),
		dynamic_depletion_y = as.numeric(rep$dynamic_depletion_y),
		cpue_data = inputs$data$cpue_data,
		cpue_pred = cpue_pred,
		par_sel = par_mle$par_sel,
		sd_summary = if (is.null(sdr)) NULL else summary(sdr, "report"),
		pdHess = if (is.null(sdr)) NA else sdr$pdHess,
		estimability = summarise_estimability(obj, opt$par)
	)
}
```

### Fixing parameters estimated on a bound

A parameter resting on a bound contributes no curvature and inflates the
`sdreport()` covariance for everything correlated with it. `fix_bounded_parameters()`
removes each such slot from estimation by setting the corresponding map entries to
`NA`, holding the parameter either at its estimated (bound) value or at its value
in the pre-seeded template.

Because the map is edited through the element keys rather than through positions
in the active vector, repeated hits within one parameter (several `par_sel`
coefficients, say) are handled correctly, and the keys of the surviving slots are
unchanged.

```{r}
#| label: fix-bounded
#| echo: true
#| message: false
fix_bounded_parameters <- function(parameters, map, bound_report,
																	 fix_at = c("estimate", "initial"),
																	 template = NULL) {
	fix_at <- match.arg(fix_at)
	if (identical(fix_at, "initial") && is.null(template)) {
		stop("fix_at = 'initial' requires a parameter template.")
	}
	if (!nrow(bound_report)) {
		return(list(parameters = parameters, map = map,
								fixed_keys = character(0), report = bound_report))
	}

	new_parameters <- parameters
	new_map <- map
	fixed_at <- numeric(nrow(bound_report))
	elements <- character(nrow(bound_report))

	for (j in seq_len(nrow(bound_report))) {
		nm <- bound_report$parameter[[j]]
		lead <- bound_report$element[[j]]
		if (is.null(new_map[[nm]])) {
			new_map[[nm]] <- factor(seq_along(new_parameters[[nm]]))
		}
		m <- new_map[[nm]]
		if (is.na(m[[lead]])) {
			stop("Element ", lead, " of ", nm, " is already fixed; duplicate key?")
		}
		elems <- which(!is.na(m) & m == m[[lead]])
		value <- if (identical(fix_at, "estimate")) {
			bound_report$estimate[[j]]
		} else {
			as.numeric(template[[nm]][[lead]])
		}
		new_parameters[[nm]][elems] <- value

		labels <- as.character(m)
		labels[elems] <- NA_character_
		# Preserve first-appearance level order so the active layout is unchanged.
		new_map[[nm]] <- factor(labels, levels = unique(labels[!is.na(labels)]))

		fixed_at[[j]] <- value
		elements[[j]] <- paste(elems, collapse = ",")
	}

	list(parameters = new_parameters, map = new_map,
			 fixed_keys = bound_report$key,
			 report = bound_report %>%
				 mutate(mapped_elements = elements, fixed_at = fixed_at))
}
```

### Convergence reporting

Two gradient summaries are reported for every model. `max_grad` is the raw
maximum absolute gradient, and `kkt_grad` is the Karush--Kuhn--Tucker adjusted
maximum: at a parameter resting on a bound, only the component pointing back into
the feasible region counts. A large `max_grad` with a small `kkt_grad` is the
signature of an active bound rather than a failure to converge, which is why both
are needed. Once bounded parameters have been fixed the two should agree.

```{r}
#| label: convergence-row
#| echo: true
convergence_row <- function(fit, label) {
	tibble(
		model = label,
		time_steps = fit$keep,
		active_parameters = length(fit$keys),
		objective = fit$objective,
		nlminb_convergence = fit$convergence,
		max_grad = fit$max_grad,
		kkt_grad = fit$kkt_grad,
		n_at_bound = fit$n_at_bound,
		pdHess = fit$pdHess
	)
}

bound_table <- function(fit, label) {
	if (!nrow(fit$bound_report)) {
		return(tibble(model = label, key = "None", parameter = NA_character_,
									bound = NA_character_, bound_value = NA_real_,
									estimate = NA_real_))
	}
	fit$bound_report %>%
		transmute(model = label, key, parameter, bound, bound_value, estimate)
}
```

### Cache provenance

Both baseline fits and the peel series share one cache file, but each entry
carries its own provenance record. Changing `n_peel` therefore reruns only the
peel series; changing `fix_bounded_at` reruns the refitted baseline and the peel
series but not the original baseline.

```{r}
#| label: cache-provenance
#| echo: true
# Provenance is recorded per cache entry and chained, so that changing the peel
# depth does not invalidate the two baseline fits.
provenance_baseline <- list(
	schema_version = 12L,
	reference_cache = normalizePath(base_fit_path),
	fit_config = bet_config[c("lf_fisheries", "wf_fisheries", "lf_var_adjust",
															"lf_var_override", "wf_var_adjust", "wf_var_override",
															"lf_switch", "wf_switch", "cpue_switch", "cpue_tau",
															"quarters_per_year")],
	start_values = "base_fit$opt$par, keyed by parameter element",
	estimability = "opal::check_estimability(obj, h = obj$he(par))"
)
provenance_fixed <- c(provenance_baseline, list(
	fix_bounded_at = bet_config$fix_bounded_at,
	fix_policy = "single pass, determined from the original baseline"
))
provenance_peels <- c(provenance_fixed, list(n_peel = bet_config$n_peel))

# Snapshot the stored provenance before any chunk writes to the cache.
stored_provenance <- retro_cache$provenance
if (is.null(stored_provenance)) stored_provenance <- list()

needs_fit <- function(key, provenance, companions = character(0)) {
	if (!identical(stored_provenance[[key]], provenance)) return(TRUE)
	if (is.null(retro_cache[[key]])) return(TRUE)
	any(vapply(companions, function(nm) is.null(retro_cache[[nm]]), logical(1)))
}

# Records the entry's provenance and flushes the shared cache file.
write_cache <- function(key, provenance) {
	retro_cache$provenance[[key]] <<- provenance
	saveRDS(retro_cache, cache_file)
}
```

### Original baseline

The reconstructed retained fit, refitted here under the peel machinery with no
parameters fixed beyond those already mapped off in the retained configuration.
This is the model whose bound behaviour determines what is fixed in the next
step.

```{r}
#| label: fit-baseline-initial
#| echo: true
#| message: false
if (needs_fit("baseline_initial", provenance_baseline)) {
	message("Fitting the original baseline.")
	inputs_initial <- peel_bet_inputs(base_data, bet_parameters, bet_map,
																		lf_wide, wf_wide, n_drop = 0L,
																		config = bet_config)
	retro_cache$baseline_initial <- fit_peel(inputs_initial, keyed_start,
																					 keyed_lower, keyed_upper)
	write_cache("baseline_initial", provenance_baseline)
}
baseline_initial <- retro_cache$baseline_initial
```

```{r}
#| label: tbl-baseline-initial
#| echo: true
#| tbl-cap: Convergence diagnostics for the original baseline fit. Both the raw and KKT-adjusted maximum gradients are shown.
convergence_row(baseline_initial, "Original baseline") %>%
	mutate(across(c(objective, max_grad, kkt_grad), ~ signif(.x, 6))) %>%
	knitr::kable()
```

```{r}
#| label: tbl-baseline-initial-bounds
#| echo: true
#| tbl-cap: Parameters estimated on a bound in the original baseline fit.
bound_table(baseline_initial, "Original baseline") %>%
	mutate(across(c(bound_value, estimate), ~ signif(.x, 5))) %>%
	knitr::kable()
```

### Refitted baseline

Every parameter listed above is removed from estimation by setting its map entry
to `NA` and holding it at `bet_config$fix_bounded_at`. The model is then refitted.
Fixing at the estimated value pins the parameter to the bound it was already
resting on, so the objective should be unchanged to within optimiser tolerance and
`max_grad` should fall to meet `kkt_grad`.

```{r}
#| label: fit-baseline-fixed
#| echo: true
#| message: false
fixed_companions <- c("baseline_parameters", "baseline_map",
													"baseline_keyed_start", "baseline_fix_report")
if (needs_fit("baseline_fixed", provenance_fixed, fixed_companions)) {
	message("Fixing bounded parameters and refitting the baseline.")
	fix_info <- fix_bounded_parameters(
		parameters = bet_parameters, map = bet_map,
		bound_report = baseline_initial$bound_report,
		fix_at = bet_config$fix_bounded_at,
		template = bet_parameters_template
	)

	# Fitted values from the original baseline seed the refit.
	baseline_keyed_start <- keyed_start
	baseline_keyed_start[names(baseline_initial$par)] <- baseline_initial$par

	inputs_fixed <- peel_bet_inputs(base_data, fix_info$parameters, fix_info$map,
																	lf_wide, wf_wide, n_drop = 0L,
																	config = bet_config)
	retro_cache$baseline_fixed <- fit_peel(inputs_fixed, baseline_keyed_start,
																				 keyed_lower, keyed_upper)
	retro_cache$baseline_parameters <- fix_info$parameters
	retro_cache$baseline_map <- fix_info$map
	retro_cache$baseline_keyed_start <- baseline_keyed_start
	retro_cache$baseline_fix_report <- fix_info$report
	write_cache("baseline_fixed", provenance_fixed)
}
baseline_fixed <- retro_cache$baseline_fixed
baseline_parameters <- retro_cache$baseline_parameters
baseline_map <- retro_cache$baseline_map
baseline_keyed_start <- retro_cache$baseline_keyed_start
baseline_fix_report <- retro_cache$baseline_fix_report
```

```{r}
#| label: tbl-baseline-fixed
#| echo: true
#| tbl-cap: Convergence diagnostics for the original and refitted baselines. A drop in active parameters with an unchanged objective is the expected result of fixing parameters at their bound values.
bind_rows(
	convergence_row(baseline_initial, "Original baseline"),
	convergence_row(baseline_fixed, "Refitted baseline")
) %>%
	mutate(across(c(objective, max_grad, kkt_grad), ~ signif(.x, 6))) %>%
	knitr::kable()
```

```{r}
#| label: tbl-baseline-fix-report
#| echo: true
#| tbl-cap: Parameters removed from estimation, the map elements they control, and the values at which they were fixed.
if (is.null(baseline_fix_report) || !nrow(baseline_fix_report)) {
	tibble(note = "No parameters were estimated on a bound.") %>% knitr::kable()
} else {
	baseline_fix_report %>%
		select(key, parameter, mapped_elements, bound, bound_value, estimate,
					 fixed_at) %>%
		mutate(across(c(bound_value, estimate, fixed_at), ~ signif(.x, 5))) %>%
		knitr::kable()
}
```

```{r}
#| label: tbl-baseline-fixed-bounds
#| echo: true
#| tbl-cap: Parameters still resting on a bound after the refit. These are reported only; the fixed set is determined once, from the original baseline.
bound_table(baseline_fixed, "Refitted baseline") %>%
	mutate(across(c(bound_value, estimate), ~ signif(.x, 5))) %>%
	knitr::kable()
```

The objective difference between the two baselines is
`r if (!is.null(baseline_fixed)) signif(baseline_fixed$objective - baseline_initial$objective, 4) else NA`.
A material increase means either that the `1e-7` bound tolerance is flagging
parameters that are not genuinely active, or that `fix_bounded_at = "initial"` has
moved a parameter away from its optimum.

### Peel series

::: {.callout-note}
## The model structure is held constant across peels

Every peel inherits the fixed set determined by the original full-data baseline,
and peel 0 is the refitted baseline itself. Allowing each peel to fix its own
bounded parameters would let the model structure vary with the amount of data, and
Mohn's $\rho$ would then confound structural change with the retrospective revision
it is meant to measure. Peels are still checked for new bound hits, which are
reported rather than acted on.
:::

```{r}
#| label: fit-peels
#| echo: true
#| message: false
if (needs_fit("peels", provenance_peels)) {
	peel_depths <- (0:bet_config$n_peel) * bet_config$quarters_per_year
	peels <- purrr::map(peel_depths, function(n_drop) {
		if (n_drop == 0L) return(baseline_fixed)
		message("Peel with ", n_drop / bet_config$quarters_per_year,
						" calendar years removed")
		inputs <- peel_bet_inputs(base_data, baseline_parameters, baseline_map,
															lf_wide, wf_wide, n_drop = n_drop,
															config = bet_config)
		fit_peel(inputs, baseline_keyed_start, keyed_lower, keyed_upper)
	})
	names(peels) <- paste0("peel_", 0:bet_config$n_peel)
	retro_cache$peels <- peels
	write_cache("peels", provenance_peels)
}
peels <- retro_cache$peels
```

```{r}
#| label: tbl-peel-bounds
#| echo: true
#| tbl-cap: Parameters resting on a bound within each peel, after the baseline fixing step.
purrr::map_dfr(peels, function(p) {
	tibble(peel = p$n_drop / bet_config$quarters_per_year,
				 on_bound = nrow(p$bound_report),
				 parameters = if (nrow(p$bound_report)) {
					 paste(sort(unique(p$bound_report$key)), collapse = ", ")
				 } else {
					 "None"
				 })
}) %>%
	knitr::kable()
```

```{r}
#| label: tbl-peel-convergence
#| echo: true
#| tbl-cap: Convergence and estimability diagnostics by retrospective peel. Estimability is evaluated only for active parameters and should be interpreted with the KKT gradient and Hessian status.
purrr::map_dfr(peels, function(p) {
	est <- p$estimability
	bad_names <- if (identical(est$status, "ok")) {
		bad <- est$bad_params %>% filter(Param_check == "Bad") %>% pull(Param)
		if (length(bad)) paste(unique(bad), collapse = ", ") else "None"
	} else {
		NA_character_
	}
	tibble(peel = p$n_drop / bet_config$quarters_per_year,
				 time_steps = p$keep, objective = p$objective,
				 max_grad = p$max_grad, kkt_grad = p$kkt_grad,
				 n_at_bound = p$n_at_bound, 
				 estimability = est$status,
				 min_eigen = est$min_eigen,
				 n_bad_directions = est$n_bad_directions,
				 bad_parameters = bad_names,
				 estimability_message = est$message)
}) %>%
	mutate(retained = kkt_grad < 1e-2,
				 across(c(objective, max_grad, kkt_grad, min_eigen), ~ signif(.x, 4))) %>%
	knitr::kable()
```

`check_estimability()` examines the Hessian only for parameters active in the
peel's RTMB objective. A `Bad` result therefore identifies an active parameter
that contributes materially to a near-null Hessian direction; it is not a
convergence verdict. Read it alongside the KKT gradient, and any
estimability diagnostic error shown in the table.

# Peel time series

The plot below overlays all peels within each metric facet. Pointwise Wald 95%
intervals come from `sdreport()` and use the same colour as their peel. Intervals
are clipped to the feasible scale: zero for spawning biomass and $[0, 1]$ for
depletion. A peel without a usable `sdreport()` result retains its estimate but
has no uncertainty ribbon.

```{r}
#| label: prepare-peel-time-series
#| echo: true
report_series <- function(peel_fit, peel_label, quantity, metric) {
	n <- length(peel_fit[[quantity]])
	se <- rep(NA_real_, n)
	if (!is.null(peel_fit$sd_summary) && "Std. Error" %in% colnames(peel_fit$sd_summary)) {
		rows <- which(rownames(peel_fit$sd_summary) == quantity)
		if (length(rows)) {
			se[seq_len(min(n, length(rows)))] <-
				peel_fit$sd_summary[rows[seq_len(min(n, length(rows)))], "Std. Error"]
		}
	}
	tibble(
		peel = peel_label,
		calendar_year = as.numeric(base_data$years[seq_len(n)]),
		metric = metric,
		estimate = as.numeric(peel_fit[[quantity]]),
		se = as.numeric(se)
	)
}

peel_labels <- paste0("Peel ", purrr::map_int(peels, "n_drop") / bet_config$quarters_per_year)
peel_time_series <- purrr::imap_dfr(peels, function(peel_fit, peel_name) {
	peel_label <- paste0("Peel ", peel_fit$n_drop / bet_config$quarters_per_year)
	bind_rows(
		report_series(peel_fit, peel_label, "spawning_biomass_y", "Spawning biomass"),
		report_series(peel_fit, peel_label, "static_depletion_y", "Static depletion"),
		report_series(peel_fit, peel_label, "dynamic_depletion_y", "Dynamic depletion")
	)
}) %>%
	mutate(
		peel = factor(peel, levels = peel_labels),
		lower = estimate - qnorm(0.975) * se,
		upper = estimate + qnorm(0.975) * se,
		lower = pmax(0, lower),
		upper = upper)
```

```{r}
#| label: fig-peel-time-series
#| fig-cap: Retrospective time series by metric. Colour identifies the number of calendar years removed; ribbons are pointwise 95% uncertainty intervals.
#| fig-width: 12
#| fig-height: 10
peel_time_series %>%
	ggplot(aes(calendar_year, estimate, group = peel)) +
	geom_ribbon(aes(ymin = lower, ymax = upper, fill = peel), alpha = 0.25, na.rm = TRUE) +
	geom_line(linewidth = 1.15, aes(colour = peel)) +
	facet_wrap(~ metric, scales = "free_y", ncol = 3) +
	scale_colour_viridis_d(option = "turbo", end = 0.9) +
	scale_fill_viridis_d(option = "turbo", end = 0.9) +
	labs(x = "Calendar year", y = NULL, colour = "Peel (years removed)", fill = "Peel (years removed)") +
	guides(fill = "none") +
	theme(legend.position = "bottom")
```

# Index fit by peel

The index-fit plot uses the log-scale RMSE, $
\sqrt{\operatorname{mean}[(\log I_\mathrm{obs} - \log I_\mathrm{pred})^2]}$,
calculated separately for each retained index observation in a peel.

```{r}
#| label: prepare-peel-index-fit
#| echo: true
peel_cpue_fit <- purrr::imap_dfr(peels, function(peel_fit, peel_name) {
	# CPUE predictions follow the baseline rows retained by this peel exactly.
	cpue <- base_data$cpue_data %>% filter(ts <= peel_fit$keep)
	if (!all(c("year", "month", "value") %in% names(cpue))) {
		stop("Baseline CPUE data are missing year, month, or value.")
	}
	if (is.null(peel_fit$cpue_pred)) {
		stop("Cached peel lacks CPUE predictions; rerun the fit-peels chunk to refresh the cache.")
	}
	if (nrow(cpue) != length(peel_fit$cpue_pred)) {
		stop("Baseline CPUE rows and cached peel predictions have incompatible lengths.")
	}
	cpue %>%
		transmute(
			peel = paste0("Peel ", peel_fit$n_drop / bet_config$quarters_per_year),
			calendar_year = as.numeric(year) + (as.numeric(month) - 1) / 12,
			index = factor(month, levels = sort(unique(base_data$cpue_data$month)),
				labels = month.abb[sort(unique(base_data$cpue_data$month))]),
			observed = as.numeric(value),
			predicted = as.numeric(peel_fit$cpue_pred)
		)
}) %>%
	mutate(peel = factor(peel, levels = peel_labels))

cpue_observed <- peel_cpue_fit %>%
	select(index, calendar_year, observed) %>%
	distinct()
cpue_rmse <- peel_cpue_fit %>%
	filter(observed > 0, predicted > 0) %>%
	group_by(index, peel) %>%
	summarise(rmse = sqrt(mean((log(observed) - log(predicted))^2)), .groups = "drop") %>%
	group_by(index) %>%
	arrange(peel, .by_group = TRUE) %>%
	mutate(label_rank = row_number(), label = paste0(as.character(peel), ": ", sprintf("%.3f", rmse))) %>%
	ungroup()
```

```{r}
#| label: fig-peel-index-fit
#| fig-cap: Observed and predicted CPUE by index and retrospective peel. Labels give the log-scale RMSE for each peel within an index.
#| fig-width: 12
#| fig-height: 7
ggplot(peel_cpue_fit, aes(calendar_year, predicted, colour = peel, group = peel)) +
	geom_point(data = cpue_observed, aes(x = calendar_year, y = observed), inherit.aes = FALSE,
				 colour = "grey15", size = 1.1, alpha = 0.7) +
	geom_line(linewidth = 1.1) +
	geom_text(data = cpue_rmse, aes(x = Inf, y = Inf, label = label, colour = peel,
											group = peel, vjust = 1.1 + 1.15 * label_rank),
				 hjust = 1.05, size = 2.8, show.legend = FALSE) +
	facet_wrap(~ index, scales = "free_y") +
	scale_y_log10() +
	scale_colour_viridis_d(option = "turbo", end = 0.9) +
	labs(x = "Calendar year", y = "Standardized CPUE", colour = "Peel (years removed)") +
	theme(legend.position = "bottom")
```

# Selectivity by peel

```{r}
#| label: prepare-peel-selectivity
#| echo: true
sel_logistic <- function(len, par) {
	mean_length <- mean(len)
	sd_length <- sd(len)
	a50 <- mean_length + par[1] * sd_length
	w95 <- exp(par[2]) * sd_length
	1 / (1 + exp(-log(19) * (len - a50) / w95))
}

sel_double_normal <- function(len, par) {
	mean_length <- mean(len)
	sd_length <- sd(len)
	peak <- mean_length + par[1] * sd_length
	upper_scale <- exp(par[3]) * sd_length^2
	lower_scale <- exp(par[4]) * sd_length^2
	upper_point <- plogis(par[5])
	lower_point <- plogis(par[6])
	last <- length(len)
	bin_width <- len[2] - len[1]
	second_peak <- peak + bin_width + (0.99 * len[last] - peak - bin_width) /
		(1 + exp(-par[2]))
	upper_min <- exp(-(len[1] - peak)^2 / upper_scale)
	lower_min <- exp(-(len[last] - second_peak)^2 / lower_scale)
	upper_distance <- len - peak
	lower_distance <- len - second_peak
	join_upper <- plogis((20 / (1 + abs(upper_distance))) * upper_distance)
	join_lower <- plogis((20 / (1 + abs(lower_distance))) * lower_distance)
	ascending <- upper_point + (1 - upper_point) *
		(exp(-upper_distance^2 / upper_scale) - upper_min) / (1 - upper_min)
	descending <- 1 + (lower_point - 1) *
		(exp(-lower_distance^2 / lower_scale) - 1) / (lower_min - 1)
	ascending * (1 - join_upper) + join_upper * (1 - join_lower + descending * join_lower)
}

sel_type_f <- base_fit$diagnostics$selectivity_raw$sel_type_f
if (is.null(sel_type_f)) sel_type_f <- base_data$sel_type_f
if (is.null(sel_type_f)) stop("Selectivity types are unavailable in the retained fit cache.")

peel_selectivity <- purrr::imap_dfr(peels, function(peel_fit, peel_name) {
	par_sel <- peel_fit$par_sel
	if (is.null(dim(par_sel))) stop("Cached peel selectivity parameters are not a matrix.")
	if (nrow(par_sel) != length(sel_type_f)) {
		stop("Cached peel selectivity parameters and selectivity types are incompatible.")
	}
	purrr::map_dfr(seq_len(nrow(par_sel)), function(fishery) {
		selectivity <- if (sel_type_f[[fishery]] == 1L) {
			sel_logistic(base_data$len_mid, par_sel[fishery, ])
		} else {
			sel_double_normal(base_data$len_mid, par_sel[fishery, ])
		}
		tibble(
			peel = paste0("Peel ", peel_fit$n_drop / bet_config$quarters_per_year),
			fishery = paste("Fishery", fishery),
			length = base_data$len_mid,
			selectivity = pmin(1, pmax(0, as.numeric(selectivity)))
		)
	})
}) %>%
	mutate(
		peel = factor(peel, levels = peel_labels),
		fishery = factor(fishery, levels = paste("Fishery", seq_along(sel_type_f)))
	)
```

```{r}
#| label: fig-peel-selectivity
#| fig-cap: Estimated selectivity-at-length curves for each retrospective peel. Colours identify the number of calendar years removed.
#| fig-width: 12
#| fig-height: 14
ggplot(peel_selectivity, aes(length, selectivity, colour = peel, group = peel)) +
	geom_line(linewidth = 1.1) +
	facet_wrap(~ fishery, ncol = 3) +
	coord_cartesian(ylim = c(0, 1)) +
	scale_colour_viridis_d(option = "turbo", end = 0.9) +
	labs(x = "Length", y = "Selectivity", colour = "Peel (years removed)") +
	theme(legend.position = "bottom")
```

# Mohn's rho

```{r}
#| label: mohn-rho
#| echo: true
qpy <- bet_config$quarters_per_year
mohn_rho <- function(peels, quantity, unit = c("quarter", "year"), drop_last = 0L) {
	unit <- match.arg(unit)
	full <- peels[[1]][[quantity]]
	use <- peels[-1]
	if (drop_last > 0L) use <- use[seq_len(length(use) - drop_last)]
	terms <- purrr::map_dfr(use, function(p) {
		idx <- if (unit == "quarter") p$keep else seq(p$keep - qpy + 1L, p$keep)
		x_peel <- mean(p[[quantity]][idx])
		x_full <- mean(full[idx])
		tibble(peel = p$n_drop / qpy, x_peel = x_peel, x_full = x_full,
					 rel_diff = (x_peel - x_full) / x_full)
	})
	list(rho = mean(terms$rel_diff), terms = terms)
}

quantities <- c("spawning_biomass_y", "dynamic_depletion_y", "static_depletion_y")
rho_table <- tidyr::expand_grid(quantity = quantities,
																unit = c("quarter", "year"),
																drop_last = c(0L, 1L)) %>%
	mutate(res = purrr::pmap(list(quantity, unit, drop_last),
													 ~ mohn_rho(peels, ..1, ..2, ..3)),
				 rho = purrr::map_dbl(res, "rho"),
				 min_diff = purrr::map_dbl(res, ~ min(.x$terms$rel_diff)),
				 max_diff = purrr::map_dbl(res, ~ max(.x$terms$rel_diff))) %>%
	select(-res)

rho_table %>%
	mutate(unit = recode(unit, quarter = "Terminal quarter", year = "Terminal-year mean"),
				 peels_used = if_else(drop_last == 0L, "all", "final peel dropped"),
				 across(c(rho, min_diff, max_diff), ~ round(.x, 3)),
				 outside_rule = rho < -0.15 | rho > 0.20) %>%
	select(quantity, unit, peels_used, rho, min_diff, max_diff, outside_rule) %>%
	knitr::kable()
```

# Interpretation

A retrospective pattern indicates that post-sample data update an estimate in
a way that is inconsistent with the fitted model conditioned on earlier data.
It does not identify the direction or magnitude of estimation error, and does
not justify adjusting $\rho$.

The post-sample bootstrap significance test is intentionally not executed here.
Each replicate requires all retrospective refits and should be run separately
after the deterministic peel series has been reviewed.
